```         
# 1. Install devtools if not already available
if (!requireNamespace("devtools", quietly = TRUE)) {
  install.packages("devtools")
}

# 2. Install nichenetr from GitHub
devtools::install_github("saeyslab/nichenetr")
```

```         
# Switch download method (useful on Windows if the default method fails)
options(download.file.method = "wininet")

# Retry the multinichenetr installation
remotes::install_github("saeyslab/multinichenetr")
```

`# Install devtools if not already installed`

`if (!requireNamespace("devtools", quietly = TRUE))`

`install.packages("devtools")`

`# Install BiocManager if needed`

`if (!requireNamespace("BiocManager", quietly = TRUE))`

`install.packages("BiocManager")`

`# Install Bioconductor dependencies`

`BiocManager::install("ComplexHeatmap")`

`# Install CRAN packages needed alongside nichenetr`

`install.packages(c("tidyverse", "circlize", "ggrepel", "igraph", "ggraph"))`

`# Install nichenetr from GitHub (use the saeyslab fork — the actively maintained version)`

`devtools::install_github("saeyslab/nichenetr")`

# MultiNicheNet x WGCNA Integration

```{r libraries-integration}
library(reticulate)
library(Seurat)
library(SingleCellExperiment)
library(nichenetr)
library(multinichenetr)
library(dplyr)
library(tibble)
library(purrr)
library(ggplot2)
py_require(c("anndata"))
ad <- import("anndata")


# Used further below for the circos plot. GO enrichment itself is NOT
# recomputed here -- it's reused directly from the go_full_module_df /
# go_hubs_df objects already computed in WGCNA_24_hdwgcna_HVG.Rmd (same R session).
if (!requireNamespace("circlize", quietly = TRUE)) install.packages("circlize")
if (!requireNamespace("openxlsx", quietly = TRUE)) install.packages("openxlsx")
if (!requireNamespace("visNetwork", quietly = TRUE)) install.packages("visNetwork")
if (!requireNamespace("igraph", quietly = TRUE)) install.packages("igraph")
if (!requireNamespace("ggraph", quietly = TRUE)) install.packages("ggraph")
library(circlize)
library(openxlsx)
library(visNetwork)
library(igraph)
```

```{r}
select <- dplyr::select
# o, más robusto si usas conflictos frecuentes:
library(conflicted)
conflicts_prefer(dplyr::select, dplyr::filter, dplyr::mutate)
```

## Step 0: Build the SingleCellExperiment (SCE) input for MultiNicheNet

No `.h5ad` reload, no re-mapping of `cell_ontology_class`, no re-QC: `seurat_combined_filt`, `meta` (with `meta$final_celltype` already collapsed AND restricted to `target_celltypes`) and `keep_cells` are taken exactly as they exist at the end of `WGCNA_24_hdwgcna_HVG.Rmd`. This guarantees MultiNicheNet analyzes the identical cells, the identical cell type definitions, and the identical cell type restriction that WGCNA used for its pseudobulk network -- there is no risk of the two analyses silently drifting apart.

`colData(sce)` is still built explicitly (instead of relying on `Seurat::as.SingleCellExperiment()`), because that automatic conversion was found to silently drop most metadata columns in this project (only 3 columns were carried over instead of the expected 26). Building `colData` directly from `seurat_mnn@meta.data`, aligned by cell name, guarantees every column -- including `sex` and `tech` -- ends up in the SCE.

```{r}
heart_facs    <- ad$read_h5ad("D:/SandraAnjo/Luis Grilo/WGCNAandothersOlaia/data/Heart_facs.h5ad")
heart_droplet <-ad$read_h5ad("D:/SandraAnjo/Luis Grilo/WGCNAandothersOlaia/data/Heart_droplet.h5ad")

# var_names is already the gene symbol
facs_genes    <- rownames(py_to_r(heart_facs$var))
droplet_genes <- rownames(py_to_r(heart_droplet$var))
shared_genes  <- intersect(facs_genes, droplet_genes)
cat("Shared genes:", length(shared_genes), "\n")

combined_heart <- ad$concat(
  list(heart_droplet, heart_facs),
  axis = 0L,
  join = "outer",
  keys = c("droplet", "facs"),
  label = "tech",
  index_unique = "-"
)
combined_genes <- py_to_r(combined_heart$var_names$values)
combined_cells <- py_to_r(combined_heart$obs_names$values)

cat("Combined genes:", length(combined_genes), "\n")
cat("Combined cells:", length(combined_cells), "\n")

stopifnot(length(combined_genes) == ncol(combined_heart$X))
stopifnot(length(combined_cells) == nrow(combined_heart$X))

counts_matrix <- Matrix::t(as(combined_heart$X, "CsparseMatrix"))

sanitized_genes <- gsub("_", "-", combined_genes)
dup_idx <- duplicated(sanitized_genes) | duplicated(sanitized_genes, fromLast = TRUE)
if (sum(dup_idx) > 0) {
  cat("Genes affected by sanitization collisions:", sum(dup_idx), "\n")
  print(data.frame(original = combined_genes[dup_idx],
                   sanitized = sanitized_genes[dup_idx]))
}
final_genes <- make.unique(sanitized_genes)
stopifnot(sum(duplicated(final_genes)) == 0)
stopifnot(length(final_genes) == nrow(counts_matrix))

final_cells <- as.vector(combined_cells)
stopifnot(is.null(attributes(final_cells)))

rownames(counts_matrix) <- final_genes
colnames(counts_matrix) <- final_cells

metadata <- py_to_r(combined_heart$obs)
metadata <- as.data.frame(metadata, stringsAsFactors = FALSE)
rownames(metadata) <- final_cells

# 'tech' should have exactly 2 levels (facs/droplet)
cat("'tech' values (real technology):\n")
print(table(metadata$tech))

# Rename the original 'batch' column: it's an internal sequencing batch,
# not the FACS/droplet technology
if ("batch" %in% colnames(metadata)) {
  metadata$seq_batch <- metadata$batch
  metadata$batch <- NULL
}

# sample_id is built from mouse identity (or sex+age as fallback), WITHOUT
# tech, so that metacells merge cells from both sequencing technologies
# for the same mouse / same age-sex point. This relies on 'harmony' as the
# integration reduction in MetacellsByGroups() to correct for tech-driven
# batch effects when averaging cells into metacells.
if ("mouse.id" %in% colnames(metadata)) {
  metadata$sample_id <- as.character(metadata$mouse.id)
} else {
  metadata$sample_id <- paste(metadata$sex, metadata$age, sep = "_")
}

cat("Numero de sample_id unicos:", dplyr::n_distinct(metadata$sample_id), "\n")
table(metadata$sample_id)

cat("Do counts_matrix and metadata match?",
    identical(rownames(metadata), colnames(counts_matrix)), "\n")
stopifnot(identical(rownames(metadata), colnames(counts_matrix)))

seurat_combined <- CreateSeuratObject(
  counts = counts_matrix,
  meta.data = metadata,
  project = "WGCNA_Mouse_Combined"
)
# QC thresholds 
min_features <- 200
n_before <- ncol(seurat_combined)

seurat_combined <- subset(
  seurat_combined,
  subset = nFeature_RNA > min_features 
)
cat("Cells before QC:", n_before, "\n")
cat("Cells after QC:", ncol(seurat_combined), "\n")

# Remove genes detected in very few cells globally
min_cells_gene <- 3
detected_genes <- Matrix::rowSums(GetAssayData(seurat_combined, layer = "counts") > 0)
genes_to_keep <- names(detected_genes[detected_genes >= min_cells_gene])

cat("Genes before filter:", nrow(seurat_combined), "\n")
cat("Genes after filter:", length(genes_to_keep), "\n")

seurat_combined <- subset(seurat_combined, features = genes_to_keep)
seurat_combined_raw<-seurat_combined
seurat_combined_filt_raw<-subset(seurat_combined_raw, subset = cell_ontology_class != "nan")

celltype_map <- c(
  "endothelial cell of coronary artery" = "EndothelialCells",
  "aortic endothelial cell"             = "EndothelialCells",
  "atrial myocyte"                      = "Myocytes",
  "ventricular myocyte"                 = "Myocytes",
  "fibroblast of cardiac tissue"        = "Fibroblasts",
  "leukocyte"                           = "Leukocytes", 
  "smooth muscle cell"                  = "SmoothMuscleCells", 
  "endocardial cell"                    = "EndocardialCells"
  # mast cell, valve cell -> excluded on purpose (not in this vector = dropped)
)

meta_raw <- seurat_combined_filt_raw@meta.data
meta_raw$cell_ontology_class_clean <- trimws(tolower(as.character(meta_raw$cell_ontology_class)))

names(celltype_map) <- tolower(names(celltype_map))
meta_raw$final_celltype <- celltype_map[meta_raw$cell_ontology_class_clean]

unmapped <- sort(unique(meta_raw$cell_ontology_class[is.na(meta_raw$final_celltype)]))
cat("Unmapped / excluded cell_ontology_class levels (dropped from analysis):\n")
print(unmapped)

cat("\nCells per final cell type:\n")
print(table(meta_raw$final_celltype, useNA = "ifany"))

DefaultAssay(seurat_combined_filt_raw) <- "RNA"
all_genes <- rownames(seurat_combined_filt_raw)
expr <- as.matrix(GetAssayData(seurat_combined_filt_raw, layer = "counts")[all_genes, ])

# keep only cells that were successfully mapped to a final cell type
keep_cells <- rownames(meta_raw)[!is.na(meta_raw$final_celltype)]
meta_keep  <- meta_raw[keep_cells, ]
expr_keep  <- expr[, keep_cells]

age_per_mouse <- meta_keep %>%
  distinct(sample_id, age) %>%
  group_by(sample_id) %>%
  mutate(n_ages = n()) %>%
  ungroup()

stopifnot("Some mice have more than one 'age' value -- inspect age_per_mouse" =
            all(age_per_mouse$n_ages == 1))

age_lookup <- setNames(as.character(age_per_mouse$age), age_per_mouse$sample_id)

# as.factor() sorts levels alphabetically ("18m" before "3m"), which
# scrambles chronological order -- build the ordered levels from the
# numeric part of the label instead, exactly as in the metacell workflow.
ordered_age_levels <- unique(age_lookup)
ordered_age_levels <- ordered_age_levels[
  order(as.numeric(gsub("[^0-9]", "", ordered_age_levels)))
]
```

```{r mnn-build-sce}
seurat_mnn <- seurat_combined_filt_raw[, keep_cells]

seurat_mnn$final_celltype <- meta_keep[colnames(seurat_mnn), "final_celltype"]
seurat_mnn$sample_id      <- meta_keep[colnames(seurat_mnn), "sample_id"]
seurat_mnn$age            <- as.character(meta_keep[colnames(seurat_mnn), "age"])
seurat_mnn$sex            <- meta_keep[colnames(seurat_mnn), "sex"]
seurat_mnn$tech           <- meta_keep[colnames(seurat_mnn), "tech"]  # "facs" or "droplet"

cat("Cells in the MultiNicheNet input object:", ncol(seurat_mnn), "\n")
cat("\nCells per final_celltype:\n"); print(table(seurat_mnn$final_celltype))
cat("\nCells per age:\n"); print(table(seurat_mnn$age))
cat("\nCells per sex:\n"); print(table(seurat_mnn$sex))
cat("\nCells per tech:\n"); print(table(seurat_mnn$tech))
cat("\nMice (sample_id) per age:\n")
print(table(unique(data.frame(sample_id = seurat_mnn$sample_id, age = seurat_mnn$age))$age))

# MultiNicheNet needs raw counts (it pseudobulks internally via muscat/edgeR),
# not the already log-normalized data used by WGCNA.
counts_mat <- Seurat::GetAssayData(seurat_mnn, assay = "RNA", layer = "counts")

# Explicit colData, aligned by cell name -- do NOT rely on as.SingleCellExperiment()
col_data <- S4Vectors::DataFrame(seurat_mnn@meta.data[colnames(counts_mat), ])

sce <- SingleCellExperiment::SingleCellExperiment(
  assays  = list(counts = counts_mat),
  colData = col_data
)

cat("\nSCE dimensions (genes x cells):", dim(sce), "\n")
cat("Columns in colData(sce):", ncol(colData(sce)), "\n")
stopifnot(all(c("final_celltype", "sample_id", "age", "sex", "tech") %in% colnames(colData(sce))))
```

## Step 0b: drop age levels with too few biological replicates

`multi_nichenet_analysis()` fits its per-celltype pseudobulk model across **all** age levels present in `colData(sce)` at once, not only the two levels involved in a given contrast. An age level backed by a single `sample_id` (mouse) cannot have its between-replicate variability estimated, which breaks DE estimation for every cell type, not just that level -- this must be filtered out before any contrasts are built; no `min_cells` value can fix a lack of biological replicates.

```{r mnn-filter-low-replicate-ages}
min_samples_per_group <- 2

samples_per_age <- as.data.frame(colData(sce)) %>%
  distinct(sample_id, age) %>%
  count(age, name = "n_samples")

print(samples_per_age)

valid_ages <- samples_per_age %>%
  filter(n_samples >= min_samples_per_group) %>%
  pull(age)

dropped_ages <- base::setdiff(unique(colData(sce)$age), valid_ages)
cat("\nAge levels dropped for having <", min_samples_per_group, "samples:",
    paste(dropped_ages, collapse = ", "), "\n")
cat("Age levels kept:", paste(valid_ages, collapse = ", "), "\n")

sce <- sce[, colData(sce)$age %in% valid_ages]

# Drop unused factor levels everywhere they could linger
colData(sce)$age       <- droplevels(base::as.factor(colData(sce)$age))
colData(sce)$sample_id <- droplevels(base::as.factor(colData(sce)$sample_id))

# Keep ordered_age_levels (from WGCNA_24) in sync with what actually survives here
ordered_age_levels <- ordered_age_levels[ordered_age_levels %in% valid_ages]

cat("\nRemaining cells:", ncol(sce), "\n")
cat("Remaining age levels (chronological order):", paste(ordered_age_levels, collapse = ", "), "\n")
```

## Step 1: Define a single age contrast

`ordered_age_levels` is reused from `WGCNA_24_hdwgcna_HVG.Rmd` (same chronological order used there for `mouse_age`). Instead of generating every pairwise age combination (10 contrasts x 2 directions in the original version), **exactly one contrast is analyzed here: oldest vs. youngest** -- the two extremes of the aging trajectory. This keeps `multi_nichenet_analysis()` fast, and keeps every downstream section (circos plot, LR dotplot, WGCNA-overlap validation, hub-gene ligand activity) consistently about the same single comparison, instead of mixing results from 10 different age-vs-age models.

If you want a different pair of ages, edit `contrast_age_pair` below (it must contain exactly 2 values, both present in `ordered_age_levels`) -- everything downstream (contrast string, `youngest_age`/`oldest_age`, circos plot, validation) adapts automatically.

Before building the contrast, the age labels themselves need fixing: `multi_nichenet_analysis()` requires every group label in `colData(sce)[, group_id]` to be a syntactically valid R name (see `make.names()`), and labels such as `"1m"`, `"3m"`, `"18m"` start with a digit and therefore are not valid.

```{r mnn-fix-age-names}

# make.names() would auto-prefix "1m" -> "X1m", which is inconsistent with
# how ordered_age_levels was defined in WGCNA_24_hdwgcna_HVG.Rmd -- relabel
# explicitly instead, using the same mapping in both the age vector and the
# SCE colData.

age_recode <- setNames(
  paste0("age", ordered_age_levels),
  ordered_age_levels
)
print(age_recode)

# Relabel the chronological order already used in WGCNA
ordered_age_levels <- base::unname(age_recode[ordered_age_levels])

# Relabel the age column inside the SCE colData
colData(sce)$age <- age_recode[as.character(colData(sce)$age)]

cat("\nAge levels in colData(sce) after the fix:\n")
print(table(colData(sce)$age))

stopifnot(all(make.names(colData(sce)$age) == colData(sce)$age))

# Reference points reused later (circos plot, validation section, etc.)
youngest_age <- ordered_age_levels[1]
oldest_age   <- ordered_age_levels[length(ordered_age_levels)]
cat("\nYoungest age level:", youngest_age, "| Oldest age level:", oldest_age, "\n")

# --- SINGLE CONTRAST --------------------------------------------------------
# Only ONE age pair is analyzed: oldest vs youngest by default. Change this
# to any 2 values from ordered_age_levels if you want a different pair.
contrast_age_pair <- c(youngest_age, oldest_age)  # c(younger, older)
stopifnot(
  "contrast_age_pair must have exactly 2 elements" = length(contrast_age_pair) == 2,
  "contrast_age_pair values must be in ordered_age_levels" =
    all(contrast_age_pair %in% ordered_age_levels)
)

younger <- contrast_age_pair[1]
older   <- contrast_age_pair[2]

contrast_tbl <- tibble(
  contrast = paste0(older, "-", younger),
  group    = older
)

contrasts_oi <- paste0(
  "'", paste(contrast_tbl$contrast, collapse = "','"), "'"
)

cat("\nSingle contrast analyzed (contrasts_oi):\n"); print(contrasts_oi)
cat("\ncontrast_tbl:\n"); print(contrast_tbl)

```

`sample_id` has the same problem: values such as `"18-F-50"` start with a digit and contain hyphens, which are also invalid for `make.names()`. The original label is kept as a separate column for traceability.

```{r mnn-fix-sample-id}
# Original sample_id looks like "18-F-50": starts with a digit and has
# hyphens -> invalid for make.names(). Keep the original and create a
# valid version to use in the analysis.

colData(sce)$sample_id_original <- colData(sce)$sample_id
colData(sce)$sample_id <- make.names(colData(sce)$sample_id)

cat("Example mapping original sample_id -> valid sample_id:\n")
print(head(unique(data.frame(
  original = colData(sce)$sample_id_original,
  valid    = colData(sce)$sample_id
))))

stopifnot(all(make.names(colData(sce)$sample_id) == colData(sce)$sample_id))
stopifnot(length(unique(colData(sce)$sample_id_original)) == length(unique(colData(sce)$sample_id)))
cat("\nNumber of unique sample_id values:", length(unique(colData(sce)$sample_id)), "\n")


colData(sce)$mouse_id_original <- colData(sce)$mouse.id
colData(sce)$mouse.id <- make.names(colData(sce)$mouse.id)


stopifnot(all(make.names(colData(sce)$mouse.id) == colData(sce)$mouse.id))
stopifnot(length(unique(colData(sce)$mouse_id_original)) == length(unique(colData(sce)$mouse.id)))
cat("\nNumber of unique sample_id values:", length(unique(colData(sce)$mouse.id)), "\n")
```

## Step 2: Load the NicheNet prior model (ligand-receptor + ligand-target + signaling, mouse)

Beyond `lr_network` and `ligand_target_matrix` (already used for prioritization and hub-gene ligand activity), two more prior-model objects are loaded here: `weighted_networks` (ligand-receptor + signaling + TF-target confidence, needed for the receptor heatmap and the signaling-path inference below) and `ligand_tf_matrix` (ligand-to-TF regulatory potential, needed only for signaling-path inference). Both are the mouse versions from the same NicheNet v2 Zenodo record used for `lr_network`/`ligand_target_matrix`.

```{r mnn-prior-model}
options(timeout = 300)

lr_network <- readRDS("nichenet_prior/lr_network_mouse_21122021.rds"
) %>%
  dplyr::rename(ligand = from, receptor = to) %>%
  distinct(ligand, receptor)

ligand_target_matrix <- readRDS(
  "nichenet_prior/ligand_target_matrix_nsga2r_final_mouse.rds"
)

# Needed for: (1) the ligand-receptor heatmap (Step 10-style), and
# (2) signaling-path inference (Step 13-style), both added below.
weighted_networks <- readRDS(
  "nichenet_prior/weighted_networks_nsga2r_final_mouse.rds"
)

# Needed only for signaling-path inference (ligand -> TF -> target).
ligand_tf_matrix <- readRDS(
  "nichenet_prior/ligand_tf_matrix_nsga2r_final_mouse.rds"
)

cat("LR network:", nrow(lr_network), "ligand-receptor pairs\n")
cat("Ligand-target matrix:", dim(ligand_target_matrix)[1], "targets x",
    dim(ligand_target_matrix)[2], "ligands\n")
cat("Weighted networks: lr_sig =", nrow(weighted_networks$lr_sig), "edges | gr =",
    nrow(weighted_networks$gr), "edges\n")
cat("Ligand-TF matrix:", dim(ligand_tf_matrix)[1], "TFs x",
    dim(ligand_tf_matrix)[2], "ligands\n")
```

## Steps 3-6: pseudobulk + per-celltype DE, ligand activity and prioritization

`multi_nichenet_analysis()` is the wrapper that chains steps 3 to 6 of the classic workflow (pseudobulk/abundance calculation per cell type, DE per cell type via muscat/edgeR, ligand activity analysis with the prior model, and the multi-criteria prioritization table). `min_cells` requires a minimum number of cells per sample x celltype combination for that combination to enter the analysis -- adjust if some age groups have few replicates per cell type.

**Design note on `batches`/`covariates`:** `sex` and `tech` (FACS vs. droplet) were both considered as covariate/batch terms, but checking `table(colData(sce)$age, colData(sce)$tech)` and `table(colData(sce)$age, colData(sce)$sex)` showed that in this dataset both variables are almost completely confounded with `age` (e.g. `age1m`, `age18m`, `age21m`, `age30m` are each represented by a single technique, and `age1m`/`age21m`/`age24m`/`age30m` are each represented by a single sex). Because of this, the model cannot separate an age effect from a sex/technique effect for most contrasts, so both are left out (`batches = NA`, `covariates = NA`) and this is documented here as a known limitation of the Tabula Muris Senis sampling design rather than a modeling choice.

```{r mnn-run-analysis}

multinichenet_output <- multi_nichenet_analysis(
  sce                   = sce,
  celltype_id           = "final_celltype",
  sample_id             = "sample_id",
  group_id              = "age",
  batches               = NA,
  covariates            = NA,
  lr_network            = lr_network,
  ligand_target_matrix  = ligand_target_matrix,
  contrasts_oi          = contrasts_oi,
  contrast_tbl          = contrast_tbl,
  min_cells             = 3,
  logFC_threshold       = 0.50,
  p_val_threshold       = 0.05,
  fraction_cutoff       = 0.05,
  p_val_adj             = FALSE,
  empirical_pval        = FALSE,
  top_n_target          = 250,
  verbose               = TRUE
)

saveRDS(multinichenet_output, "multinichenet_output_age.rds")
cat("Total LR pairs in the prioritization table:",
    nrow(multinichenet_output$prioritization_tables$group_prioritization_tbl), "\n")
```

```{r}
multinichenet_output<-readRDS("multinichenet_output_age.rds")
```

### Which cell types were excluded from DE, and why

`multi_nichenet_analysis()` silently drops any cell type that does not have at least 2 samples per group passing `min_cells` (it prints this to the console log, but does not return it as a structured object). This chunk reconstructs that same table directly from `colData(sce)`, so it is available for inspection without having to scroll back through the log.

```{r mnn-check-excluded-celltypes}
included_celltypes <- tryCatch(
  unique(multinichenet_output$celltype_info$sender_receiver_info$sender),
  error = function(e) character(0)
)
if (length(included_celltypes) == 0) {
  # Fallback across multinichenetr versions where the info object is structured differently
  included_celltypes <- tryCatch(
    unique(multinichenet_output$prioritization_tables$group_prioritization_tbl$sender),
    error = function(e) character(0)
  )
}

samples_passing_min_cells <- as.data.frame(colData(sce)) %>%
  count(age, final_celltype, sample_id, name = "n_cells") %>%
  filter(n_cells >= 10) %>%  # keep in sync with min_cells above
  count(age, final_celltype, name = "n_samples_passing")

celltype_status <- samples_passing_min_cells %>%
  group_by(final_celltype) %>%
  summarise(
    n_age_groups_with_2plus_samples = sum(n_samples_passing >= 2),
    n_age_groups_total              = n_distinct(age),
    .groups = "drop"
  ) %>%
  mutate(
    likely_included = final_celltype %in% included_celltypes,
    fully_covered    = n_age_groups_with_2plus_samples == n_age_groups_total
  ) %>%
  arrange(desc(n_age_groups_with_2plus_samples))

cat("Cell type coverage after the min_samples_per_group / min_cells filters:\n")
print(celltype_status)

cat("\nCell types confirmed present in multinichenet_output:",
    paste(included_celltypes, collapse = ", "), "\n")
```

From here on, `multinichenet_output`, `lr_network` and `ligand_target_matrix` are available for the two WGCNA integration sections below.

### QC: cell/sample abundance diagnostics (exported to PDF)

Sanity check before trusting the prioritization results: how many cells per sample x celltype combination actually passed `min_cells`, and how balanced is that across age groups. `multi_nichenet_analysis()` already computes this internally in `abundance_expression_info`; we reuse its built-in plot when available and fall back to a manual count table otherwise (useful given the tech/sex imbalance documented above).

```{r mnn-qc-abundance, fig.width=9, fig.height=6}
abund_info <- multinichenet_output$abundance_expression_info

p_abundance <- tryCatch(
  abund_info$abund_plot_sample,
  error = function(e) NULL
)

if (is.null(p_abundance)) {
  # Fallback: build the same kind of diagnostic manually from colData(sce)
  abundance_manual <- as.data.frame(colData(sce)) %>%
    count(sample_id, final_celltype, age, name = "n_cells")

  p_abundance <- ggplot(abundance_manual, aes(x = sample_id, y = n_cells, fill = age)) +
    geom_col() +
    facet_wrap(~ final_celltype, scales = "free_y") +
    labs(title = "Cells per sample x celltype", x = "Sample", y = "N cells") +
    theme_minimal(base_size = 9) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 6))
}

print(p_abundance)
ggsave("plot_QC_abundance_per_sample_celltype.pdf", p_abundance, width = 9, height = 6)
```

### Overview plot: cell-cell communication summary (exported to PDF)

A first exploratory summary across all age contrasts pooled together: how many distinct LR pairs were prioritized per sender-receiver cell type pair.

```{r mnn-plot-sender-receiver-overview, fig.width=8, fig.height=6}
sender_receiver_overview <- multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  distinct(group, sender, receiver, ligand, receptor) %>%
  count(sender, receiver, name = "n_LR_pairs")

p_sender_receiver <- ggplot(sender_receiver_overview, aes(x = sender, y = receiver, fill = n_LR_pairs)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
  labs(
    title = "Cell-cell communication overview",
    subtitle = "Number of distinct LR pairs per sender-receiver cell type pair (contrast age30m-age3m)",
    x = "Sender cell type", y = "Receiver cell type", fill = "N LR pairs"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_sender_receiver)

ggsave("plot_sender_receiver_overview.pdf", p_sender_receiver, width = 8, height = 6)
```

### Circos plot and LR dotplot for the oldest-vs-youngest contrast

Two of MultiNicheNet's signature plots for a single contrast: a circos diagram of sender -\> receiver communication, and a dotplot combining expression, ligand activity and logFC for the top LR pairs. Function names below follow the standard multinichenetr API; if your installed package version differs, run `ls("package:multinichenetr")` and adjust the function names accordingly -- these plotting helpers have changed across versions.

```{r}
library(dplyr)
library(ggplot2)
library(circlize)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(magick)  # para rasterizar los PDFs individuales

graphics.off()

# ==============================================================================
# 1. REORDER AGES IN MULTINICHENET (age3m AT THE START)
# ==============================================================================
age_order <- ordered_age_levels  # dynamic (was hardcoded to a fixed set of 5 age strings)

multinichenet_output$prioritization_tables <- lapply(
  multinichenet_output$prioritization_tables, function(tbl) {
    if (is.data.frame(tbl) && "group" %in% colnames(tbl)) {
      tbl$group <- factor(tbl$group, levels = age_order)
    }
    return(tbl)
  }
)

if (!is.null(multinichenet_output$celltype_info$sample_id_info)) {
  if ("group" %in% colnames(multinichenet_output$celltype_info$sample_id_info)) {
    multinichenet_output$celltype_info$sample_id_info$group <- factor(
      multinichenet_output$celltype_info$sample_id_info$group, 
      levels = age_order
    )
  }
}

# ==============================================================================
# 2. EXTRACT TOP L-R PAIRS FOR CONTRAST
# ==============================================================================
contrast_oi <- paste0(oldest_age, "-", youngest_age)

get_lr_pairs_for_contrast <- function(contrast_str, top_n = 50) {
  tbl <- multinichenet_output$prioritization_tables$group_prioritization_tbl
  stopifnot("contrast" %in% colnames(tbl))
  
  rank_col <- intersect(c("prioritization_score", "prioritization_rank"), colnames(tbl))
  tbl_contrast <- tbl %>% dplyr::filter(contrast == contrast_str)
  
  if (length(rank_col) > 0) {
    tbl_contrast <- tbl_contrast %>% 
      dplyr::arrange(if (rank_col[1] == "prioritization_rank") prioritization_rank else desc(.data[[rank_col[1]]]))
  }
  
  tbl_contrast %>% 
    dplyr::distinct(sender, receiver, ligand, receptor, .keep_all = TRUE) %>% 
    dplyr::slice_head(n = top_n)
}

lr_prioritized_50 <- get_lr_pairs_for_contrast(contrast_oi, top_n = 50)

cat("Interactions retrieved for '", contrast_oi, "': ", nrow(lr_prioritized_50), "\n")

# ==============================================================================
# 3. UNIFIED CELL-TYPE COLOR PALETTE
# ==============================================================================
all_senders   <- unique(lr_prioritized_50$sender)
all_receivers <- unique(lr_prioritized_50$receiver)
all_celltypes <- unique(c(all_senders, all_receivers))

palette_colors <- RColorBrewer::brewer.pal(max(3, length(all_celltypes)), "Set2")[1:length(all_celltypes)]
names(palette_colors) <- all_celltypes

colors_sender_vec   <- palette_colors[all_senders]
colors_receiver_vec <- palette_colors[all_receivers]

# ==============================================================================
# 4. SPLIT TOP 50 INTO 3 CHUNKS
# ==============================================================================
n_total    <- nrow(lr_prioritized_50)
n_chunks   <- 3
chunk_size <- ceiling(n_total / n_chunks)

circos_part_list <- list()
for (i in 1:n_chunks) {
  start_idx <- (i - 1) * chunk_size + 1
  end_idx   <- min(i * chunk_size, n_total)
  
  if (start_idx <= n_total) {
    tbl_i <- lr_prioritized_50 %>% 
      dplyr::slice(start_idx:end_idx) %>% 
      dplyr::select(-dplyr::any_of(c("weight", "weight.x", "weight.y")))
    colnames(tbl_i) <- make.unique(colnames(tbl_i))
    circos_part_list[[paste0("part", i)]] <- tbl_i
  }
}

# ==============================================================================
# 5. RENDERIZAR CADA CÍRCULO EN SU PROPIO PDF (método que YA sabemos que funciona)
# ==============================================================================
cat("\n--- Rendering individual circos plots (as single-page PDFs) ---\n")

tmp_dir   <- tempdir()
pdf_paths <- c()

for (i in seq_along(circos_part_list)) {
  
  part_name <- names(circos_part_list)[i]
  tbl_i     <- circos_part_list[[part_name]]
  pdf_path  <- file.path(tmp_dir, paste0("circos_", part_name, ".pdf"))
  
  pdf(pdf_path, width = 8, height = 8)
  
  result <- tryCatch({
    par(mar = c(3, 3, 4, 3))
    circlize::circos.clear()
    
    make_circos_group_comparison(
      prioritized_tbl_oi = tbl_i,
      colors_sender       = colors_sender_vec,
      colors_receiver     = colors_receiver_vec
    )
    
    title(
      main = paste0("Part ", i, " [", nrow(tbl_i), " pairs]"),
      line = 1,
      cex.main = 1.6,
      font.main = 2
    )
    
    TRUE
  }, error = function(e) {
    message("ERROR en ", part_name, ": ", conditionMessage(e))
    FALSE
  })
  
  dev.off()
  
  if (isTRUE(result)) {
    cat(part_name, "renderizado correctamente (", nrow(tbl_i), "pares )\n")
    pdf_paths <- c(pdf_paths, pdf_path)
  } else {
    cat(part_name, "NO se pudo renderizar — revisa el error de arriba\n")
  }
}

# ==============================================================================
# 6. RASTERIZAR CADA PDF INDIVIDUAL A IMAGEN DE ALTA RESOLUCIÓN (con magick)
# ==============================================================================
cat("\n--- Converting individual PDFs to high-res images ---\n")

circle_grobs <- lapply(pdf_paths, function(p) {
  img <- magick::image_read_pdf(p, density = 300)
  grid::rasterGrob(as.raster(img), interpolate = TRUE)
})

# ==============================================================================
# 7. RENDERIZAR LA LEYENDA (una sola vez) COMO SU PROPIO PDF -> IMAGEN
# ==============================================================================
legend_pdf_path <- file.path(tmp_dir, "circos_legend.pdf")

pdf(legend_pdf_path, width = 2.5, height = 8)
par(mar = c(1, 1, 1, 1))
plot.new()
legend(
  "center",
  legend = names(palette_colors),
  fill = palette_colors,
  border = NA,
  bty = "n",
  title = "Cell Types",
  title.font = 2,
  cex = 1.3,
  y.intersp = 1.6
)
dev.off()

legend_img  <- magick::image_read_pdf(legend_pdf_path, density = 300)
legend_grob <- grid::rasterGrob(as.raster(legend_img), interpolate = TRUE)

# ==============================================================================
# 8. ENSAMBLAR: CÍRCULOS EN ORDEN A LA IZQUIERDA + LEYENDA UNA SOLA VEZ A LA DERECHA
# ==============================================================================
cat("\n--- Assembling final single-page PDF ---\n")

circos_pdf_filename <- "plot_circos_3panels_top50.pdf"

if (length(circle_grobs) == 0) {
  stop("Ningún círculo se renderizó correctamente. Revisa los mensajes de error de arriba antes de continuar.")
}

pdf(circos_pdf_filename, width = 24, height = 7.5)

grid.arrange(
  grobs = c(circle_grobs, list(legend_grob)),
  ncol  = length(circle_grobs) + 1,
  widths = c(rep(3, length(circle_grobs)), 1),
  top = grid::textGrob(
    paste0("Top L-R Interactions (", contrast_oi, ")"),
    gp = grid::gpar(fontsize = 16, fontface = "bold")
  )
)

dev.off()



# ==============================================================================
# 9. LR DOTPLOTS (TOP 50 SPLIT INTO 3 PARTS) — SIN CAMBIOS
# ==============================================================================
cat("\n--- Generating LR Dotplots (Top 50 split into 3 parts) ---\n")

for (part_name in names(circos_part_list)) {
  i <- as.integer(sub("part", "", part_name))
  start_idx <- (i - 1) * chunk_size + 1
  end_idx   <- min(i * chunk_size, n_total)
  tbl_sub   <- lr_prioritized_50 %>% dplyr::slice(start_idx:end_idx)
  
  p_sub <- tryCatch(
    make_sample_lr_prod_activity_plots(
      prioritization_tables = multinichenet_output$prioritization_tables,
      prioritized_tbl_oi    = tbl_sub
    ),
    error = function(e) {
      message(" Error in Dotplot of ", part_name, ": ", conditionMessage(e))
      NULL
    }
  )
  
  if (!is.null(p_sub)) {
    p_sub_wide <- p_sub & 
      theme(
        axis.text.x = element_text(size = 6.5, angle = 45, hjust = 0, vjust = 0.5),
        axis.text.y = element_text(size = 7.5),
        strip.text  = element_text(size = 7.5, face = "bold"),
        plot.margin = margin(t = 35, r = 15, b = 15, l = 15, unit = "pt")
      )
    
    dotplot_pdf_name <- paste0("plot_LR_dotplot_", part_name, "_ultra_wide.pdf")
    ggsave(dotplot_pdf_name, p_sub_wide, width = 22, height = 13, limitsize = FALSE)
    cat("LR Dotplot saved as:", dotplot_pdf_name, "\n")
  }
}

```

## Validation: overlap between age-associated WGCNA modules and MultiNicheNet prioritization

Idea: if a ligand or receptor prioritized by MultiNicheNet as relevant for the age contrast **also** falls inside a WGCNA module that the independent WGCNA analysis already flagged as age-associated -- and, moreover, in the **same cell type** (sender for the ligand, receiver for the receptor) -- that is cross-method evidence from two approaches that start from different information (co-expression topology vs. ligand-receptor prior model + DE).

```{r}

# ============================================================
# 0. NUEVO -- combinar correlación lineal (significant_modules)
#    y trayectoria (significant_trajectory_modules) en un solo
#    criterio de "módulo asociado a edad". Va ANTES de todo lo
#    demás que dependa de significant_modules.
# ============================================================
age_association_summary <- module_trait_table %>%
  dplyr::select(module, correlation, p_value_corr = p_value, fdr_corr = fdr) %>%
  full_join(
    module_trajectory %>%
      dplyr::select(module, shape, category, p_value_traj = p_value,
                    fdr_traj = fdr, r_squared),
    by = "module"
  ) %>%
  mutate(
    sig_correlation = !is.na(fdr_corr) & fdr_corr < 0.05,
    sig_trajectory  = !is.na(fdr_traj) & fdr_traj < 0.05,
    age_associated  = sig_correlation | sig_trajectory
  ) %>%
  arrange(desc(age_associated), fdr_corr, fdr_traj)

age_associated_modules <- age_association_summary %>%
  filter(age_associated) %>%
  pull(module)

cat("Módulos asociados a edad (correlación lineal y/o trayectoria):",
    length(age_associated_modules), "\n")
print(age_association_summary %>% filter(age_associated) %>%
        dplyr::select(module, correlation, fdr_corr, shape, category, fdr_traj))

```

```{r multinichenet-validation-setup}

# Same fix as the circos/dotplot section above: filter by the exact
# `contrast` string, not by the pooled `group` label, which would otherwise
# mix several different age comparisons together.
age_group_oi <- oldest_age  # kept for readable labels/titles below

# Target the single contrast defined in Step 1 (dynamic, not hardcoded --
# with contrast_tbl now holding exactly one row, this is that one contrast).
contrast_oi <- contrast_tbl$contrast[1]

lr_prioritized <- multinichenet_output$prioritization_tables$group_prioritization_tbl %>% dplyr::filter(contrast == contrast_oi) %>% dplyr::slice_max(order_by = prioritization_score, n = 250)
```

```{r multinichenet-validation-crossref}

# ============================================================
module_ct_lookup <- gene_celltype_module %>%
  filter(module %in% age_associated_modules) %>%   # antes: significant_modules
  distinct(module, celltype, gene)

lr_prioritized_annotated <- lr_prioritized %>%
  left_join(
    module_ct_lookup %>% rename(ligand = gene, sender = celltype, ligand_module = module),
    by = c("ligand", "sender")
  ) %>%
  left_join(
    module_ct_lookup %>% rename(receptor = gene, receiver = celltype, receptor_module = module),
    by = c("receptor", "receiver")
  )

overlap_summary <- lr_prioritized_annotated %>%
  summarise(
    n_total_pairs            = n(),
    n_ligand_in_age_module   = sum(!is.na(ligand_module)),
    n_receptor_in_age_module = sum(!is.na(receptor_module)),
    n_both_in_age_module     = sum(!is.na(ligand_module) & !is.na(receptor_module))
  )

cat("Overlap summary (age-prioritized LR pairs vs. age-associated WGCNA modules):\n")
print(overlap_summary)
write.csv(lr_prioritized_annotated, "multinichenet_age_LR_wgcna_overlap.csv", row.names = FALSE)
```

### Is the overlap greater than expected by chance?

Global hypergeometric test: of all ligand/receptor genes prioritized by MultiNicheNet for the age contrast (that were also part of the tested universe in WGCNA), how many fall in an age-associated module, compared to the proportion expected under random assignment?

```{r multinichenet-validation-hypergeom}

age_module_genes <- gene_celltype_module %>%
  filter(module %in% age_associated_modules) %>%   # antes: significant_modules
  distinct(gene) %>%
  pull(gene)

lr_genes_tested <- unique(c(lr_prioritized$ligand, lr_prioritized$receptor))
lr_genes_tested <- intersect(lr_genes_tested, universe_symbols)

n_overlap  <- length(intersect(lr_genes_tested, age_module_genes))
n_age_mod  <- length(intersect(age_module_genes, universe_symbols))
n_universe <- length(universe_symbols)
n_lr       <- length(lr_genes_tested)

p_enrich <- phyper(n_overlap - 1, n_age_mod, n_universe - n_age_mod, n_lr, lower.tail = FALSE)

cat("MultiNicheNet ligand/receptor genes also present in the WGCNA network:", n_lr, "\n")
cat("From which,", n_overlap, "fall into a WGCNA module asociated with age \n")
cat("Hipergeometric enrichment, p =", signif(p_enrich, 3), "\n")
```

```{r multinichenet-validation-plot, fig.width=7, fig.height=5}

plot_data <- bind_rows(
  lr_prioritized_annotated %>% filter(!is.na(ligand_module)) %>%
    transmute(module = ligand_module, role = "Ligand"),
  lr_prioritized_annotated %>% filter(!is.na(receptor_module)) %>%
    transmute(module = receptor_module, role = "Receptor")
)

p_validation <- ggplot(plot_data, aes(x = module, fill = role)) +
  geom_bar(position = "dodge") +
  labs(x = "WGCNA module (age-associated)", y = "Number of prioritized LR pairs",
       title = "MultiNicheNet LR pairs falling in age-associated WGCNA modules",
       subtitle = paste0("Contrast: ", age_group_oi, " | global hypergeometric p = ", signif(p_enrich, 3))) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_validation)

ggsave("plot_LR_wgcna_module_overlap.pdf", p_validation, width = 7, height = 5)
```

### Validation across contrasts (generic, works for 1..N contrasts)

The spotlight above covered the single contrast defined in Step 1 (`r contrast_oi`). The same overlap + hypergeometric logic is generalized here and applied to **every contrast in `contrast_tbl$contrast`** -- iterating over `contrast`, not the pooled `group` label, since a single age level can be the "group" of several different comparisons at once (see the note in the circos/dotplot section above). With the pipeline now restricted to a single age contrast (Step 1), this loop runs over exactly that one contrast and reproduces the spotlight result above; the code is kept generic (rather than hardcoded to one contrast) so it keeps working unchanged if you ever add more rows to `contrast_tbl`.

```{r}


# 1. Función para extraer los pares usando prioritization_tables
get_lr_pairs_for_contrast <- function(contrast_str, multinichenet_res, top_n = 250) {
  
  # Acceso directo a prioritization_tables (estructura estándar de MultiNicheNet)
  # Si prioritization_tables es una lista de tablas, toma group_prioritization_tbl;
  # si es el data frame directo, lo asigna.
  prior_tbl <- if (is.data.frame(multinichenet_res$prioritization_tables)) {
    multinichenet_res$prioritization_tables
  } else if (!is.null(multinichenet_res$prioritization_tables$group_prioritization_tbl)) {
    multinichenet_res$prioritization_tables$group_prioritization_tbl
  } else {
    NULL
  }
  
  if (is.null(prior_tbl)) {
    warning(paste("No se encontró la tabla de priorización para el contraste:", contrast_str))
    return(NULL)
  }
  
  # Detectar automáticamente si la columna se llama prioritization_score o score
  score_col <- if ("prioritization_score" %in% colnames(prior_tbl)) {
    "prioritization_score"
  } else if ("score" %in% colnames(prior_tbl)) {
    "score"
  } else {
    colnames(prior_tbl)[grep("score|rank", colnames(prior_tbl))[1]]
  }
  
  # Filtrar los top N pares por contraste
  lr_top <- prior_tbl %>%
    filter(contrast == contrast_str) %>%
    slice_max(order_by = .data[[score_col]], n = top_n, with_ties = FALSE)
  
  return(lr_top)
}

# 2. Función de validación por contraste
run_validation_for_contrast <- function(contrast_str, multinichenet_res) {
  
  lr_oi <- get_lr_pairs_for_contrast(
    contrast_str = contrast_str, 
    multinichenet_res = multinichenet_res, 
    top_n = 250
  )
  
  # Retornar tibble estructurado si no hay datos
  if (is.null(lr_oi) || nrow(lr_oi) == 0) {
    return(tibble(
      contrast                 = contrast_str,
      n_total_pairs            = 0,
      n_ligand_in_age_module   = 0,
      n_receptor_in_age_module = 0,
      n_lr_genes_tested        = 0,
      n_lr_genes_in_age_module = 0,
      p_hypergeometric         = NA_real_
    ))
  }
  
  # Anotación con los módulos por celda/gen
  lr_oi_annotated <- lr_oi %>%
    left_join(
      module_ct_lookup %>% rename(ligand = gene, sender = celltype, ligand_module = module),
      by = c("ligand", "sender")
    ) %>%
    left_join(
      module_ct_lookup %>% rename(receptor = gene, receiver = celltype, receptor_module = module),
      by = c("receptor", "receiver")
    )
  
  # Genes únicos involucrados en las interacciones
  lr_genes_oi <- unique(c(lr_oi$ligand, lr_oi$receptor))
  lr_genes_oi <- intersect(lr_genes_oi, universe_symbols)
  
  n_overlap_oi <- length(intersect(lr_genes_oi, age_module_genes))
  n_lr_oi      <- length(lr_genes_oi)
  
  # Test hipergeométrico
  p_oi <- if (n_lr_oi > 0) {
    phyper(n_overlap_oi - 1, n_age_mod, n_universe - n_age_mod, n_lr_oi, lower.tail = FALSE)
  } else {
    NA_real_
  }
  
  tibble(
    contrast                 = contrast_str,
    n_total_pairs            = nrow(lr_oi),
    n_ligand_in_age_module   = sum(!is.na(lr_oi_annotated$ligand_module)),
    n_receptor_in_age_module = sum(!is.na(lr_oi_annotated$receptor_module)),
    n_lr_genes_tested        = n_lr_oi,
    n_lr_genes_in_age_module = n_overlap_oi,
    p_hypergeometric         = p_oi
  )
}

# 3. Ejecución corregida sobre todos los contrastes
validation_all_contrasts <- purrr::map_dfr(
  unique(contrast_tbl$contrast), 
  ~ run_validation_for_contrast(.x, multinichenet_res = multinichenet_output)
) %>%
  mutate(
    fdr_hypergeometric = if (all(is.na(p_hypergeometric))) {
      NA_real_
    } else {
      p.adjust(p_hypergeometric, method = "BH")
    }
  ) %>%
  arrange(p_hypergeometric)
```

### GO enrichment of the LR / age-module overlap gene set (reused from WGCNA_24)

No new enrichment is run here. `WGCNA_24.Rmd` already computed GO Biological Process enrichment per module -- `go_full_module_df` (all module genes) and `go_hubs_df` (hub genes only), both already restricted to `pvalueCutoff = 0.05` and BH-adjusted -- and both objects are still in the environment from that earlier run. This section just **filters** those precomputed results down to the terms whose gene set (`geneID`, slash-separated gene symbols since `readable = TRUE` was used) overlaps with the genes that are *both* LR-prioritized by MultiNicheNet *and* already in an age-associated module (`overlap_genes`, same definition as the hypergeometric test above), restricted to `significant_modules`.

```{r}

library(tidyverse)

# Resolver conflictos de nombres entre tidyverse y paquetes tipo Bioconductor/lubridate
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select, .quiet = TRUE)
  conflicts_prefer(dplyr::filter, .quiet = TRUE)
  conflicts_prefer(dplyr::mutate, .quiet = TRUE)
  conflicts_prefer(dplyr::arrange, .quiet = TRUE)
  conflicts_prefer(base::intersect, .quiet = TRUE)
}

overlap_genes <- lr_genes_tested[lr_genes_tested %in% age_module_genes]
cat("Genes entrando al filtro GO (LR-priorizados Y en un módulo asociado a edad):",
    length(overlap_genes), "\n")

filter_precomputed_go <- function(go_df, source_label) {
  if (is.null(go_df) || nrow(go_df) == 0) return(NULL)
  
  go_df %>%
    dplyr::filter(module %in% age_associated_modules) %>%   
    dplyr::filter(!is.na(geneID) & geneID != "") %>%
    dplyr::mutate(
      # str_split devuelve una lista directamente sin requerir índices numéricos arriesgados
      current_genes     = stringr::str_split(as.character(geneID), "/"),
      intersect_genes   = purrr::map(current_genes, ~ base::intersect(.x, overlap_genes)),
      n_overlap_genes   = purrr::map_int(intersect_genes, length),
      overlap_hit_genes = purrr::map_chr(intersect_genes, ~ paste(.x, collapse = "/"))
    ) %>%
    dplyr::filter(n_overlap_genes > 0) %>%
    dplyr::mutate(go_source = source_label) %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::select(-current_genes, -intersect_genes)
}

go_overlap_full <- filter_precomputed_go(go_bp_by_module_celltype, "GO_BP (todos los genes del módulo)")

hub_lookup <- hub_genes_table %>% dplyr::distinct(module, celltype, gene)

go_overlap_hubs <- go_overlap_full %>%
  dplyr::mutate(
    # Aplicamos la misma lógica segura con stringr para evitar subíndices fuera de límites
    split_hits    = stringr::str_split(ifelse(is.na(overlap_hit_genes), "", overlap_hit_genes), "/"),
    target_hubs   = purrr::map2(module, celltype, ~ hub_lookup$gene[hub_lookup$module == .x & hub_lookup$celltype == .y]),
    intersect_hub = purrr::map2(split_hits, target_hubs, base::intersect),
    hub_hit_genes = purrr::map_chr(intersect_hub, ~ paste(.x, collapse = "/"))
  ) %>%
  dplyr::filter(hub_hit_genes != "") %>%
  dplyr::mutate(go_source = "GO_BP (subset con hub genes)") %>%
  dplyr::select(-split_hits, -target_hubs, -intersect_hub)

go_overlap_combined <- dplyr::bind_rows(go_overlap_hubs, go_overlap_full)

cat("Términos GO BP (de WGCNA_24) que tocan el set de solape LR/módulo:",
    nrow(go_overlap_combined), "\n")

write.csv(go_overlap_combined, "GO_LR_wgcna_overlap_from_precomputed.csv", row.names = FALSE)

go_enrich_plot <- NULL
if (!is.null(go_overlap_combined) && nrow(go_overlap_combined) > 0) {
  
  go_overlap_top <- go_overlap_combined %>%
    dplyr::group_by(go_source) %>%
    dplyr::slice_min(p.adjust, n = 15, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      # Etiqueta más corta y envuelta en varias líneas para que no aplaste el panel
      label = stringr::str_wrap(paste(Description, celltype, sep = " | "), width = 45)
    )
  
  cat("Distribución de términos por go_source:\n")
  print(table(go_overlap_top$go_source))
  
  go_enrich_plot <- ggplot(
    go_overlap_top,
    aes(x = reorder(label, -p.adjust),
        y = -log10(p.adjust), size = n_overlap_genes, color = module)
  ) +
    geom_point(alpha = 0.8) +
    coord_flip() +
    facet_wrap(~ go_source, scales = "free_y", ncol = 1) +
    scale_size_continuous(range = c(2, 7)) +
    labs(x = NULL, y = "-log10(p.adjust)",
         color = "Módulo", size = "N genes solapados",
         title = "Términos GO (BP) de WGCNA_24 que tocan el set de solape LR/WGCNA",
         subtitle = "Enrichment reutilizado tal cual de WGCNA_24.Rmd -- no recalculado aquí") +
    theme_minimal(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11, margin = margin(b = 4)),
      plot.subtitle = element_text(size = 8, color = "gray40", margin = margin(b = 10)),
      axis.text.y = element_text(size = 7.5, lineheight = 0.85),
      strip.text = element_text(face = "bold", size = 9),
      legend.position = "right",
      plot.margin = margin(t = 10, r = 15, b = 10, l = 10)
    )
  
  print(go_enrich_plot)
  
  # Alto dinámico según número de términos y facetas, para que no se aplasten
  n_terms     <- nrow(go_overlap_top)
  n_sources   <- dplyr::n_distinct(go_overlap_top$go_source)
  plot_height <- max(6, n_terms / n_sources * 0.35 * n_sources)
  
  ggsave("plot_GO_LR_wgcna_overlap_from_precomputed.pdf",
         go_enrich_plot, width = 11, height = plot_height, limitsize = FALSE)
  
} else {
  cat("Ninguno de los términos GO BP precomputados (WGCNA_24) contiene un gen del set de solape LR/módulo.\n")
}
```

```{r}
library(tidyverse)

# Resolver conflictos de nombres entre tidyverse y otros paquetes cargados
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select, .quiet = TRUE)
  conflicts_prefer(dplyr::filter, .quiet = TRUE)
  conflicts_prefer(dplyr::mutate, .quiet = TRUE)
  conflicts_prefer(dplyr::arrange, .quiet = TRUE)
  conflicts_prefer(dplyr::count, .quiet = TRUE)
  conflicts_prefer(base::intersect, .quiet = TRUE)
}

module_celltype_candidates <- hub_genes_table %>%
  dplyr::filter(celltype %in% target_celltypes) %>%
  dplyr::count(module, celltype, name = "n_hub_genes") %>%
  dplyr::inner_join(age_association_summary, by = "module") %>%   # antes: module_trait_table + filter(fdr<0.05)
  dplyr::filter(age_associated) %>%
  dplyr::mutate(abs_correlation = abs(correlation)) %>%
  dplyr::arrange(fdr_corr, fdr_traj, desc(n_hub_genes))

cat("Candidatos (module x celltype) asociados a edad:\n")
print(module_celltype_candidates)
write.csv(module_celltype_candidates, "module_celltype_candidates_prelim.csv", row.names = FALSE)
cat("Candidatos (version preliminar) exportados a module_celltype_candidates_prelim.csv\n")

SELECTION_METRIC <- "fdr_corr"  # "fdr_corr" | "abs_correlation" | "fdr_traj" | "r_squared"

top_hit <- switch(
  SELECTION_METRIC,
  fdr_corr        = module_celltype_candidates %>% dplyr::filter(sig_correlation) %>% dplyr::slice_min(fdr_corr, n = 1, with_ties = FALSE),
  abs_correlation = module_celltype_candidates %>% dplyr::filter(sig_correlation) %>% dplyr::slice_max(abs_correlation, n = 1, with_ties = FALSE),
  fdr_traj        = module_celltype_candidates %>% dplyr::filter(sig_trajectory) %>% dplyr::slice_min(fdr_traj, n = 1, with_ties = FALSE),
  r_squared       = module_celltype_candidates %>% dplyr::filter(sig_trajectory) %>% dplyr::slice_max(r_squared, n = 1, with_ties = FALSE),
  stop("SELECTION_METRIC debe ser 'fdr_corr', 'abs_correlation', 'fdr_traj' o 'r_squared'")
)

module_oi   <- top_hit$module
receiver_oi <- top_hit$celltype
```

## Mechanistic hypothesis: age-associated module hub genes as the target gene set

Instead of using MultiNicheNet's standard DE genes as the gene set of interest in the ligand activity analysis, here we use the **hub genes** (top kME) of a specific module, restricted to the cell type where that module's genes originate. This directly asks: *which ligands, from which candidate cell types, could be driving the co-expression program that WGCNA found in this cell type?* This whole section follows the classic NicheNet receiver-centric logic (Steps 2-3, 5-11, 13 of the [ngs101.com NicheNet tutorial](https://ngs101.com/how-to-analyze-single-cell-rna-seq-data-complete-beginners-guide-part-10-cell-cell-communication-analysis-using-nichenet/)), adapted so the receiver, the gene set of interest, and the receiver/sender cell types all come from WGCNA rather than from a fresh DE test.

Note on `module_oi`/`receiver_oi` selection: WGCNA's modules are built from a **single combined network** of `gene_celltype` features (all `target_celltypes` mixed together), so a module-age correlation (`module_trait_table`) exists only **per module**, not per (module, celltype) -- there is no separate age-correlation value for "this module in Fibroblasts" vs. "this module in Myocytes". What legitimately varies by cell type is *which* cell types actually contributed hub genes to a given significant module (`hub_genes_table$celltype`). `module_celltype_candidates` below combines both pieces of information correctly: the module-level FDR/correlation from `module_trait_table`, crossed with every (module, celltype) combination that actually has hub genes, restricted to `target_celltypes` (so senders/receiver both stay within the same 4 cell types WGCNA and MultiNicheNet were restricted to).

### Step 3 (WGCNA-adapted): real expressed genes per cell type

The previous version of this section used `gene_celltype_module$gene` (the \~20K WGCNA age-dispersion HVGs) as a stand-in for "genes expressed in this cell type", for both the background gene universe and the sender ligand pool. That is too narrow: those \~20K genes are the ones WGCNA *selected* as highly age-variable, not the full set of genes actually expressed in each cell type, and using them as the ligand-activity background biases the corrected AUPR statistic (a smaller, non-random background changes what "enriched relative to background" means). Following Step 3 of the NicheNet tutorial, real expressed genes are computed here with `nichenetr::get_expressed_genes()` (\>= 10% of a cell type's cells, the same convention the tutorial uses) directly on `seurat_mnn` -- the exact same cells/cell types WGCNA and MultiNicheNet both use. WGCNA's gene set (`gene_celltype_module`) is still used for what it's actually meant for: defining `geneset_oi`, the hub genes themselves.

```{r}
library(tidyverse)

# Resolver conflictos de nombres entre tidyverse y otros paquetes cargados
if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select, .quiet = TRUE)
  conflicts_prefer(dplyr::filter, .quiet = TRUE)
  conflicts_prefer(dplyr::mutate, .quiet = TRUE)
  conflicts_prefer(dplyr::arrange, .quiet = TRUE)
  conflicts_prefer(dplyr::count, .quiet = TRUE)
  conflicts_prefer(base::intersect, .quiet = TRUE)
  conflicts_prefer(base::union, .quiet = TRUE)
}

EXPRESSION_PCT <- 0.10  # same convention as the NicheNet tutorial (>=10% of cells)

expr_mat_mnn    <- Seurat::GetAssayData(seurat_mnn, assay = "RNA", layer = "counts")
cell_idents_mnn <- seurat_mnn$final_celltype

expressed_genes_by_celltype <- setNames(
  lapply(target_celltypes, function(ct) {
    nichenetr::get_expressed_genes(
      celltype_oi    = ct,
      object         = expr_mat_mnn,
      celltype_annot = cell_idents_mnn,
      pct            = EXPRESSION_PCT
    )
  }),
  target_celltypes
)

cat("Real expressed genes per cell type (>=", EXPRESSION_PCT * 100, "% of cells):\n")
print(vapply(expressed_genes_by_celltype, length, integer(1)))

# Union across all 4 target_celltypes = the real "expressed_genes_sender" pool
# used by the sender-agnostic ligand activity approach further below.
expressed_genes_union <- Reduce(base::union, expressed_genes_by_celltype)
cat("\nUnion of expressed genes across all target_celltypes:", length(expressed_genes_union), "\n")

rm(expr_mat_mnn, cell_idents_mnn)
```

### Setup: module/celltype selection, gene set of interest, real background

```{r hub-genes-ligand-activity-setup}
# Build the (module x celltype) candidate table: module-level significance
# from module_trait_table, crossed with which cell types actually have hub
# genes in that module (from hub_genes_table), restricted to target_celltypes.
module_celltype_candidates <- hub_genes_table %>%
  filter(celltype %in% target_celltypes) %>%
  count(module, celltype, name = "n_hub_genes") %>%
  inner_join(
    module_trait_table %>% dplyr::select(module, correlation, p_value, fdr),
    by = "module"
  ) %>%
  filter(p_value < 0.05) %>%
  mutate(abs_correlation = abs(correlation)) %>%
  arrange(fdr, desc(n_hub_genes))

cat("Significant (module x celltype) candidates (FDR < 0.05, restricted to target_celltypes):\n")
print(module_celltype_candidates)
write.csv(module_celltype_candidates, "module_celltype_candidates.csv", row.names = FALSE)
cat("Candidatos exportados a module_celltype_candidates.csv\n")

# Which criterion to rank candidates by, before picking the top one:
#  - "fdr"              : most statistically confident module-age association (default)
#  - "abs_correlation"   : strongest module-age effect size, regardless of FDR rank
SELECTION_METRIC <- "fdr"  # change to "abs_correlation" to select by effect size instead

top_hit <- switch(
  SELECTION_METRIC,
  fdr             = module_celltype_candidates %>% slice_min(fdr, n = 1, with_ties = FALSE),
  abs_correlation = module_celltype_candidates %>% slice_max(abs_correlation, n = 1, with_ties = FALSE),
  stop("SELECTION_METRIC must be 'fdr' or 'abs_correlation'")
)
# To force a specific module/celltype manually instead, uncomment and edit:
# top_hit <- module_celltype_candidates %>% filter(module == "turquoise", celltype == "Fibroblasts")

module_oi   <- top_hit$module
receiver_oi <- top_hit$celltype

cat("\nSelection metric:", SELECTION_METRIC, "\n")
cat("Module of interest:", module_oi, "| Receiver cell type:", receiver_oi,
    "| Module-age FDR:", signif(top_hit$fdr, 3),
    "| Module-age correlation:", signif(top_hit$correlation, 3), "\n")

# Hub genes of THAT module, selected specifically from THAT cell type
# (the `celltype` column in hub_genes_table = origin celltype of the gene_celltype feature)
geneset_oi <- hub_genes_table %>%
  filter(module == module_oi, celltype == receiver_oi) %>%
  pull(gene) %>%
  unique()

geneset_oi <- geneset_oi[geneset_oi %in% rownames(ligand_target_matrix)]
cat("Hub genes used as the target gene set:", length(geneset_oi), "\n")

# The tutorial recommends 50-500 genes for a stable ligand activity signal;
# with a small module this can easily fall well below that. Warn loudly
# rather than silently reporting an AUPR built on very few genes.
MIN_GENESET_SIZE <- 20
if (length(geneset_oi) < MIN_GENESET_SIZE) {
  warning(
    "geneset_oi has only ", length(geneset_oi), " genes (< ", MIN_GENESET_SIZE,
    "). Ligand activity (AUPR) estimates from this few genes are unstable -- ",
    "consider a larger/less strict module, a lower hub_percentile in WGCNA_24, ",
    "or interpreting the ranking below qualitatively rather than quantitatively."
  )
}

# REAL background: genes actually expressed in the receiver cell type
# (>=10% of its cells, computed above), NOT just the WGCNA-selected HVGs --
# this is the statistically correct background for predict_ligand_activities().
background_expressed_genes <- expressed_genes_by_celltype[[receiver_oi]] %>%
  intersect(rownames(ligand_target_matrix))

cat("Real background expressed genes in", receiver_oi, ":", length(background_expressed_genes), "\n")
cat("(for comparison, WGCNA-HVG-only background would have been:",
    length(intersect(unique(gene_celltype_module$gene[gene_celltype_module$celltype == receiver_oi]),
                      rownames(ligand_target_matrix))), ")\n")
```

### Ligand activity: sender-focused AND sender-agnostic (Step 5-7 style)

Two candidate-ligand definitions are run and compared, as recommended by the NicheNet tutorial: **sender-focused** (ligand expressed in a specific `target_celltypes` sender AND its receptor expressed in the receiver -- iterated per sender cell type, as before) and **sender-agnostic** (any ligand in `lr_network` whose receptor is expressed in the receiver, regardless of which cell type expresses the ligand). Ligands that rank highly in *both* are the strongest candidates; ligands unique to the sender-agnostic list may come from cell types outside `target_celltypes` (i.e. cell types WGCNA/MultiNicheNet excluded) and should be interpreted with that caveat.

```{r}
run_candidate_plots <- function(module_oi, receiver_oi, verbose = TRUE) {
  
  geneset_oi <- hub_genes_table %>%
    dplyr::filter(module == module_oi, celltype == receiver_oi) %>%
    dplyr::pull(gene) %>%
    unique()
  geneset_oi <- geneset_oi[geneset_oi %in% rownames(ligand_target_matrix)]
  
  if (verbose) cat("\n===== Módulo:", module_oi, "| Receptor:", receiver_oi,
                   "| Hub genes:", length(geneset_oi), "=====\n")
  
  if (length(geneset_oi) < MIN_GENESET_SIZE) {
    warning("[", module_oi, "/", receiver_oi, "] geneset_oi tiene solo ", length(geneset_oi),
            " genes (< ", MIN_GENESET_SIZE, "). Las estimaciones de ligand activity (AUPR) ",
            "serán inestables.")
  }
  if (length(geneset_oi) < 5) {
    cat("  -> Se salta", module_oi, "/", receiver_oi, ": geneset_oi demasiado pequeño (<5 genes).\n")
    return(NULL)
  }
  
  result <- tryCatch({
    
    background_expressed_genes <- expressed_genes_by_celltype[[receiver_oi]] %>%
      intersect(rownames(ligand_target_matrix))
    
    sender_celltypes    <- base::setdiff(target_celltypes, receiver_oi)
    expressed_receptors <- base::intersect(lr_network$receptor, background_expressed_genes)
    
    # --- Sender-focused: una corrida de ligand activity POR sender celltype ---
    ligand_activities_by_sender <- purrr::map_dfr(sender_celltypes, function(sender_ct) {
      sender_expressed_genes <- expressed_genes_by_celltype[[sender_ct]]
      expressed_ligands <- base::intersect(lr_network$ligand, sender_expressed_genes)
      
      potential_ligands <- lr_network %>%
        dplyr::filter(ligand %in% expressed_ligands, receptor %in% expressed_receptors) %>%
        dplyr::pull(ligand) %>%
        unique()
      
      if (length(potential_ligands) < 1) return(NULL)
      
      act <- tryCatch(
        nichenetr::predict_ligand_activities(
          geneset                    = geneset_oi,
          background_expressed_genes = background_expressed_genes,
          ligand_target_matrix       = ligand_target_matrix,
          potential_ligands          = potential_ligands
        ),
        error = function(e) {
          message("  [", module_oi, "/", receiver_oi, "] Sender '", sender_ct,
                  "': ligand activity falló (", conditionMessage(e), ")")
          NULL
        }
      )
      if (is.null(act)) return(NULL)
      
      act$sender   <- sender_ct
      act$approach <- "sender-focused"
      act
    })
    
    if (nrow(ligand_activities_by_sender) == 0 || !"aupr_corrected" %in% names(ligand_activities_by_sender)) {
      cat("  -> Ningún sender produjo ligand activities válidas para", module_oi, "/", receiver_oi,
          "-- se salta.\n")
      return(NULL)
    }
    ligand_activities_by_sender <- ligand_activities_by_sender %>%
      dplyr::arrange(desc(aupr_corrected))
    
    # --- Sender-agnostic: UNA corrida, cualquier ligando con receptor expresado en el receptor ---
    potential_ligands_agnostic <- lr_network %>%
      dplyr::filter(ligand %in% expressed_genes_union, receptor %in% expressed_receptors) %>%
      dplyr::pull(ligand) %>%
      unique()
    
    ligand_activities_agnostic <- tryCatch(
      nichenetr::predict_ligand_activities(
        geneset                    = geneset_oi,
        background_expressed_genes = background_expressed_genes,
        ligand_target_matrix       = ligand_target_matrix,
        potential_ligands          = potential_ligands_agnostic
      ),
      error = function(e) {
        message("  [", module_oi, "/", receiver_oi, "] Sender-agnostic ligand activity falló (",
                conditionMessage(e), ")")
        NULL
      }
    )
    if (!is.null(ligand_activities_agnostic)) {
      ligand_activities_agnostic <- ligand_activities_agnostic %>%
        dplyr::arrange(desc(aupr_corrected)) %>%
        dplyr::mutate(sender = NA_character_, approach = "sender-agnostic")
    }
    
    write.csv(
      ligand_activities_by_sender,
      paste0("ligand_activity_hubgenes_", module_oi, "_", receiver_oi, "_sender_focused.csv"),
      row.names = FALSE
    )
    if (!is.null(ligand_activities_agnostic)) {
      write.csv(
        ligand_activities_agnostic,
        paste0("ligand_activity_hubgenes_", module_oi, "_", receiver_oi, "_sender_agnostic.csv"),
        row.names = FALSE
      )
    }
    
    top_ligands_focused_set <- ligand_activities_by_sender %>%
      dplyr::slice_max(aupr_corrected, n = TOP_N_LIGANDS, with_ties = FALSE) %>%
      dplyr::pull(test_ligand) %>% unique()
    
    top_ligands_agnostic_set <- if (!is.null(ligand_activities_agnostic)) {
      ligand_activities_agnostic %>%
        dplyr::slice_max(aupr_corrected, n = TOP_N_LIGANDS, with_ties = FALSE) %>%
        dplyr::pull(test_ligand) %>% unique()
    } else character(0)
    
    best_upstream_ligands <- base::union(top_ligands_focused_set, top_ligands_agnostic_set)
    shared_ligands         <- base::intersect(top_ligands_focused_set, top_ligands_agnostic_set)
    
    # --- Plot 1: actividad de ligandos por sender ---
    top_ligands_plot <- ligand_activities_by_sender %>%
      dplyr::group_by(sender) %>%
      dplyr::slice_max(aupr_corrected, n = 5, with_ties = FALSE) %>%
      dplyr::ungroup() %>%
      dplyr::mutate(is_shared = test_ligand %in% shared_ligands)
    
    p_ligand_activity <- ggplot(top_ligands_plot,
                                aes(x = reorder(test_ligand, aupr_corrected), y = aupr_corrected,
                                    fill = sender, alpha = is_shared)) +
      geom_col() +
      coord_flip() +
      facet_wrap(~ sender, scales = "free_y") +
      scale_alpha_manual(values = c(`TRUE` = 1, `FALSE` = 0.55), guide = "none") +
      labs(
        x = "Ligand", y = "Predicted activity (corrected AUPR)",
        title = paste0("Candidate ligands by sender celltype -> hub genes of module ",
                       module_oi, " in ", receiver_oi),
        subtitle = "Target gene set = module hub genes (top kME); solid bars = also top-ranked sender-agnostic"
      ) +
      theme_minimal(base_size = 11) +
      theme(legend.position = "none")
    
    if (verbose) print(p_ligand_activity)
    ggsave(
      paste0("plot_ligand_activity_hubgenes_", module_oi, "_", receiver_oi, ".pdf"),
      p_ligand_activity, width = 10, height = 7
    )
    
    # --- Plot 2: heatmap ligando-target ---
    active_ligand_target_links_df <- best_upstream_ligands %>%
      lapply(get_weighted_ligand_target_links,
             geneset              = geneset_oi,
             ligand_target_matrix = ligand_target_matrix,
             n                    = 100) %>%
      bind_rows() %>%
      tidyr::drop_na()
    
    p_ligand_target_heatmap <- NULL
    if (nrow(active_ligand_target_links_df) > 0) {
      active_ligand_target_links <- prepare_ligand_target_visualization(
        ligand_target_df     = active_ligand_target_links_df,
        ligand_target_matrix = ligand_target_matrix,
        cutoff               = 1/3
      )
      
      order_ligands <- intersect(best_upstream_ligands, colnames(active_ligand_target_links)) %>% rev()
      order_targets <- active_ligand_target_links_df$target %>%
        unique() %>%
        intersect(rownames(active_ligand_target_links))
      
      vis_ligand_target <- active_ligand_target_links[order_targets, order_ligands, drop = FALSE]
      
      p_ligand_target_heatmap <- vis_ligand_target %>%
        make_heatmap_ggplot(
          y_name       = paste0("Hub genes of module ", module_oi, " (", receiver_oi, ")"),
          x_name       = "Prioritized ligands (sender-focused U sender-agnostic)",
          color        = "purple",
          legend_title = "Regulatory\npotential"
        ) +
        theme(axis.text.x = element_text(face = "italic", angle = 90, hjust = 1))
      
      if (verbose) print(p_ligand_target_heatmap)
      ggsave(
        paste0("plot_ligand_target_heatmap_", module_oi, "_", receiver_oi, ".pdf"),
        p_ligand_target_heatmap, width = 10, height = 8
      )
    } else {
      cat("  -> Ningún link ligando-target pasó el cutoff para", module_oi, "/", receiver_oi,
          "-- se salta el heatmap.\n")
    }
    
    # --- Ligando top único + tabla de sus genes target ---
    top_ligand <- dplyr::bind_rows(
      ligand_activities_by_sender %>% dplyr::select(test_ligand, aupr_corrected),
      if (!is.null(ligand_activities_agnostic)) {
        ligand_activities_agnostic %>% dplyr::select(test_ligand, aupr_corrected)
      } else NULL
    ) %>%
      dplyr::arrange(desc(aupr_corrected)) %>%
      dplyr::slice(1) %>%
      dplyr::pull(test_ligand)
    
    active_ligand_targets <- get_weighted_ligand_target_links(
      ligand               = top_ligand,
      geneset              = geneset_oi,
      ligand_target_matrix = ligand_target_matrix,
      n                    = 100
    )
    
    # --- Plot 3: heatmap ligando-receptor ---
    local_lr_network <- lr_network
    if (!"from" %in% colnames(local_lr_network) && "ligand" %in% colnames(local_lr_network)) {
      local_lr_network <- local_lr_network %>% dplyr::rename(from = ligand, to = receptor)
    }
    
    ligand_receptor_links_df <- get_weighted_ligand_receptor_links(
      best_upstream_ligands,
      expressed_receptors,
      local_lr_network,
      weighted_networks$lr_sig
    )
    
    p_ligand_receptor_heatmap <- NULL
    if (nrow(ligand_receptor_links_df) > 0) {
      vis_ligand_receptor_network <- prepare_ligand_receptor_visualization(
        ligand_receptor_links_df,
        best_upstream_ligands,
        order_hclust = "both"
      )
      
      p_ligand_receptor_heatmap <- t(vis_ligand_receptor_network) %>%
        make_heatmap_ggplot(
          y_name       = "Ligands",
          x_name       = paste0("Receptors expressed in receiver (", receiver_oi, ")"),
          color        = "mediumvioletred",
          legend_title = "Prior interaction\npotential"
        ) +
        theme(axis.text.x = element_text(face = "italic", angle = 90, hjust = 1))
      
      if (verbose) print(p_ligand_receptor_heatmap)
      ggsave(
        paste0("plot_ligand_receptor_heatmap_", module_oi, "_", receiver_oi, ".pdf"),
        p_ligand_receptor_heatmap, width = 9, height = 7
      )
      
      write.csv(
        ligand_receptor_links_df,
        paste0("ligand_receptor_links_", module_oi, "_", receiver_oi, ".csv"),
        row.names = FALSE
      )
    } else {
      cat("  -> No se encontraron links ligando-receptor para", module_oi, "/", receiver_oi, ".\n")
    }
    
    # --- Plot 4: signaling path (ligando -> receptor -> TF -> target hub gene) ---
    p_signaling_path  <- NULL
    sig_path_combined <- NULL
    if (nrow(active_ligand_targets) > 0) {
      top_target_for_path <- active_ligand_targets %>%
        dplyr::arrange(desc(weight)) %>%
        dplyr::slice(1) %>%
        dplyr::pull(target)
      
      sig_path <- tryCatch(
        get_ligand_signaling_path(
          ligand_tf_matrix  = ligand_tf_matrix,
          ligands_all       = top_ligand,
          targets_all       = top_target_for_path,
          top_n_regulators  = 4,
          weighted_networks = weighted_networks,
          minmax_scaling    = TRUE
        ),
        error = function(e) {
          message("  [", module_oi, "/", receiver_oi, "] Signaling path inference falló (",
                  conditionMessage(e), ")")
          NULL
        }
      )
      
      if (!is.null(sig_path) && nrow(sig_path$sig) + nrow(sig_path$gr) > 0) {
        sig_path_combined <- dplyr::bind_rows(sig_path$sig, sig_path$gr)
        sig_graph <- igraph::graph_from_data_frame(sig_path_combined, directed = TRUE)
        
        p_signaling_path <- ggraph::ggraph(sig_graph, layout = "kk") +
          ggraph::geom_edge_link(
            aes(edge_width = weight, edge_alpha = weight),
            arrow   = grid::arrow(length = unit(3, "mm"), type = "closed"),
            end_cap = ggraph::circle(3, "mm"),
            color   = "grey50"
          ) +
          ggraph::geom_node_point(size = 5, color = "steelblue") +
          ggraph::geom_node_text(aes(label = name), repel = TRUE, size = 3) +
          ggraph::scale_edge_width(range = c(0.5, 2)) +
          theme_void() +
          labs(title = paste("Signaling path:", top_ligand, "->", top_target_for_path),
               subtitle = paste0("Module ", module_oi, " hub gene, receiver = ", receiver_oi))
        
        if (verbose) print(p_signaling_path)
        ggsave(
          paste0("plot_signaling_path_", module_oi, "_", receiver_oi, "_",
                 top_ligand, "_", top_target_for_path, ".pdf"),
          p_signaling_path, width = 8, height = 6
        )
      } else {
        cat("  -> No se pudo inferir signaling path para", module_oi, "/", receiver_oi,
            "(", top_ligand, "->", top_target_for_path, ").\n")
      }
    } else {
      cat("  -> active_ligand_targets vacío para", module_oi, "/", receiver_oi,
          "-- no hay target para trazar signaling path.\n")
    }
    
    list(
      module_oi                   = module_oi,
      receiver_oi                 = receiver_oi,
      geneset_oi                  = geneset_oi,
      background_expressed_genes  = background_expressed_genes,
      ligand_activities_by_sender = ligand_activities_by_sender,
      ligand_activities_agnostic  = ligand_activities_agnostic,
      best_upstream_ligands       = best_upstream_ligands,
      shared_ligands              = shared_ligands,
      top_ligand                  = top_ligand,
      active_ligand_targets       = active_ligand_targets,
      sig_path_combined           = sig_path_combined,
      p_ligand_activity           = p_ligand_activity,
      p_ligand_target_heatmap     = p_ligand_target_heatmap,
      p_ligand_receptor_heatmap   = p_ligand_receptor_heatmap,
      p_signaling_path            = p_signaling_path
    )
    
  }, error = function(e) {
    message("  [", module_oi, "/", receiver_oi,
            "] run_candidate_plots fallo con un error no controlado: ", conditionMessage(e))
    NULL
  })
  
  return(result)
}
```

### Candidato principal (`top_hit`)

Se corre primero para el candidato principal seleccionado por `SELECTION_METRIC`; sus resultados se asignan a las variables globales `ligand_activities_by_sender` / `ligand_activities_agnostic` que usan las secciones de más abajo (golden table, heatmap general, export multi-hoja), igual que en la versión anterior.

```{r hub-genes-top-hit-run, fig.width=10, fig.height=7}
TOP_N_LIGANDS<-10
top_hit_result <- run_candidate_plots(module_oi, receiver_oi)

if (is.null(top_hit_result)) {
  stop("El candidato top_hit (", module_oi, "/", receiver_oi,
       ") no produjo resultados válidos -- revisa geneset_oi/background.")
}

geneset_oi                  <- top_hit_result$geneset_oi
background_expressed_genes  <- top_hit_result$background_expressed_genes
ligand_activities_by_sender <- top_hit_result$ligand_activities_by_sender
ligand_activities_agnostic  <- top_hit_result$ligand_activities_agnostic
best_upstream_ligands       <- top_hit_result$best_upstream_ligands
shared_ligands               <- top_hit_result$shared_ligands
top_ligand                  <- top_hit_result$top_ligand
active_ligand_targets       <- top_hit_result$active_ligand_targets

cat("\nCandidatos ligandos top (sender-focused, cualquier sender celltype) para el programa hub del módulo",
    module_oi, "en", receiver_oi, ":\n")
print(head(ligand_activities_by_sender, 15))

if (!is.null(ligand_activities_agnostic)) {
  cat("\nCandidatos ligandos top (sender-agnostic) para el mismo programa hub:\n")
  print(head(ligand_activities_agnostic, 15))
}

cat("\nLigandos compartidos por AMBOS enfoques (candidatos más fuertes):\n")
print(shared_ligands)
```

### Todos los candidatos: mismos gráficos, en loop

Se repite exactamente el mismo bloque de análisis y gráficos de arriba (actividad de ligandos por sender, heatmap ligando-target, heatmap ligando-receptor, signaling path) para **cada fila** de `module_celltype_candidates` -- no solo para `top_hit`. Cada candidato genera su propio set de PDFs/CSVs (ya nombrados por `module_oi`/`receiver_oi`, así que no se pisan entre sí). Los resultados de todos los candidatos quedan además en `all_candidates_results` (lista con un elemento por module x celltype) para inspección posterior en R sin tener que releer los CSVs.

```{r hub-genes-all-candidates-loop, fig.width=10, fig.height=7}
cat("Candidatos a procesar en el loop:", nrow(module_celltype_candidates), "\n")

all_candidates_results <- purrr::map2(
  module_celltype_candidates$module,
  module_celltype_candidates$celltype,
  run_candidate_plots
)
names(all_candidates_results) <- paste(
  module_celltype_candidates$module, module_celltype_candidates$celltype, sep = "_"
)

# Candidatos que se saltaron (geneset_oi < 5 genes o ligand activity fallida en todos los senders)
skipped_candidates <- names(all_candidates_results)[vapply(all_candidates_results, is.null, logical(1))]
if (length(skipped_candidates) > 0) {
  cat("\nCandidatos sin resultados (geneset_oi < 5 genes o sin ligand activities válidas):\n")
  print(skipped_candidates)
}

# Tabla resumen: mejor ligando y su AUPR corregido por cada candidato procesado
all_candidates_results_ok <- Filter(Negate(is.null), all_candidates_results)

candidates_summary <- purrr::map_dfr(all_candidates_results_ok, function(res) {
  tibble::tibble(
    module            = res$module_oi,
    celltype          = res$receiver_oi,
    n_hub_genes       = length(res$geneset_oi),
    n_best_upstream   = length(res$best_upstream_ligands),
    n_shared_ligands  = length(res$shared_ligands),
    top_ligand        = res$top_ligand,
    top_ligand_aupr   = max(
      dplyr::bind_rows(
        res$ligand_activities_by_sender %>% dplyr::select(test_ligand, aupr_corrected),
        if (!is.null(res$ligand_activities_agnostic)) {
          res$ligand_activities_agnostic %>% dplyr::select(test_ligand, aupr_corrected)
        } else NULL
      )$aupr_corrected,
      na.rm = TRUE
    )
  )
})

write.csv(candidates_summary, "all_candidates_summary.csv", row.names = FALSE)

cat("\nResumen de todos los candidatos procesados (guardado en all_candidates_summary.csv):\n")
print(candidates_summary)
```

## Summary & Integration: Golden Table, Heatmaps, and Multi-sheet Export

### 1. Build the consolidated "golden table"

The golden table is a single denormalized data frame where every row is a unique ligand-receptor pair, enriched with: the contrasts where it was prioritized, module membership, SASP annotation, and top predicted ligand activity across all sender-celltype combinations. This is the table to use for filtering, sorting, and publication-quality figures.

```{r golden-table-build}
# Start with all prioritized LR pairs
golden_table <- multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  distinct(sender, receiver, ligand, receptor)

# Which contrasts contain each LR pair (pivot to comma-separated string)
lr_per_contrast <- multinichenet_output$prioritization_tables$group_prioritization_tbl %>%
  distinct(contrast, sender, receiver, ligand, receptor) %>%
  group_by(sender, receiver, ligand, receptor) %>%
  summarise(contrasts_active = paste(sort(unique(contrast)), collapse = "; "),
            n_contrasts = n_distinct(contrast), .groups = "drop")

golden_table <- golden_table %>%
  left_join(lr_per_contrast, by = c("sender", "receiver", "ligand", "receptor"))

# WGCNA module membership: ligand in sender cell type?
# ============================================================
# {r golden-table-build}  -- ligand_modules / receptor_modules
# ============================================================
ligand_modules <- gene_celltype_module %>%
  filter(module %in% age_associated_modules) %>%   # antes: significant_modules
  group_by(gene, celltype) %>%
  summarise(module = paste(sort(unique(module)), collapse = "; "), .groups = "drop") %>%
  rename(ligand = gene, sender = celltype, ligand_modules = module)

receptor_modules <- gene_celltype_module %>%
  filter(module %in% age_associated_modules) %>%   # antes: significant_modules
  group_by(gene, celltype) %>%
  summarise(module = paste(sort(unique(module)), collapse = "; "), .groups = "drop") %>%
  rename(receptor = gene, receiver = celltype, receptor_modules = module)



# Top ligand activity for this LR pair (across all senders/receivers)
if (nrow(ligand_activities_by_sender) > 0) {
  top_ligand_activity <- ligand_activities_by_sender %>%
    group_by(test_ligand) %>%
    slice_max(aupr_corrected, n = 1, with_ties = FALSE) %>%
    rename(ligand = test_ligand, max_aupr_corrected = aupr_corrected, aupr_sender_ct = sender) %>%
    dplyr::select(ligand, max_aupr_corrected, aupr_sender_ct)

  golden_table <- golden_table %>%
    left_join(top_ligand_activity, by = c("ligand"))
}

# Sort by activity (if available) or by n_contrasts
golden_table <- golden_table %>%
  arrange(desc(max_aupr_corrected %||% 0), desc(n_contrasts))

cat("Golden table built:", nrow(golden_table), "unique LR pairs\n")
print(head(golden_table, 10))

write.csv(golden_table, "golden_table_LR_pairs.csv", row.names = FALSE)
```

### 2. Ligand activity heatmap (all ligands × sender cell types)

A heatmap showing which ligands are active (high aupr_corrected) in which sender cell types. Restricted to top 30 ligands for readability.

```{r ligand-activity-heatmap, fig.width=12, fig.height=8}
heatmap_data <- ligand_activities_by_sender %>%
  dplyr::select(test_ligand, sender, aupr_corrected) %>%
  group_by(test_ligand, sender) %>%
  slice_max(aupr_corrected, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(desc(aupr_corrected)) %>%
  slice_head(n = 30)

p_ligand_heatmap <- ggplot(heatmap_data, aes(x = sender, y = reorder(test_ligand, aupr_corrected),
                                              fill = aupr_corrected)) +
  geom_tile(color = "white", size = 0.5) +
  scale_fill_gradient(low = "#fee5d9", high = "#a50f15") +
  labs(
    x = "Sender cell type",
    y = "Ligand",
    fill = "AUPR (corrected)",
    title = "Ligand Activity Heatmap",
    subtitle = "Top 30 ligands by max activity; color = predicted activity for receptor target genes"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_ligand_heatmap)
ggsave("plot_ligand_activity_heatmap_top30.pdf", p_ligand_heatmap, width = 12, height = 8)
```

### 3. Network plot of cell-cell communication (interactive visNetwork)

An interactive network where nodes are cell types and edges represent LR pairs. Edge width = number of distinct ligand-receptor pairs; edge color reflects the proportion of pairs overlapping age-associated WGCNA modules.

```{r network-visualization, fig.width=9, fig.height=7}
library(tidyverse)
library(visNetwork)

network_summary <- golden_table %>%
  group_by(sender, receiver) %>%
  summarise(
    n_pairs       = n(),
    n_wgcna       = sum(!is.na(ligand_modules) | !is.na(receptor_modules)),
    .groups = "drop"
  ) %>%
  mutate(
    # Evitamos división por cero o NaNs asegurando que n_pairs > 0
    pct_wgcna = if_else(n_pairs > 0, 100 * n_wgcna / n_pairs, 0),
    # Reemplazamos cualquier NA residual por 0
    pct_wgcna = replace_na(pct_wgcna, 0)
  )

network_nodes <- unique(c(network_summary$sender, network_summary$receiver))

nodes_df <- data.frame(
  id    = network_nodes,
  label = network_nodes,
  title = network_nodes,
  color = "#2c7fb8",
  size  = 30
)

# Creamos el gradiente de forma segura controlando el rango de colores
edge_colors <- colorRamp(c("lightgrey", "#2ca25f"))(pmin(pmax(network_summary$pct_wgcna / 100, 0), 1))

# Manejo seguro por si quedara algún elemento nulo (asegurando dimensiones de matriz)
if (is.matrix(edge_colors)) {
  edge_colors[is.na(edge_colors)] <- 200 
  edge_colors_hex <- apply(edge_colors, 1, function(x) rgb(x[1], x[2], x[3], maxColorValue = 255))
} else {
  edge_colors_hex <- rep("#d3d3d3", nrow(network_summary))
}

# Construcción de los bordes (edges_df) requeridos por visNetwork a partir del resumen
edges_df <- data.frame(
  from  = network_summary$sender,
  to    = network_summary$receiver,
  value = network_summary$n_pairs,
  color = edge_colors_hex,
  title = paste0("Pairs: ", network_summary$n_pairs, "<br>WGCNA %: ", round(network_summary$pct_wgcna, 1), "%")
)

p_network <- visNetwork(nodes_df, edges_df, directed = TRUE, physics = TRUE) %>%
  visLayout(randomSeed = 42) %>%
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) %>%
  visPhysics(enabled = TRUE, stabilization = list(iterations = 200))

visSave(p_network, "network_LR_celltype_interactive.html")
cat("Interactive network saved to: network_LR_celltype_interactive.html\n")
```

### 4. Session info for reproducibility

```{r session-info}
cat("\n=== R Session Info ===\n")
session_info <- capture.output(sessionInfo())
writeLines(session_info, "session_info.txt")
print(sessionInfo())
```

### 5. Export all key tables to multi-sheet Excel workbook

```{r excel-export}
# Clean up list columns for Excel compatibility
golden_table_export <- golden_table %>% 
  mutate(across(where(is.list), ~ as.character(.)))

# Build workbook with multiple sheets
wb <- createWorkbook()

sheet_data <- list(
  "Golden_Table"              = golden_table_export,
  "LR_WGCNA_Overlap"          = lr_prioritized_annotated, 
  "GO_Overlap_Terms"          = go_overlap_combined %>%
    dplyr::select(any_of(c("Description", "module", "p.adjust", "n_overlap_genes", "go_source"))),
  "Ligand_Activity"           = ligand_activities_by_sender %>%
    dplyr::select(any_of(c("test_ligand", "sender", "aupr", "aupr_corrected", "p_value", "p.value"))),
  "Ligand_Activity_Agnostic"  = if (!is.null(ligand_activities_agnostic)) {
    ligand_activities_agnostic %>% dplyr::select(any_of(c("test_ligand", "aupr", "aupr_corrected", "p_value", "p.value")))
  } else data.frame(),
  "Ligand_Receptor_Links"     = ligand_receptor_links_df,
  "Ligand_Activity_Robustness" = robustness_results %>%
    dplyr::select(any_of(c("test_ligand", "sender", "receiver", "module", "aupr_corrected")))
)

# Add sheets with formatting
for (sheet_name in names(sheet_data)) {
  addWorksheet(wb, sheet_name)

  if (is.null(sheet_data[[sheet_name]]) || ncol(sheet_data[[sheet_name]]) == 0) {
    writeData(wb, sheet = sheet_name, x = "No data available for this section")
    next
  }

  writeData(wb, sheet = sheet_name, x = sheet_data[[sheet_name]])

  # Style headers
  headerStyle <- createStyle(
    fgFill = "#2c7fb8", fontColour = "white", textDecoration = "bold", halign = "center"
  )
  addStyle(wb, sheet = sheet_name, style = headerStyle, rows = 1,
           cols = 1:ncol(sheet_data[[sheet_name]]))
  setColWidths(wb, sheet = sheet_name, cols = 1:ncol(sheet_data[[sheet_name]]), widths = "auto")
  freezePane(wb, sheet = sheet_name, firstRow = TRUE)
}

saveWorkbook(wb, "multinichenet_wgcna_integration_tables.xlsx", overwrite = TRUE)
cat("Multi-sheet Excel export saved to: multinichenet_wgcna_integration_tables.xlsx\n")
```
