---
title: "WGCNA_24"
output: html_document
date: "2026-07-24"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(reticulate)
library(Seurat)
library(Matrix)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(patchwork)
library(knitr)
library(hdWGCNA)
library(WGCNA)
library(harmony)
library(pheatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(enrichplot)
library(GOSemSim)
library(ReactomePA)
library(viridis)
library(jsonlite)    
library(scales)       
library(ggpubr)        
library(ggtext)         
library(RColorBrewer)
library(ggalluvial)   
library(ggrepel)      
py_require(c("anndata"))
ad <- import("anndata")
```

```{r}
dir.create("plots_pdf", showWarnings = FALSE)

theme_pub <- function(base_size = 13) {
  theme_bw(base_size = base_size) +
    theme(
      panel.grid.minor  = element_blank(),
      panel.grid.major  = element_line(linewidth = 0.25, color = "grey85"),
      plot.title        = element_text(face = "bold", size = base_size + 1),
      plot.subtitle     = element_text(size = base_size - 1, color = "grey30"),
      axis.title        = element_text(face = "bold"),
      legend.title      = element_text(face = "bold", size = base_size - 1),
      legend.text       = element_text(size = base_size - 2),
      strip.background  = element_rect(fill = "grey90", color = NA),
      strip.text        = element_text(face = "bold")
    )
}


save_pub_pdf <- function(plot, filename, width = 10, height = 8) {
  path <- file.path("plots_pdf", filename)
  ggsave(path, plot, width = width, height = height, device = cairo_pdf, limitsize = FALSE)
  cat("Saved:", path, "\n")
}

```

## Retrieving the data

Downloaded from Figshare as two archives: heart_facs and heart_droplet.

```{r, include=FALSE}
heart_facs    <- ad$read_h5ad("C:/Users/olaia/Desktop/WGCNA_mouse/data/Heart_facs.h5ad")
heart_droplet <-ad$read_h5ad("C:/Users/olaia/Desktop/WGCNA_mouse/data/Heart_droplet.h5ad")
```

## Combining datasets, normalization and data preparation

```{r}
# var_names is already the gene symbol
facs_genes    <- rownames(py_to_r(heart_facs$var))
droplet_genes <- rownames(py_to_r(heart_droplet$var))
shared_genes  <- intersect(facs_genes, droplet_genes)
cat("Shared genes:", length(shared_genes), "\n")

combined_heart <- ad$concat(
  list(heart_droplet, heart_facs),
  axis = 0L,
  join = "outer",
  keys = c("droplet", "facs"),
  label = "tech",
  index_unique = "-"
)
combined_genes <- py_to_r(combined_heart$var_names$values)
combined_cells <- py_to_r(combined_heart$obs_names$values)


rm(heart_facs, heart_droplet)
gc()

cat("Combined genes:", length(combined_genes), "\n")
cat("Combined cells:", length(combined_cells), "\n")

stopifnot(length(combined_genes) == ncol(combined_heart$X))
stopifnot(length(combined_cells) == nrow(combined_heart$X))

counts_matrix <- Matrix::t(as(combined_heart$X, "CsparseMatrix"))

sanitized_genes <- gsub("_", "-", combined_genes)
dup_idx <- duplicated(sanitized_genes) | duplicated(sanitized_genes, fromLast = TRUE)
if (sum(dup_idx) > 0) {
  cat("Genes affected by sanitization collisions:", sum(dup_idx), "\n")
  print(data.frame(original = combined_genes[dup_idx],
                   sanitized = sanitized_genes[dup_idx]))
}
final_genes <- make.unique(sanitized_genes)
stopifnot(sum(duplicated(final_genes)) == 0)
stopifnot(length(final_genes) == nrow(counts_matrix))

final_cells <- as.vector(combined_cells)
stopifnot(is.null(attributes(final_cells)))

rownames(counts_matrix) <- final_genes
colnames(counts_matrix) <- final_cells

metadata <- py_to_r(combined_heart$obs)
metadata <- as.data.frame(metadata, stringsAsFactors = FALSE)
rownames(metadata) <- final_cells

# 'tech' should have exactly 2 levels (facs/droplet)
cat("'tech' values (real technology):\n")
print(table(metadata$tech))


if ("batch" %in% colnames(metadata)) {
  metadata$seq_batch <- metadata$batch
  metadata$batch <- NULL
}


if ("mouse.id" %in% colnames(metadata)) {
  metadata$sample_id <- as.character(metadata$mouse.id)
} else {
  metadata$sample_id <- paste(metadata$sex, metadata$age, sep = "_")
}

cat("Numero de sample_id unicos:", dplyr::n_distinct(metadata$sample_id), "\n")
table(metadata$sample_id)

cat("Do counts_matrix and metadata match?",
    identical(rownames(metadata), colnames(counts_matrix)), "\n")
stopifnot(identical(rownames(metadata), colnames(counts_matrix)))

seurat_combined <- CreateSeuratObject(
  counts = counts_matrix,
  meta.data = metadata,
  project = "WGCNA_Mouse_Combined"
)


rm(counts_matrix, combined_heart, combined_genes, combined_cells,
   sanitized_genes, final_genes, final_cells, metadata)
gc()
```

```{r}
# QC thresholds 
min_features <- 200
n_before <- ncol(seurat_combined)

seurat_combined <- subset(
  seurat_combined,
  subset = nFeature_RNA > min_features 
)
cat("Cells before QC:", n_before, "\n")
cat("Cells after QC:", ncol(seurat_combined), "\n")

# Remove genes detected in very few cells globally
min_cells_gene <- 3
detected_genes <- Matrix::rowSums(GetAssayData(seurat_combined, layer = "counts") > 0)
genes_to_keep <- names(detected_genes[detected_genes >= min_cells_gene])

cat("Genes before filter:", nrow(seurat_combined), "\n")
cat("Genes after filter:", length(genes_to_keep), "\n")

seurat_combined <- subset(seurat_combined, features = genes_to_keep)
seurat_combined <- NormalizeData(seurat_combined)
seurat_combined <- FindVariableFeatures(seurat_combined)
seurat_combined <- ScaleData(seurat_combined)
seurat_combined <- RunPCA(seurat_combined)

seurat_combined <- RunHarmony(
  seurat_combined,
  group.by.vars = "tech",
  reduction.use = "pca",
  reduction.save = "harmony"
)

seurat_combined <- FindNeighbors(seurat_combined, reduction = "harmony", dims = 1:20)
seurat_combined <- RunUMAP(seurat_combined, reduction = "harmony", dims = 1:15)
seurat_combined_filt <- subset(seurat_combined, subset = cell_ontology_class != "nan")


rm(seurat_combined, detected_genes, genes_to_keep)
gc()
```

## Collapse cell_ontology_class into final cell type categories

```{r inspect-cell-ontology}

cat("Raw cell_ontology_class levels present in seurat_combined_filt:\n")
print(sort(unique(as.character(seurat_combined_filt$cell_ontology_class))))
```

```{r celltype-mapping}
# Collapse into 8 final categories. Cell types not listed here (e.g. mast
# cell, valve cell) are intentionally dropped from the analysis.

celltype_map <- c(
  "endothelial cell of coronary artery" = "EndothelialCells",
  "aortic endothelial cell"             = "EndothelialCells",
  "atrial myocyte"                      = "Myocytes",
  "ventricular myocyte"                 = "Myocytes",
  "cardiac neuron"                      = "CardiacNeurons",
  qaaaaaaa,
  "erythrocyte"                         = "Erythrocytes",
  "fibroblast of cardiac tissue"        = "Fibroblasts",
  "leukocyte"                           = "Leukocytes",
  "smooth muscle cell"                  = "SmoothMuscleCells"
  
)


meta <- seurat_combined_filt@meta.data
meta$cell_ontology_class_clean <- trimws(tolower(as.character(meta$cell_ontology_class)))


names(celltype_map) <- tolower(names(celltype_map))
meta$final_celltype <- celltype_map[meta$cell_ontology_class_clean]

unmapped <- sort(unique(meta$cell_ontology_class[is.na(meta$final_celltype)]))
cat("Unmapped / excluded cell_ontology_class levels (dropped from analysis):\n")
print(unmapped)

cat("\nCells per final cell type:\n")
print(table(meta$final_celltype, useNA = "ifany"))

```

## Restrict to target cell types (before imputation)

```{r restrict-celltypes}
target_celltypes <- c("Fibroblasts", "EndothelialCells", "Myocytes", "Leukocytes", "SmoothMuscleCells", "EndocardialCells" )

cat("Celulas por tipo celular final ANTES de restringir a los tipos objetivo:\n")
print(table(meta$final_celltype, useNA = "ifany"))

meta$final_celltype[!(meta$final_celltype %in% target_celltypes)] <- NA

cat("\nCelulas por tipo celular final DESPUES de restringir a (",
    paste(target_celltypes, collapse = ", "), "):\n", sep = "")
print(table(meta$final_celltype, useNA = "ifany"))
```

```{r celltype-by-age-breakdown}

meta_mapped <- meta[!is.na(meta$final_celltype), ]

age_breakdown <- table(meta_mapped$final_celltype, meta_mapped$age)
cat("Cell counts per final cell type x age:\n")
print(age_breakdown)

age_breakdown_df <- as.data.frame.matrix(age_breakdown)
age_breakdown_df <- cbind(CellType = rownames(age_breakdown_df), age_breakdown_df)
rownames(age_breakdown_df) <- NULL
print(age_breakdown_df)

```

## Build the pseudobulk matrix: mean expression (linear space) per gene x cell type x mouse

```{r build-pseudobulk-matrix}
DefaultAssay(seurat_combined_filt) <- "RNA"

all_genes<-rownames(GetAssayData(seurat_combined_filt, layer = "data"))

expr <- GetAssayData(seurat_combined_filt, layer = "data")[all_genes, ]

keep_cells <- rownames(meta)[!is.na(meta$final_celltype)]
meta_keep  <- meta[keep_cells, ]
expr_keep  <- expr[, keep_cells]

cat("Cells kept for pseudobulk:", ncol(expr_keep),
    "out of", ncol(expr), "\n")

final_celltypes <- sort(unique(meta_keep$final_celltype))
all_samples     <- sort(unique(meta_keep$sample_id))

cat("\nFinal cell types included:\n")
print(final_celltypes)
cat("\nNumber of mice (sample_id):", length(all_samples), "\n")

n_genes <- nrow(expr_keep)
block_list <- vector("list", length(final_celltypes))
names(block_list) <- final_celltypes

for (ct in final_celltypes) {
  ct_mat <- matrix(
    NA_real_, nrow = n_genes, ncol = length(all_samples),
    dimnames = list(rownames(expr_keep), all_samples)
  )
  for (s in all_samples) {
    cells_gs <- rownames(meta_keep)[meta_keep$sample_id == s & meta_keep$final_celltype == ct]
    if (length(cells_gs) > 0) {
      
      block_vals <- as.matrix(expr_keep[, cells_gs, drop = FALSE])
      ct_mat[, s] <- log1p(matrixStats::rowMeans2(expm1(block_vals)))
    }
    # length(cells_gs) == 0 -> stays NA_real_ (no data for this mouse/celltype)
  }
  rownames(ct_mat) <- paste(rownames(ct_mat), ct, sep = "_")
  block_list[[ct]] <- ct_mat
}

pseudobulk_matrix <- do.call(rbind, block_list)
cat("Pseudobulk matrix dimensions (gene_celltype x mouse):",
    nrow(pseudobulk_matrix), "x", ncol(pseudobulk_matrix), "\n")


n_dup <- sum(duplicated(rownames(pseudobulk_matrix)))
cat("Duplicated gene_celltype row names (should be 0):", n_dup, "\n")
stopifnot(n_dup == 0)

cat("\nCells contributing to each (mouse x celltype) combination:\n")
cell_count_tab <- table(meta_keep$sample_id, meta_keep$final_celltype)
print(cell_count_tab)
cat("\nNumber of (mouse x celltype) combinations with 0 cells (-> NA in the matrix):",
    sum(cell_count_tab == 0), "out of", length(cell_count_tab), "\n")

cat("\nOverall fraction of NA in pseudobulk_matrix:",
    round(mean(is.na(pseudobulk_matrix)), 4), "\n")
cat("Overall fraction of TRUE zero-expression values (non-NA, ==0):",
    round(mean(pseudobulk_matrix == 0, na.rm = TRUE), 4), "\n")


rm(expr, expr_keep, block_list, ct_mat)
gc()
```

## Diagnostic: which mice are missing which cell types?

```{r diagnose-missing-celltype-per-mouse}

# cell_count_tab: rows = mice, columns = final_celltype, values = n cells
missing_tab <- cell_count_tab == 0  # TRUE where a mouse has 0 cells of that type

# --- Per cell type: how many / which mice are missing it ---
celltype_missing_summary <- data.frame(
  celltype       = colnames(missing_tab),
  n_mice_missing = colSums(missing_tab),
  pct_mice_missing = round(100 * colSums(missing_tab) / nrow(missing_tab), 1),
  missing_mice   = sapply(colnames(missing_tab), function(ct) {
    paste(rownames(missing_tab)[missing_tab[, ct]], collapse = ", ")
  })
)
celltype_missing_summary <- celltype_missing_summary[order(-celltype_missing_summary$n_mice_missing), ]
rownames(celltype_missing_summary) <- NULL

cat("Cell types ranked by how many mice are missing them (out of",
    nrow(missing_tab), "mice):\n\n")
print(celltype_missing_summary, right = FALSE)

# --- Per mouse: how many / which cell types it's missing ---
mouse_missing_summary <- data.frame(
  mouse            = rownames(missing_tab),
  n_celltypes_missing = rowSums(missing_tab),
  missing_celltypes   = apply(missing_tab, 1, function(r) {
    paste(colnames(missing_tab)[r], collapse = ", ")
  })
)
mouse_missing_summary <- mouse_missing_summary[order(-mouse_missing_summary$n_celltypes_missing), ]
rownames(mouse_missing_summary) <- NULL

cat("\n\nMice ranked by how many cell types they're missing (out of",
    ncol(missing_tab), "cell types):\n\n")
print(mouse_missing_summary, right = FALSE)

# Save both for inspection outside the console
write.csv(celltype_missing_summary, "diagnostic_missing_by_celltype.csv", row.names = FALSE)
write.csv(mouse_missing_summary, "diagnostic_missing_by_mouse.csv", row.names = FALSE)
```

## Drop majority-missing cell types, impute the rest

```{r}

MAX_MISSING_FRAC <- 0.5  # cell types missing in >50% of mice are dropped entirely

celltype_missing_summary$missing_frac <- celltype_missing_summary$pct_mice_missing / 100
celltypes_to_drop <- celltype_missing_summary$celltype[celltype_missing_summary$missing_frac > MAX_MISSING_FRAC]
celltypes_to_keep <- celltype_missing_summary$celltype[celltype_missing_summary$missing_frac <= MAX_MISSING_FRAC]

cat("Cell types DROPPED entirely (>", MAX_MISSING_FRAC * 100, "% of mice missing this type):\n")
print(celltypes_to_drop)
cat("\nCell types KEPT (imputing the few missing mice):\n")
print(celltypes_to_keep)


row_celltype <- sub("^.*_", "", rownames(pseudobulk_matrix))
pseudobulk_matrix_clean <- pseudobulk_matrix[row_celltype %in% celltypes_to_keep, , drop = FALSE]

cat("\nDimensions after dropping majority-missing cell types:",
    dim(pseudobulk_matrix_clean), "(gene_celltype x sample_id)\n")
cat("NA fraction before imputation:",
    round(mean(is.na(pseudobulk_matrix_clean)), 4), "\n")

# Mean-impute remaining NAs (per feature/row, restricted to mice of the SAME age group).
sample_ages <- meta$age[match(colnames(pseudobulk_matrix_clean), meta$sample_id)]
n_na_before <- sum(is.na(pseudobulk_matrix_clean))
global_row_means <- rowMeans(pseudobulk_matrix_clean, na.rm = TRUE)

for (a in unique(sample_ages)) {
  cols_a <- which(sample_ages == a)
  
  # Calculate row means for this specific age group
  age_row_means <- rowMeans(pseudobulk_matrix_clean[, cols_a, drop = FALSE], na.rm = TRUE)
  
  # Fallback to global row mean if all mice in this age group are NA for a feature
  nan_mask <- is.nan(age_row_means)
  if (any(nan_mask)) {
    age_row_means[nan_mask] <- global_row_means[nan_mask]
  }
  
  # Impute NAs within this age group's subset
  sub_mat <- pseudobulk_matrix_clean[, cols_a, drop = FALSE]
  na_sub_idx <- which(is.na(sub_mat), arr.ind = TRUE)
  
  if (nrow(na_sub_idx) > 0) {
    sub_mat[na_sub_idx] <- age_row_means[na_sub_idx[, 1]]
    pseudobulk_matrix_clean[, cols_a] <- sub_mat
  }
}

cat("\nValues imputed:", n_na_before, "\n")
cat("NAs remaining after imputation (should be 0):",
    sum(is.na(pseudobulk_matrix_clean)), "\n")
stopifnot(sum(is.na(pseudobulk_matrix_clean)) == 0)

```

```{r}
pseudobulk_export <- data.frame(
  gene_celltype = rownames(pseudobulk_matrix_clean),
  pseudobulk_matrix_clean,
  check.names = FALSE,   # evita que R modifique los nombres de columna (sample_id)
  row.names = NULL
)


write.csv(
  pseudobulk_export,
  file = "pseudobulk_matrix_gene_celltype_x_mouse_clean.csv",
  row.names = FALSE
)
```

```{r filter-complete-case}

cat("Pseudobulk matrix (post drop+impute) before this sanity check:\n")
cat("  Dimensions:", dim(pseudobulk_matrix_clean), "(gene_celltype × sample_id)\n")
cat("  NA count:", sum(is.na(pseudobulk_matrix_clean)), "\n")
cat("  NA fraction:", round(mean(is.na(pseudobulk_matrix_clean)), 4), "\n")


features_complete <- rownames(pseudobulk_matrix_clean)[rowSums(is.na(pseudobulk_matrix_clean)) == 0]
cat("\nFeatures with COMPLETE data (no NAs):", length(features_complete), "\n")
cat("Features eliminated (with ANY NA):", nrow(pseudobulk_matrix_clean) - length(features_complete), "\n")

# Keep only complete features (subset ROWS, not columns)
pseudobulk_complete <- pseudobulk_matrix_clean[features_complete, , drop = FALSE]

cat("\nPseudobulk matrix after complete-case filtering:\n")
cat("  Dimensions:", dim(pseudobulk_complete), "(gene_celltype x sample_id)\n")
cat("  NA fraction:", round(mean(is.na(pseudobulk_complete)), 4), "(should be 0)\n")

# Diagnostic: which cell types were hit hardest?
features_celltype <- sub("^.*_", "", rownames(pseudobulk_matrix_clean))
features_complete_celltype <- sub("^.*_", "", features_complete)

cat("\nFeature retention by cell type:\n")
retention_by_type <- data.frame()
for (ct in sort(unique(features_celltype))) {
  total <- sum(features_celltype == ct)
  retained <- sum(features_complete_celltype == ct)
  pct <- round(100 * retained / total, 1)
  retention_by_type <- rbind(retention_by_type, data.frame(
    celltype = ct, total = total, retained = retained, pct = pct
  ))
}
print(retention_by_type)

write.csv(retention_by_type, "feature_retention_by_celltype.csv", row.names = FALSE)


rm(pseudobulk_matrix_clean)
gc()
```

## Remove genes with zero expression across all samples

Instead of excluding genes belonging to a previously computed "grey"
module (from a separate global kME table), features are now filtered
directly from the pseudobulk matrix itself: any (gene x celltype)
feature whose expression is exactly 0 in **every** mouse is removed,
since it carries no signal and no variability to contribute to network
construction.

```{r remove-zero-expression-features}
n_before_zero_filter <- nrow(pseudobulk_complete)

all_zero_feature <- rowSums(pseudobulk_complete != 0) == 0
cat("Features with expression == 0 across ALL samples (to REMOVE):",
    sum(all_zero_feature), "\n")

pseudobulk_complete <- pseudobulk_complete[!all_zero_feature, , drop = FALSE]

cat("Features before zero-expression filter:", n_before_zero_filter, "\n")
cat("Features after zero-expression filter:", nrow(pseudobulk_complete), "\n")


features_complete <- rownames(pseudobulk_complete)
features_complete_celltype <- sub("^.*_", "", features_complete)

cat("\nFeatures remaining per cell type after zero-expression filter:\n")
print(table(features_complete_celltype))

rm(all_zero_feature, n_before_zero_filter)
```

## Build Final Pseudobulk Matrix

```{r build-final-pseudobulk}

pseudobulk_hvg <- pseudobulk_complete

cat("Final pseudobulk matrix (selected features):\n")
cat("  Dimensions:", dim(pseudobulk_hvg), "(gene_celltype x sample_id)\n")
cat("  NA count:", sum(is.na(pseudobulk_hvg)), "(should be 0)\n")
cat("  NA fraction:", round(mean(is.na(pseudobulk_hvg)), 4), "\n")

# Check for duplicates
n_dup <- sum(duplicated(rownames(pseudobulk_hvg)))
cat("  Duplicate row names:", n_dup, "\n")
stopifnot(n_dup == 0)

cat("\nMice (columns) in final matrix:", ncol(pseudobulk_hvg), "\n")


```

```{r enable-wgcna-fast-path}
allowWGCNAThreads()
cat("WGCNA threads enabled:", WGCNA::WGCNAnThreads(), "\n")

datExpr<-t(pseudobulk_hvg)

```

## Soft-thresholding power selection

```{r pick-soft-threshold, fig.width=10, fig.height=5}
# 
# powers <- c(1:10, seq(12, 30, 2))
# sft <- pickSoftThreshold(
#   datExpr,
#   powerVector = powers,
#   networkType = "signed",
#   corFnc = "bicor",
#   corOptions = list(use = "all.obs",
#                      maxPOutliers = 0.1,
#                      pearsonFallback = "individual"),
#   verbose = 5
# )
# 
# par(mfrow = c(1, 2))
# plot(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
#      xlab = "Soft Threshold (power)",
#      ylab = "Scale Free Topology Model Fit, signed R^2",
#      main = "Scale independence", type = "n")
# text(sft$fitIndices[, 1], -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
#      labels = powers, cex = 0.7, col = "red")
# abline(h = 0.90, col = "red", lty = 2)
# mtext("Threshold p = 0.90", side = 4, at = 0.90, las = 1, line = 0.5)
# 
# plot(sft$fitIndices[, 1], sft$fitIndices[, 5],
#      xlab = "Soft Threshold (power)",
#      ylab = "Mean connectivity",
#      main = "Mean connectivity", type = "n")
# text(sft$fitIndices[, 1], sft$fitIndices[, 5],
#      labels = powers, cex = 0.7, col = "red")
# 
# if (!is.na(sft$powerEstimate)) {
#   cat("Recommended soft-threshold power:", sft$powerEstimate, "\n")
# } else {
#   cat("No power could be recommended; inspect the plots above and choose manually.\n")
# }
```

## Network construction (blockwiseModules)

```{r choose-power}
# soft_power <- sft$powerEstimate
# if (is.na(soft_power)) {
#   # fallback if no power reaches R^2 = 0.8; pick manually after inspecting the plot above
#   soft_power <- 12
#   cat("No power reached the 0.8 R^2 threshold automatically; using fallback power =",
#       soft_power, "-- inspect the plot above and adjust if needed.\n")
# } else {
#   cat("Selected soft power:", soft_power, "\n")
# }
# 
# n_feat <- ncol(datExpr)
# cat("Features (genes x cell types) to cluster:", n_feat, "\n")

```

```{r}

RANDOM_SEED <- 12345

BICOR_OPTIONS <- "use = 'all.obs', maxPOutliers = 0.1, pearsonFallback = 'individual'"

```

### hdWGCNA pseudobulk setup: reuse the same imputed `datExpr` inside a minimal Seurat container

```{r setup-hdwgcna-object}

pseudobulk_mat_for_hdwgcna <- pseudobulk_hvg  # gene_celltype (rows) x mice (columns)
stopifnot(!anyNA(pseudobulk_mat_for_hdwgcna))

original_gc_names          <- rownames(pseudobulk_mat_for_hdwgcna)
sanitized_gc_names         <- gsub("_", "-", original_gc_names)
stopifnot(sum(duplicated(sanitized_gc_names)) == 0)  # sanity: no collisions after sanitizing
hdwgcna_name_map            <- setNames(original_gc_names, sanitized_gc_names)  # sanitized -> original

rownames(pseudobulk_mat_for_hdwgcna) <- sanitized_gc_names

hdwgcna_dummy_meta <- data.frame(
  row.names = colnames(pseudobulk_mat_for_hdwgcna),
  sample_id = colnames(pseudobulk_mat_for_hdwgcna)
)


seurat_hd <- CreateSeuratObject(
  counts    = pseudobulk_mat_for_hdwgcna,
  meta.data = hdwgcna_dummy_meta,
  project   = "WGCNA_pseudobulk_hdWGCNA"
)


stopifnot(all(rownames(pseudobulk_mat_for_hdwgcna) %in% rownames(seurat_hd)))

seurat_hd <- SetupForWGCNA(
  seurat_hd,
  features   = rownames(pseudobulk_mat_for_hdwgcna),
  wgcna_name = "pseudobulk"
)

seurat_hd <- SetDatExpr(
  seurat_hd,
  mat        = t(pseudobulk_mat_for_hdwgcna),
  wgcna_name = "pseudobulk"
)


cat("hdWGCNA container ready:", nrow(pseudobulk_mat_for_hdwgcna), "features x",
    ncol(pseudobulk_mat_for_hdwgcna), "mice\n")

dir.create("TOM_pseudobulk", showWarnings = FALSE)
```

### Parameter scan: find the minModuleSize / mergeCutHeight / deepSplit combo that keeps the most information.

```{r parameter-scan}
# 
# grid <- expand.grid(
#   minModuleSize  = c(15, 20, 30),
#   deepSplit      = c(1, 2, 3),
#   mergeCutHeight = c(0.15, 0.20, 0.25),
#   KEEP.OUT.ATTRS = FALSE
# )
# 
# # --- Función de escaneo corregida ---
# score_combo <- function(min_size, deep_split, merge_height, build_tom = FALSE) {
#   tryCatch({
#     res_obj <- ConstructNetwork(
#       seurat_hd,
#       soft_power      = soft_power,
#       corType         = "bicor",
#       corOptions      = BICOR_OPTIONS,
#       networkType     = "signed",
#       TOMType         = "signed",
#       tom_name        = "pseudobulk_scan",
#       tom_outdir      = "TOM_pseudobulk",
#       overwrite_tom   = build_tom,
#       maxBlockSize    = 20500,
#       randomSeed      = RANDOM_SEED,
#       deepSplit       = deep_split,
#       minModuleSize   = min_size,
#       mergeCutHeight  = merge_height,
#       pamStage        = TRUE,
#       detectCutHeight = 0.995,
#       wgcna_name      = "pseudobulk"
#     )
#     
#     # Extraer métricas pasando wgcna_name
#     mods <- GetModules(res_obj, wgcna_name = "pseudobulk")
#     tab <- table(mods$module)
#     
#     n_mods <- length(tab) - 1 # Excluir grey
#     grey_frac <- mean(mods$module == "grey")
#     med_size <- median(tab[names(tab) != "grey"])
#     
#     return(data.frame(
#       minModuleSize = min_size,
#       deepSplit = deep_split,
#       mergeCutHeight = merge_height,
#       n_modules = n_mods,
#       grey_fraction = grey_frac,
#       median_module_size = med_size,
#       stringsAsFactors = FALSE
#     ))
#   }, error = function(e) {
#     message(paste0("Combo failed (minModuleSize=", min_size, 
#                    ", deepSplit=", deep_split, 
#                    ", mergeCutHeight=", merge_height, "): ", e$message))
#     return(NULL)
#   })
# }
# 
# cat("Scanning", nrow(grid), "parameter combinations",
#     "(TOM built once on combo 1, reused/cheap tree-cut on the other",
#     nrow(grid) - 1, ")...\n")
# 
# scan_results <- vector("list", nrow(grid))
# for (i in seq_len(nrow(grid))) {
#   scan_results[[i]] <- score_combo(
#     grid$minModuleSize[i], grid$deepSplit[i], grid$mergeCutHeight[i],
#     build_tom = TRUE
#   )
# }
# scan_results <- bind_rows(scan_results)
# 
# # Score: reward low grey fraction (information kept) and penalize both
# # extremes -- too few big modules (no resolution) and too many tiny ones
# # (noise, hard to interpret/GO-enrich).
# scan_results <- scan_results %>%
#   mutate(
#     score = -grey_fraction
#     - 0.01 * pmax(0, 6 - n_modules)
#     - 0.01 * pmax(0, n_modules - 30)
#     - 0.002 * pmax(0, 15 - median_module_size)
#   ) %>%
#   arrange(desc(score))
# 
# cat("Top 10 parameter combinations (best first):\n")
# print(head(scan_results, 10))
# 
# best_params <- scan_results[1, ]
# cat("\nChosen combo -> minModuleSize:", best_params$minModuleSize,
#     "| deepSplit:", best_params$deepSplit,
#     "| mergeCutHeight:", best_params$mergeCutHeight,
#     "| grey fraction:", round(best_params$grey_fraction, 3),
#     "| n_modules:", best_params$n_modules, "\n")
# 
# write.csv(scan_results, "network_parameter_scan.csv", row.names = FALSE)

```

### Final network, built with the best-scoring parameter combination

```{r blockwise-final-after-scan}
soft_power<-10

seurat_hd <- ConstructNetwork(
  seurat_hd,
  soft_power      = soft_power,
  corType         = "bicor",
  corOptions      = BICOR_OPTIONS,
  networkType     = "signed",
  TOMType         = "signed",
  tom_name        = "pseudobulk_scan",
  tom_outdir      = "TOM_pseudobulk",
  overwrite_tom   = TRUE,
  maxBlockSize    = 110000, 
  randomSeed      = RANDOM_SEED,
  deepSplit       = 2,
  minModuleSize   = 15, 
  mergeCutHeight  = 0.15,
  pamStage        = TRUE,
  detectCutHeight = 0.995,
  wgcna_name      = "pseudobulk"
)

# 1. Extraer dendrograma y módulos
net <- GetNetworkData(seurat_hd, wgcna_name = "pseudobulk")
dendro <- net$dendrograms[[1]]
mods <- GetModules(seurat_hd, wgcna_name = "pseudobulk")



if (!is.null(net$blockGenes)) {
  genes_block1 <- net$blockGenes[[1]]
  colors_aligned <- as.character(mods$color[genes_block1])
} else {
  n_tree <- length(dendro$order)
  colors_aligned <- as.character(mods$color[1:n_tree])
}


cat("Clase del objeto de colores:", class(colors_aligned), "\n") # DEBE decir 'character'
cat("Primeros 5 colores:", head(colors_aligned), "\n")
cat("Longitud del árbol vs colores:", length(dendro$order), "vs", length(colors_aligned), "\n")



pdf("dendogram.pdf", width = 10, height = 6)


par(mar = c(3, 4, 3, 1))


WGCNA::plotDendroAndColors(
  dendro          = dendro,
  colors          = colors_aligned,     
  groupLabels     = "Moduls",
  dendroLabels    = FALSE,
  hang            = 0.03,
  addGuide        = FALSE,
  setLayout       = TRUE,
  autoColorHeight = FALSE,
  colorHeight     = 0.2,               
  main            = "Gene dendrogram and module colors"
)

dev.off()
```

```{r map-modules-back}

gene_celltype_module <- data.frame(
  gene_celltype = rownames(pseudobulk_hvg),
  module = mods$color,
  stringsAsFactors = FALSE
)

gene_celltype_module$celltype <- sub("^.*_", "", gene_celltype_module$gene_celltype)
gene_celltype_module$gene     <- sub("_[^_]+$", "", gene_celltype_module$gene_celltype)

cat("Module composition by cell type:\n")
print(table(gene_celltype_module$module, gene_celltype_module$celltype))

prop.table(table(gene_celltype_module$module == "grey")) * 100

gene_celltype_module<- subset(gene_celltype_module, module != "grey")



```

```{r}

keep_idx <- mods$color != "grey"

pseudobulk_hvg_nogrey <- pseudobulk_hvg[keep_idx, ]
colors_nogrey         <- mods$color[keep_idx]


MEs <- moduleEigengenes(
  t(pseudobulk_hvg_nogrey), 
  colors    = colors_nogrey, 
  softPower = soft_power
)$eigengenes

MEs <- orderMEs(MEs)
rownames(MEs) <- rownames(t(pseudobulk_hvg_nogrey))  # mouse (sample_id)

module_names <- colnames(MEs)

head(MEs)

pseudobulk_hvg<-pseudobulk_hvg_nogrey
```

## Module-age correlation

Unlike the metacell-based approach, `datExpr` is **already one row per
mouse** (the pseudobulk aggregation collapsed all cells of a given mouse
x cell type into a single value). This means there is no
pseudoreplication to correct for here: `n` in every correlation below is
genuinely the number of mice, not the number of cells, so no separate
cell-level/mouse-level aggregation step is needed as it was in the
metacell workflow.

```{r age-per-mouse}

age_per_mouse <- meta_keep %>%
  distinct(sample_id, age) %>%
  group_by(sample_id) %>%
  mutate(n_ages = n()) %>%
  ungroup()

stopifnot("Some mice have more than one 'age' value -- inspect age_per_mouse" =
            all(age_per_mouse$n_ages == 1))

age_lookup <- setNames(as.character(age_per_mouse$age), age_per_mouse$sample_id)

ordered_age_levels <- unique(age_lookup)
ordered_age_levels <- ordered_age_levels[
  order(as.numeric(gsub("[^0-9]", "", ordered_age_levels)))
]
cat("Chronological order of 'age' used throughout the analysis:\n")
print(ordered_age_levels)

mouse_age <- factor(age_lookup[rownames(t(pseudobulk_hvg))], levels = ordered_age_levels, ordered = TRUE)
stopifnot(!any(is.na(mouse_age)))

age_num <- as.numeric(gsub("[^0-9.]", "", as.character(mouse_age)))
stopifnot(!any(is.na(age_num)))

cat("\nMice per age group in datExpr:\n")
print(table(mouse_age))
cat("\nNumeric age values used as 'age_num' (real units, not rank):\n")
print(sort(unique(age_num)))
```

```{r module-trait-correlation}
n_mice <- nrow(MEs)

module_trait_cor <- sapply(MEs, function(me) cor(me, age_num, method = "pearson"))
module_trait_p <- sapply(module_trait_cor, function(r) {
  if (is.na(r) || n_mice <= 3) return(1)
  t_val <- r * sqrt((n_mice - 2) / (1 - r^2))
  2 * pt(-abs(t_val), df = n_mice - 2)
})

module_trait_table <- data.frame(
  module      = sub("^ME", "", names(module_trait_cor)),
  correlation = as.numeric(module_trait_cor),
  p_value     = as.numeric(module_trait_p),
  n_mice      = n_mice
) %>%
  mutate(fdr = p.adjust(p_value, method = "BH")) %>%
  arrange(fdr)

cat("Module-age correlation (n =", n_mice, "mice):\n")
print(module_trait_table)

write.csv(module_trait_table, "module_age_correlation.csv", row.names = FALSE)

significant_modules <- module_trait_table %>% filter(fdr < 0.05) %>% pull(module)
cat("\nModules significantly associated with age (FDR < 0.05):\n")
print(significant_modules)
```

```{r module-trait-heatmap, fig.width=6, fig.height=8}
trait_mat <- matrix(module_trait_cor, ncol = 1,
                     dimnames = list(sub("^ME", "", names(module_trait_cor)), "age"))
p_mat <- matrix(module_trait_p, ncol = 1)
text_mat <- paste0(signif(trait_mat, 2), "\n(", signif(p_mat, 1), ")")
dim(text_mat) <- dim(trait_mat)

labeledHeatmap(
  Matrix = trait_mat,
  xLabels = "age",
  yLabels = rownames(trait_mat),
  ySymbols = rownames(trait_mat),
  colorLabels = FALSE,
  colors = blueWhiteRed(50),
  textMatrix = text_mat,
  setStdMargins = FALSE,
  cex.text = 0.6,
  zlim = c(-1, 1),
  main = "Module-age correlations (pseudobulk, n = mice)"
)
```

### Dot plot: module-age association :

```{r}

library(ggplot2)
library(dplyr)
library(ggpubr)   
library(ggtext)   
library(RColorBrewer)


dotplot_data_pub <- data.frame(
  module = sub("^ME", "", names(module_trait_cor)),
  correlation = as.vector(module_trait_cor),
  pvalue = as.vector(module_trait_p),
  stringsAsFactors = FALSE
) %>%
  mutate(
    fdr = p.adjust(pvalue, method = "BH"),
    neg_log_fdr = -log10(pmax(fdr, 1e-300)),
    is_significant = ifelse(fdr < 0.05, "FDR < 0.05", "Not significant"),
    # Factorizar módulos ordenados por el valor de correlación
    module_ordered = factor(module, levels = module[order(correlation)])
  )


p_publication <- ggplot(dotplot_data_pub, aes(x = correlation, y = module_ordered)) +
  
  geom_segment(
    aes(x = 0, xend = correlation, y = module_ordered, yend = module_ordered),
    color = "grey80", linewidth = 0.5
  ) +
  # Línea neutra en el cero
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.4) +
  # Puntos con mapa de color divergente (RdBu)
  geom_point(
    aes(size = neg_log_fdr, fill = correlation, color = is_significant),
    shape = 21, stroke = 0.8
  ) +
  # Escala de color continuo (Azul a Rojo)
  scale_fill_distiller(
    palette = "RdBu", direction = -1, limits = c(-1, 1),
    name = "Correlation (*r*)"
  ) +
  
  scale_color_manual(
    values = c("FDR < 0.05" = "black", "Not significant" = "grey85"),
    name = "Significance"
  ) +
  # Escala de tamaños
  scale_size_continuous(name = "-log<sub>10</sub>(FDR)", range = c(2.5, 7.5)) +
  # Textos y etiquetas con Markdown
  labs(
    x = "Module Eigengene Correlation (*r*)",
    y = "WGCNA Module",
    title = "Overall Module-Age Trait Association",
    subtitle = "Point fill indicates **correlation**, size indicates **-log<sub>10</sub>(FDR)**"
  ) +
  # Tema pubr ajustado
  theme_pubr(base_size = 11, border = TRUE) +
  theme(
    plot.title = element_markdown(face = "bold", size = 13),
    plot.subtitle = element_markdown(size = 9, color = "grey35"),
    axis.title.x = element_markdown(size = 10, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold"),
    legend.title = element_markdown(size = 9, face = "bold"),
    legend.text = element_text(size = 8),
    legend.box = "vertical",
    legend.position = "right",
    panel.grid.major.x = element_line(color = "#f0f0f0", linetype = "dotted")
  )

print(p_publication)

# Exportar en formato vectorial (PDF HD)
ggsave(
  filename = "figure_module_age_association.pdf",
  plot = p_publication,
  width = 5.5,
  height = 6.5,
  device = cairo_pdf,
  dpi = 300
)
```

### Panel: regression bars + p-value dots (module colors)

```{r}
# Total genes per module (grey excluded), reused for point size on the
# FDR panel -- same "number of genes" encoding as the reference figure.
module_gene_counts <- gene_celltype_module %>%
  dplyr::filter(module != "grey") %>%
  dplyr::distinct(module, gene) %>%
  dplyr::count(module, name = "n_genes")

fig_regression_data <- module_trait_table %>%
  dplyr::left_join(module_gene_counts, by = "module") %>%
  dplyr::mutate(
    module_color = module,                       # module name IS a valid R color
    fdr_plot     = pmax(fdr, 1e-4),               # floor for the log10 axis -- ajusta 'fdr' al nombre real de tu columna
    sig_fdr      = fdr < 0.05,
    point_border = ifelse(sig_fdr, "grey15", "grey15"),   # borde rojo si es significativo
    point_stroke = ifelse(sig_fdr, 1.0, 0.4),
    module = factor(module, levels = module[order(correlation)])
  )

# --- Left panel: horizontal bars, one per module, filled with its own color -
p_reg_bars <- ggplot(fig_regression_data, aes(x = correlation, y = module)) +
  geom_col(aes(fill = module_color), color = "grey15", linewidth = 0.3, width = 0.72) +
  scale_fill_identity() +
  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.4) +
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
  labs(x = "Regression (Pearson r vs. age)", y = NULL, title = "Regression") +
  theme_pub(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

# --- Right panel: -log10 FDR as dot position (log10 x-axis), dot size ------
# encodes module size (number of genes), dot filled with the module's color,
# and outlined in red when FDR < 0.05 to flag significant modules at a glance.
p_reg_fdr <- ggplot(fig_regression_data, aes(x = fdr_plot, y = module)) +
  geom_point(aes(size = n_genes, fill = module_color, color = point_border, stroke = point_stroke),
             shape = 21) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_size_continuous(name = "Number\nof genes", range = c(2, 9)) +
  scale_x_log10(limits = c(1e-4, 1), breaks = c(0.0001, 0.001, 0.01, 0.05, 0.1, 1)) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  labs(x = "FDR", y = NULL, title = "FDR") +
  theme_pub(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

p_regression_panel <- p_reg_bars + p_reg_fdr +
  patchwork::plot_layout(widths = c(1.3, 1)) +
  patchwork::plot_annotation(
    title = "Module-age regression across all WGCNA modules",
    subtitle = "Highlighted border= FDR < 0.05",
    theme = theme_pub(base_size = 13)
  )

print(p_regression_panel)

save_pub_pdf(p_regression_panel, "figure_module_age_regression.pdf",
             width = 9, height = max(4, 0.42 * nrow(fig_regression_data) + 1.5))
```

```{r module-trait-publication-panel, fig.width=9, fig.height=7}
# Total genes per module (grey excluded), reused for point size on the
# p-value panel -- same "number of genes" encoding as the reference figure.
module_gene_counts <- gene_celltype_module %>%
  filter(module != "grey") %>%
  distinct(module, gene) %>%
  dplyr::count(module, name = "n_genes")

fig_regression_data <- module_trait_table %>%
  left_join(module_gene_counts, by = "module") %>%
  mutate(
    module_color = module,                       # module name IS a valid R color
    p_plot       = pmax(p_value, 1e-4),           # floor for the log10 axis
    module = factor(module, levels = module[order(correlation)])
  )

# --- Left panel: horizontal bars, one per module, filled with its own color -
p_reg_bars <- ggplot(fig_regression_data, aes(x = correlation, y = module)) +
  geom_col(aes(fill = module_color), color = "grey15", linewidth = 0.3, width = 0.72) +
  scale_fill_identity() +
  geom_vline(xintercept = 0, color = "grey30", linewidth = 0.4) +
  scale_x_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.5)) +
  labs(x = "Regression (Pearson r vs. age)", y = NULL, title = "Regression") +
  theme_pub(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

# --- Right panel: -log10 p-value as dot position (log10 x-axis), dot size --
# encodes module size (number of genes), dot filled with the module's color,
# exactly mirroring the reference figure's p-value / gene-count panel.
p_reg_pvals <- ggplot(fig_regression_data, aes(x = p_plot, y = module)) +
  geom_point(aes(size = n_genes, fill = module_color), shape = 21, color = "grey15", stroke = 0.4) +
  scale_fill_identity() +
  scale_size_continuous(name = "Number\nof genes", range = c(2, 9)) +
  scale_x_log10(limits = c(1e-4, 1), breaks = c(0.0001, 0.001, 0.01, 0.05, 0.1, 1)) +
  geom_vline(xintercept = 0.05, linetype = "dashed", color = "grey40", linewidth = 0.4) +
  labs(x = "P-value", y = NULL, title = "P-value") +
  theme_pub(base_size = 12) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid.major.y = element_blank(),
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5)
  )

p_regression_panel <- p_reg_bars + p_reg_pvals +
  patchwork::plot_layout(widths = c(1.3, 1)) +
  patchwork::plot_annotation(
    title = "Module-age regression across all WGCNA modules",
    theme = theme_pub(base_size = 13)
  )

print(p_regression_panel)

save_pub_pdf(p_regression_panel, "figure_module_age_regression_publication.pdf",
             width = 9, height = max(4, 0.42 * nrow(fig_regression_data) + 1.5))
```

## Hub genes (top kME per module)

`signedKME` gives each gene_celltype feature its correlation (module
membership, kME) to every module eigengene. As in the metacell workflow,
a fixed kME cutoff isn't comparable across modules of different
size/cohesion, so hub genes are defined as the **top kME percentile
within each module**.

```{r hub-genes}
kME_table <- signedKME(
  datExpr, MEs, corFnc = "bicor",
  # datExpr has no NAs after imputation -- all.obs uses bicor's fast path
  corOptions = "use = 'all.obs', maxPOutliers = 0.1, pearsonFallback = 'individual'"
)
colnames(kME_table) <- sub("^kME", "", colnames(kME_table))

gene_celltype_module$kME <- sapply(seq_len(nrow(gene_celltype_module)), function(i) {
  mod <- gene_celltype_module$module[i]
  feat <- gene_celltype_module$gene_celltype[i]
  if (mod == "grey" || !mod %in% colnames(kME_table)) return(NA_real_)
  kME_table[feat, mod]
})

hub_percentile <- 0.90  # top 10% by kME within each module

hub_genes_table <- gene_celltype_module %>%
  filter(module != "grey") %>%
  group_by(module) %>%
  filter(kME >= quantile(kME, hub_percentile, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(module, desc(kME))

write.csv(hub_genes_table, "hub_genes_by_module.csv", row.names = FALSE)

cat("Hub genes per module (top", (1 - hub_percentile) * 100, "% kME):\n")
print(table(hub_genes_table$module))

cat("\nHub gene composition by module x cell type of origin:\n")
print(table(hub_genes_table$module, hub_genes_table$celltype))
```

## Module co-expression network interactions

The TOM computed during `ConstructNetwork()` in the parameter scan above
was cached to disk by hdWGCNA (`tom_outdir/tom_name*.rds`). This chunk
loads that cached TOM (or rebuilds it safely via block-wise computation
if something goes wrong), then every network plot below reuses it.

```{r final-tom-cache}
# ============ CONFIGURACIÓN ============
# Cambia esta ruta si TOM_pseudobulk no está en el working directory actual.
# Puede ser relativa (ej. "TOM_pseudobulk", "resultados/TOM_pseudobulk")
# o absoluta (ej. "/home/usuario/proyecto/TOM_pseudobulk",
#              "C:/Users/usuario/proyecto/TOM_pseudobulk").
tom_outdir_used <- "D:/SandraAnjo/Luis Grilo/WGCNAandothersOlaia/WGCNA_12_08/TOM_pseudobulk"

tom_name_used   <- "pseudobulk_scan"   # prefijo usado en ConstructNetwork(tom_name = ...)
wgcna_name_used <- "pseudobulk"        # el wgcna_name usado en ConstructNetwork()

tom_final_path      <- "final_TOM_combined.rds"
force_recompute_tom <- FALSE
# ========================================

if (!dir.exists(tom_outdir_used)) {
  stop("La carpeta '", tom_outdir_used, "' no existe desde el working directory actual (",
       getwd(), "). Ajusta 'tom_outdir_used' arriba con la ruta correcta.")
}

if (file.exists(tom_final_path) && !force_recompute_tom) {
  
  final_TOM <- readRDS(tom_final_path)
  cat("Loaded cached combined TOM from", tom_final_path, "(",
      nrow(final_TOM), "x", ncol(final_TOM), ")\n")
  
} else {
  
  cat("No cached", tom_final_path, "yet -- building it from the network",
      "already computed in seurat_hd /", tom_outdir_used, "(no recomputation needed).\n")
  cat("Working directory:", getwd(), "\n")
  cat("Contents of", tom_outdir_used, ":\n")
  print(list.files(tom_outdir_used, full.names = FALSE))
  
  # ---- Primary path: GetTOM() reads the combined TOM file that hdWGCNA
  # already wrote to disk during ConstructNetwork() (the "*_TOM*" file,
  # NOT the per-block "*_block.N*" files). This is the expected, cheap path
  # whenever seurat_hd already has the network built (e.g. loaded from .RData).
  cat("Attempting to load TOM via GetTOM(seurat_hd, wgcna_name = '", wgcna_name_used, "')...\n", sep = "")
  
  final_TOM <- tryCatch(
    {
      tom_from_hdwgcna <- GetTOM(seurat_hd, wgcna_name = wgcna_name_used)
      
      # GetTOM() returns row/colnames using hdWGCNA's SANITIZED feature
      # names ("-" instead of "_"); translate back via hdwgcna_name_map so
      # downstream lookups by original gene_celltype name (which use "_")
      # keep working.
      tom_names_original <- hdwgcna_name_map[rownames(tom_from_hdwgcna)]
      stopifnot("Some GetTOM() feature names are missing from hdwgcna_name_map -- name mismatch upstream" =
                  !anyNA(tom_names_original))
      rownames(tom_from_hdwgcna) <- tom_names_original
      colnames(tom_from_hdwgcna) <- tom_names_original
      
      cat("Successfully loaded TOM from hdWGCNA cache (",
          nrow(tom_from_hdwgcna), "x", ncol(tom_from_hdwgcna), ")\n")
      tom_from_hdwgcna
    },
    error = function(e) {
      cat("GetTOM() failed:", conditionMessage(e), "\n")
      cat("Falling back to manually combining the per-block TOM files",
          "already on disk in", tom_outdir_used, "...\n")
      NULL
    }
  )
  
  # ---- Fallback: GetTOM() failed (e.g. path mismatch after mover el proyecto)
  # pero los archivos TOM por bloque de ConstructNetwork() siguen en
  # tom_outdir_used. Los combina directamente usando el mapeo gen->bloque
  # ya guardado en `net` (de GetNetworkData(), calculado más arriba en el
  # script) -- sin necesidad de rerun blockwiseModules().
  if (is.null(final_TOM)) {
    
    stopifnot("`net` (from GetNetworkData(seurat_hd, ...)) is not in the environment -- rerun the 'blockwise-final-after-scan' chunk first" =
                exists("net") && !is.null(net$blockGenes))
    
    block_files <- list.files(
      tom_outdir_used,
      pattern    = paste0("^", tom_name_used, ".*block.*\\.(rds|RData|rda)$"),
      full.names = TRUE,
      ignore.case = TRUE
    )
    cat("Found", length(block_files), "candidate block TOM file(s):\n")
    print(block_files)

    if (length(block_files) == 0) {
      stop("No per-block TOM files matching '", tom_name_used, "*block*' found in ", tom_outdir_used,
           " -- check tom_name_used/tom_outdir_used above match what ConstructNetwork() actually used.")
    }

    load_tom_block <- function(path) {
      if (grepl("\\.rds$", path, ignore.case = TRUE)) {
        readRDS(path)
      } else {
        # .RData/.rda block files from blockwiseModules(saveTOMs = TRUE)
        # store the TOM as a single object -- typically named 'TOM'.
        e <- new.env()
        loaded <- load(path, envir = e)
        e[[loaded[1]]]
      }
    }

    all_features <- unlist(net$blockGenes)
    final_TOM <- matrix(NA_real_, nrow = length(all_features), ncol = length(all_features))
    rownames(final_TOM) <- colnames(datExpr)[all_features]
    colnames(final_TOM) <- colnames(datExpr)[all_features]

    n_combined <- 0
    for (block_idx in seq_along(net$blockGenes)) {
      block_file <- block_files[grepl(paste0("block\\.?", block_idx, "(\\.|$)"), block_files)]
      if (length(block_file) == 0) {
        warning("No TOM file found for block ", block_idx, " -- leaving those entries as NA")
        next
      }
      tom_block <- load_tom_block(block_file[1])
      block_features <- net$blockGenes[[block_idx]]
      final_TOM[block_features, block_features] <- as.matrix(tom_block)
      n_combined <- n_combined + 1
    }

    stopifnot("Could not reconstruct final_TOM from any block file" = n_combined > 0)
    cat("Combined", n_combined, "of", length(net$blockGenes), "block TOMs into one",
        nrow(final_TOM), "x", ncol(final_TOM), "matrix\n")

    tom_names_original <- hdwgcna_name_map[rownames(final_TOM)]
    stopifnot("Some block-TOM feature names are missing from hdwgcna_name_map -- name mismatch upstream" =
                !anyNA(tom_names_original))
    rownames(final_TOM) <- tom_names_original
    colnames(final_TOM) <- tom_names_original
  }

  stopifnot(!anyNA(diag(final_TOM)))  # diagonal should be 1, never NA

  # Cache the final TOM for future runs
  saveRDS(final_TOM, tom_final_path)
  cat("Cached combined TOM to", tom_final_path, "\n")
}

stopifnot(nrow(final_TOM) == ncol(final_TOM))
stopifnot(all(rownames(final_TOM) == colnames(final_TOM)))
cat("Final TOM ready:", nrow(final_TOM), "x", ncol(final_TOM), "features\n")
```

### Per-module network (top hub genes, intra-module TOM edges)

One circular network per module, edges = TOM similarity among that
module's top-kME features, node labels = gene name (cell-type suffix
dropped for readability, but each node is still a single `gene_celltype`
feature).

```{r per-module-network, fig.width=7, fig.height=7}
library(igraph)
library(ggraph)
library(dplyr)
library(reshape2)


dir.create("module_networks", showWarnings = FALSE)

n_show <- 25           # Número de top genes por módulo a evaluar
top_edge_prop <- 0.25  # Mostrar solo el Top 25% de aristas TOM más fuertes


plot_module_network <- function(mod, n_show = 25, top_edge_prop = 0.25) {
  # Limpiar el prefijo "ME" si está presente
  mod_clean <- sub("^ME", "", mod)

  # Extraer y ordenar features por kME
  feats <- kME_table[, mod_clean]
  names(feats) <- rownames(kME_table)
  feats <- feats[!is.na(feats)]

  top_feats <- names(sort(feats, decreasing = TRUE))[seq_len(min(n_show, length(feats)))]
  top_feats <- intersect(top_feats, rownames(final_TOM))

  if (length(top_feats) < 3) return(NULL)

  kme_vals <- feats[top_feats]

  # Sub-matriz TOM y extracción de aristas
  sub_tom <- final_TOM[top_feats, top_feats]
  sub_tom[upper.tri(sub_tom, diag = TRUE)] <- NA

  edges <- reshape2::melt(sub_tom) %>%
    dplyr::rename(from = Var1, to = Var2, weight = value) %>%
    dplyr::filter(!is.na(weight))

  # Filtrar solo el top % de conexiones TOM más fuertes (evita la bola de pelo)
  tom_cutoff <- quantile(edges$weight, probs = 1 - top_edge_prop, na.rm = TRUE)
  edges_filtered <- edges %>% dplyr::filter(weight >= tom_cutoff)

  # Creación del grafo inicial
  g <- igraph::graph_from_data_frame(edges_filtered, vertices = data.frame(name = top_feats), directed = FALSE)

  # --- ELIMINAR NODOS AISLADOS (Grado = 0) ---
  g <- igraph::delete_vertices(g, V(g)[igraph::degree(g) == 0])

  # Verificar si quedan suficientes nodos para graficar
  if (igraph::vcount(g) < 3) return(NULL)

  # Asignar atributos a los nodos
  V(g)$gene   <- sub("_[^_]+$", "", V(g)$name)
  V(g)$kME    <- kme_vals[V(g)$name]
  V(g)$degree <- igraph::degree(g)

  mod_color <- mod_clean

  # Renderizado visual con ggraph (Layout 'stress' para centralizar hubs)
  p <- ggraph(g, layout = "stress") +
    # Aristas
    geom_edge_link(
      aes(alpha = weight, width = weight),
      color = "grey40",
      show.legend = FALSE
    ) +
    scale_edge_alpha_continuous(range = c(0.15, 0.65)) +
    scale_edge_width_continuous(range = c(0.3, 1.1)) +

    # Nodos (El tamaño escala según kME)
    geom_node_point(
      aes(size = kME),
      fill = mod_color,
      color = "black",
      shape = 21,
      stroke = 0.5
    ) +
    scale_size_continuous(range = c(3, 7.5), name = "Module Membership\n(kME)") +

    # Etiquetas con halo blanco de protección tras el texto
    geom_node_text(
      aes(label = gene),
      repel = TRUE,
      size = 3.2,
      fontface = "italic",
      bg.color = "white",
      bg.r = 0.15,
      max.overlaps = 30
    ) +

    # Títulos y metadatos
    labs(
      title = paste0("Module ", mod_clean, " Co-expression Network"),
      subtitle = paste0("Showing connected top genes | Top ", round(top_edge_prop * 100), "% TOM connections")
    ) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 13, margin = margin(b = 3)),
      plot.subtitle = element_text(hjust = 0.5, size = 8.5, color = "grey35", margin = margin(b = 10)),
      legend.position = "right",
      legend.title = element_text(size = 8.5, face = "bold"),
      legend.text = element_text(size = 8),
      plot.margin = margin(12, 12, 12, 12)
    )

  return(p)
}


for (mod in module_names) {
  cat("Procesando módulo:", mod, "...\n")

  p <- plot_module_network(mod, n_show = n_show, top_edge_prop = top_edge_prop)

  if (is.null(p)) {
    cat("  -> Omitido (insuficientes genes conectados)\n")
    next
  }

  # Imprimir en consola / visor de RStudio
  print(p)

  # Guardar archivo PDF individual
  file_name <- file.path("module_networks", paste0("network_", mod, ".pdf"))
  ggsave(file_name, p, width = 6.5, height = 6, limitsize = FALSE, dpi = 300)
  cat("  -> Guardado en:", file_name, "\n")
}


```

### Comprehensive network: all modules, hub genes, inter-module edges in grey

Combines every module's hub genes into one graph. Intra-module edges are
colored by module (top 15% kept per module, minimum 5, so the plot stays
legible); inter-module edges -- where two hub genes from different
modules are still strongly TOM-connected -- are shown in grey,
highlighting cross-module structure.

```{r comprehensive-network, fig.width=14, fig.height=12}
# n_hubs_combined <- 15
#
# hub_feats_combined <- hub_genes_table %>%
#   group_by(module) %>%
#   slice_max(kME, n = n_hubs_combined, with_ties = FALSE) %>%
#   ungroup() %>%
#   pull(gene_celltype) %>%
#   intersect(rownames(final_TOM)) %>%
#   unique()
#
# cur_tom <- final_TOM[hub_feats_combined, hub_feats_combined]
# cur_tom[upper.tri(cur_tom, diag = TRUE)] <- NA
#
# cur_network <- reshape2::melt(cur_tom) %>%
#   dplyr::rename(feat1 = Var1, feat2 = Var2, weight = value) %>%
#   filter(!is.na(weight))
#
# feat_module_lookup <- gene_celltype_module %>%
#   filter(gene_celltype %in% hub_feats_combined) %>%
#   distinct(gene_celltype, module) %>%
#   tibble::deframe()
#
# cur_network <- cur_network %>%
#   mutate(
#     module1 = feat_module_lookup[as.character(feat1)],
#     module2 = feat_module_lookup[as.character(feat2)],
#     edge_color = ifelse(module1 == module2, module1, "grey85")
#   )
#
# edge_percent <- 0.15
# min_edges_per_module <- 5
#
# cur_network_intra <- cur_network %>%
#   filter(edge_color != "grey85") %>%
#   group_by(edge_color) %>%
#   group_modify(~ slice_max(.x, order_by = weight,
#                             n = pmax(min_edges_per_module, round(nrow(.x) * edge_percent)))) %>%
#   ungroup()
#
# cur_network_inter <- cur_network %>% filter(edge_color == "grey85")
# cur_network_final <- bind_rows(cur_network_intra, cur_network_inter)
#
# g <- graph_from_data_frame(
#   cur_network_final %>% dplyr::select(feat1, feat2, weight, edge_color),
#   directed = FALSE
# )
# V(g)$module <- feat_module_lookup[V(g)$name]
# V(g)$gene   <- sub("_[^_]+$", "", V(g)$name)
#
# module_palette <- setNames(unique(V(g)$module), unique(V(g)$module))  # module name IS the color
#
# p_combined <- ggraph(g, layout = "fr") +
#   geom_edge_link(aes(alpha = weight, color = edge_color), width = 0.3) +
#   geom_node_point(aes(color = module), size = 3) +
#   geom_node_text(aes(label = gene), repel = TRUE, max.overlaps = 20, size = 2.3, fontface = "italic") +
#   scale_color_manual(values = module_palette, name = "Module") +
#   scale_edge_color_manual(values = c(module_palette, "grey85" = "grey85"), guide = "none") +
#   scale_edge_alpha_continuous(range = c(0.1, 0.6), name = "TOM") +
#   theme_void() +
#   theme(legend.position = "right")
#
# ggsave("comprehensive_module_network.pdf", p_combined, width = 14, height = 12, limitsize = FALSE)
# print(p_combined)
```

```{r}

library(igraph)
library(ggraph)
library(tidyverse)
library(uwot)      # Para el cálculo de UMAP
library(ggforce)   # Para los sombreados (hulls)


n_hubs_combined <- 15

hub_feats_table <- hub_genes_table %>%
  dplyr::mutate(module_clean = sub("^ME", "", module)) %>%
  dplyr::group_by(module_clean) %>%
  dplyr::slice_max(kME, n = n_hubs_combined, with_ties = FALSE) %>%
  dplyr::ungroup()

hub_feats_combined <- hub_feats_table %>%
  dplyr::pull(gene_celltype) %>%
  intersect(rownames(final_TOM)) %>%
  unique()


cur_tom_full <- final_TOM[hub_feats_combined, hub_feats_combined]
tom_dist <- as.dist(1 - cur_tom_full)

set.seed(42)
umap_res <- uwot::umap(
  tom_dist,
  n_neighbors = min(15, length(hub_feats_combined) - 1),
  min_dist = 0.3,
  metric = "euclidean"
)

layout_umap <- data.frame(
  name = hub_feats_combined,
  x = umap_res[, 1],
  y = umap_res[, 2]
)

cur_tom <- cur_tom_full
cur_tom[upper.tri(cur_tom, diag = TRUE)] <- NA

cur_network <- reshape2::melt(cur_tom) %>%
  dplyr::rename(feat1 = Var1, feat2 = Var2, weight = value) %>%
  dplyr::filter(!is.na(weight))

feat_module_lookup <- gene_celltype_module %>%
  dplyr::mutate(module_clean = sub("^ME", "", module)) %>%
  dplyr::filter(gene_celltype %in% hub_feats_combined) %>%
  dplyr::distinct(gene_celltype, module_clean) %>%
  tibble::deframe()

cur_network <- cur_network %>%
  dplyr::mutate(
    module1 = feat_module_lookup[as.character(feat1)],
    module2 = feat_module_lookup[as.character(feat2)],
    edge_color = ifelse(module1 == module2, module1, "grey85")
  )

# Filtrado de aristas
edge_percent <- 0.15
min_edges_per_module <- 5

cur_network_intra <- cur_network %>%
  dplyr::filter(edge_color != "grey85") %>%
  dplyr::group_by(edge_color) %>%
  dplyr::group_modify(~ dplyr::slice_max(.x, order_by = weight,
                                         n = pmax(min_edges_per_module, round(nrow(.x) * edge_percent)))) %>%
  dplyr::ungroup()

cur_network_inter <- cur_network %>%
  dplyr::filter(edge_color == "grey85") %>%
  dplyr::slice_max(weight, prop = 0.02)

cur_network_final <- dplyr::bind_rows(cur_network_intra, cur_network_inter)

# -----------------------------------------------------------------------------
# 4. Construcción del Grafo (Llamadas explícitas a igraph)
# -----------------------------------------------------------------------------
g <- igraph::graph_from_data_frame(
  cur_network_final %>% dplyr::select(feat1, feat2, weight, edge_color),
  directed = FALSE
)

V(g)$module <- feat_module_lookup[V(g)$name]
V(g)$gene   <- sub("_[^_]+$", "", V(g)$name)

# Cambio clave aquí: llamamos directamente a igraph::degree
V(g)$degree <- igraph::degree(g)

layout_graph <- ggraph::create_layout(
  g, layout = "manual",
  x = layout_umap$x[match(V(g)$name, layout_umap$name)],
  y = layout_umap$y[match(V(g)$name, layout_umap$name)]
)

modules_present <- unique(V(g)$module)
module_palette <- setNames(modules_present, modules_present)


p_combined <- ggraph(layout_graph) +
  geom_mark_hull(
    aes(x = x, y = y, fill = module, group = module),
    concavity = 2,
    expand = unit(4, "mm"),
    alpha = 0.12,
    color = NA,
    show.legend = FALSE
  ) +
  geom_edge_link(
    aes(filter = (edge_color == "grey85"), alpha = weight),
    color = "grey80", width = 0.25, linetype = "dashed"
  ) +
  geom_edge_link(
    aes(filter = (edge_color != "grey85"), alpha = weight, color = edge_color),
    width = 0.5
  ) +
  geom_node_point(aes(color = module, size = degree), alpha = 0.9) +
  geom_node_text(
    aes(label = gene),
    repel = TRUE,
    max.overlaps = 25,
    size = 2.8,
    fontface = "italic",
    point.padding = unit(0.2, "lines")
  ) +
  scale_color_manual(values = module_palette, name = "Module") +
  scale_fill_manual(values = module_palette) +
  scale_edge_color_manual(values = c(module_palette, "grey85" = "grey85"), guide = "none") +
  scale_edge_alpha_continuous(range = c(0.15, 0.7), name = "TOM Co-expression") +
  scale_size_continuous(range = c(2.5, 6), name = "Degree") +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 10),
    legend.text = element_text(size = 9),
    plot.margin = margin(10, 10, 10, 10),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  ) +
  labs(title = "Integrated Co-expression Hub Network (UMAP Layout)")

ggsave("comprehensive_module_network_umap.pdf", p_combined, width = 14, height = 12, limitsize = FALSE, dpi = 300)
print(p_combined)
```

## Entrez mapping and enrichment universe

Every downstream enrichment (GO / KEGG / Reactome, always split by cell
type -- see next section) is tested against the same background
universe: every gene that entered the WGCNA network
(`gene_celltype_module`, grey module excluded upstream), mapped to
Entrez IDs once here and reused everywhere below.
`convert_to_entrez_flexible()` is the shared helper that maps a named
list of gene vectors (one per group -- module, or module x cell type) to
Entrez IDs, dropping groups with too few mapped genes.

```{r entrez-universe-and-helpers}
cat("Genes in gene_celltype_module (grey excluded):",
    length(unique(gene_celltype_module$gene)), "\n")
cat("Genes in the matrix that entered blockwiseModules (datExpr):",
    ncol(datExpr), "\n")

universe_symbols <- gene_celltype_module %>%
  distinct(gene) %>%
  pull(gene) %>%
  unique()

universe_entrez_map <- bitr(
  universe_symbols, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db
)

cat("Symbols in universe:", length(universe_symbols), "\n")
cat("Mapped to Entrez:", nrow(universe_entrez_map), "\n")
cat("Not mapped:", length(universe_symbols) - nrow(universe_entrez_map), "\n")

universe_entrez <- universe_entrez_map$ENTREZID %>% unique() %>% na.omit()
cat("Final Entrez universe:", length(universe_entrez), "\n\n")

convert_to_entrez_flexible <- function(gene_list, min_genes = 1, verbose = TRUE) {
  result <- list()
  dropped_modules <- data.frame(
    module = character(),
    n_genes_original = integer(),
    n_genes_mapped = integer(),
    reason = character()
  )

  for (mod in names(gene_list)) {
    genes <- gene_list[[mod]]

    converted <- tryCatch(
      bitr(genes, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Mm.eg.db),
      error = function(e) NULL
    )

    if (is.null(converted) || nrow(converted) == 0) {
      if (verbose) {
        dropped_modules <- rbind(dropped_modules, data.frame(
          module = mod, n_genes_original = length(genes),
          n_genes_mapped = 0, reason = "No genes mapped to ENTREZID"
        ))
      }
      next
    }

    entrez_ids <- unique(na.omit(converted$ENTREZID))

    if (length(entrez_ids) < min_genes) {
      if (verbose) {
        dropped_modules <- rbind(dropped_modules, data.frame(
          module = mod, n_genes_original = length(genes),
          n_genes_mapped = length(entrez_ids),
          reason = paste0("Only ", length(entrez_ids), " genes mapped (min: ", min_genes, ")")
        ))
      }
      next
    }

    result[[mod]] <- entrez_ids
  }

  if (verbose && nrow(dropped_modules) > 0) {
    cat("\nGroups excluded:\n")
    print(dropped_modules, row.names = FALSE)
  }

  return(result)
}
```

## Functional enrichment by cell type within each module (GO / KEGG / Reactome)

```{r enrichment-by-celltype-within-module}
gene_counts_by_module_celltype <- gene_celltype_module %>%
  filter(module != "grey") %>%
  dplyr::count(module, celltype, name = "n_genes")


run_enrichment_by_celltype <- function(fun_name,
                                       database_label,
                                       extra_args = list(),
                                       cache_file,
                                       force_recompute = TRUE) {
  
  if (file.exists(cache_file) && !force_recompute) {
    df_cached <- read.csv(cache_file, stringsAsFactors = FALSE)
    cat("[", database_label, "] Loaded cached results from ", cache_file,
        " (", nrow(df_cached), " rows). Set force_recompute_enrichment_celltype <- TRUE to redo.\n", sep = "")
    return(df_cached)
  }
  
  result_list <- list()
  modules_all <- unique(gene_celltype_module$module[gene_celltype_module$module != "grey"])
  
  for (mod in modules_all) {
    genes_mod <- gene_celltype_module %>% filter(module == mod)
    
    gene_list_ct <- split(genes_mod$gene, genes_mod$celltype)
    gene_list_ct <- lapply(gene_list_ct, unique)
    
    gene_list_ct_entrez <- convert_to_entrez_flexible(gene_list_ct, min_genes = 3, verbose = FALSE)
    if (length(gene_list_ct_entrez) == 0) next
    
    call_args <- c(
      list(
        geneCluster   = gene_list_ct_entrez,
        fun           = fun_name,
        universe      = universe_entrez,
        pAdjustMethod = "BH",
        pvalueCutoff  = 0.05
      ),
      extra_args
    )
    
    ego_mod <- tryCatch(
      do.call(compareCluster, call_args),
      error = function(e) {
        message("[", database_label, "] Module ", mod, ": failed (", conditionMessage(e), ")")
        NULL
      }
    )
    if (is.null(ego_mod)) next
    
    df_mod <- as.data.frame(ego_mod)
    if (nrow(df_mod) == 0) next
    
    df_mod$module   <- mod
    df_mod$database <- database_label
    colnames(df_mod)[colnames(df_mod) == "Cluster"] <- "celltype"
    result_list[[mod]] <- df_mod
  }
  
  df_out <- bind_rows(result_list)
  write.csv(df_out, cache_file, row.names = TRUE)
  cat("[", database_label, "] Computed fresh and saved to ", cache_file, "\n", sep = "")
  df_out
}

force_recompute_enrichment_celltype <- TRUE

go_bp_by_module_celltype <- run_enrichment_by_celltype(
  fun_name       = "enrichGO",
  database_label = "GO_BP",
  extra_args     = list(OrgDb = org.Mm.eg.db, ont = "BP", qvalueCutoff = 0.2, readable = TRUE),
  cache_file     = "GO_BP_by_module_and_celltype.csv",
  force_recompute = force_recompute_enrichment_celltype
)

kegg_by_module_celltype <- run_enrichment_by_celltype(
  fun_name       = "enrichKEGG",
  database_label = "KEGG",
  extra_args     = list(organism = "mmu"),
  cache_file     = "KEGG_by_module_and_celltype.csv",
  force_recompute = force_recompute_enrichment_celltype
)

reactome_by_module_celltype <- run_enrichment_by_celltype(
  fun_name       = "enrichPathway",
  database_label = "Reactome",
  extra_args     = list(organism = "mouse"),
  cache_file     = "Reactome_by_module_and_celltype.csv",
  force_recompute = force_recompute_enrichment_celltype
)


filter_out_kegg_disease <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  is_disease <- grepl("^[a-z]{3,4}05\\d{3}$", df$ID)
  n_removed <- sum(is_disease)
  if (n_removed > 0) {
    cat("KEGG: removed", n_removed, "term(s) from the 'Human Diseases' category (map 05xxx), e.g.:\n")
    print(head(unique(df$Description[is_disease]), 5))
  }
  df[!is_disease, , drop = FALSE]
}


get_reactome_disease_ids <- function(species = "Mus musculus") {
  url <- paste0("https://reactome.org/ContentService/data/eventsHierarchy/",
                utils::URLencode(species))
  res <- tryCatch(jsonlite::fromJSON(url, simplifyVector = FALSE), error = function(e) NULL)
  if (is.null(res)) return(character(0))
  
  collect_ids <- function(node) {
    ids <- character(0)
    if (!is.null(node$stId)) ids <- c(ids, node$stId)
    if (!is.null(node$children)) ids <- c(ids, unlist(lapply(node$children, collect_ids)))
    ids
  }
  
  disease_nodes <- Filter(function(n) isTRUE(n$name == "Disease"), res)
  if (length(disease_nodes) == 0) return(character(0))
  unique(unlist(lapply(disease_nodes, collect_ids)))
}

reactome_disease_ids <- tryCatch(get_reactome_disease_ids("Mus musculus"), error = function(e) character(0))
if (length(reactome_disease_ids) == 0) {
  cat("Reactome: could not resolve the 'Disease' branch from the ContentService API",
      "(offline, or the service is unavailable) -- falling back to keyword-only filtering.\n")
} else {
  cat("Reactome: resolved", length(reactome_disease_ids), "pathway IDs under the 'Disease' top-level category.\n")
}

# Keyword fallback / extra safety net -- catches disease terms even when the
# ID-based hierarchy above is unavailable or misses a branch.
reactome_disease_keywords <- c(
  "cancer", "carcinoma", "tumou?r", "leukemia", "lymphoma", "melanoma",
  "sarcoma", "glioma", "neoplasm", "metastas", "diabetes", "alzheimer",
  "parkinson", "huntington", "prion disease", "amyotrophic lateral sclerosis",
  "hepatitis", "influenza", "malaria", "tuberculosis", "leishmania",
  "trypanosom", "sars-cov|coronavirus|covid", "hiv infection|aids",
  "addiction", "substance abuse", "infection (by|with|of)", "viral inf",
  "bacterial inf", "human papillomavirus|\\bhpv\\b", "measles", "salmonella",
  "listeria|shigella", "legionella", "yersinia", "chlamydia",
  "helicobacter pylori", "asthma", "rheumatoid arthritis", "systemic lupus",
  "psoriasis", "graft-versus-host", "diseases associated with"
)
reactome_disease_kw_regex <- paste(reactome_disease_keywords, collapse = "|")

filter_out_reactome_disease <- function(df, disease_ids) {
  if (is.null(df) || nrow(df) == 0) return(df)
  by_id      <- df$ID %in% disease_ids
  by_keyword <- grepl(reactome_disease_kw_regex, df$Description, ignore.case = TRUE)
  is_disease <- by_id | by_keyword
  n_removed <- sum(is_disease)
  if (n_removed > 0) {
    cat("Reactome: removed", n_removed, "disease-related term(s), e.g.:\n")
    print(head(unique(df$Description[is_disease]), 5))
  }
  df[!is_disease, , drop = FALSE]
}

kegg_by_module_celltype     <- filter_out_kegg_disease(kegg_by_module_celltype)
reactome_by_module_celltype <- filter_out_reactome_disease(reactome_by_module_celltype, reactome_disease_ids)

# KEGG/Reactome results are readable=FALSE by default (Entrez IDs in geneID);
# convert them to gene symbols so downstream code is uniform across databases.
make_geneid_readable <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)
  df$geneID <- vapply(df$geneID, function(ids) {
    entrez <- strsplit(ids, "/")[[1]]
    sym <- suppressMessages(
      tryCatch(bitr(entrez, fromType = "ENTREZID", toType = "SYMBOL", OrgDb = org.Mm.eg.db),
               error = function(e) NULL)
    )
    if (is.null(sym) || nrow(sym) == 0) return(ids)
    paste(sym$SYMBOL[match(entrez, sym$ENTREZID)], collapse = "/")
  }, character(1))
  df
}

kegg_by_module_celltype     <- make_geneid_readable(kegg_by_module_celltype)
reactome_by_module_celltype <- make_geneid_readable(reactome_by_module_celltype)

# Stacked long table used by every plotting function below
enrichment_by_module_celltype <- bind_rows(
  go_bp_by_module_celltype,
  kegg_by_module_celltype,
  reactome_by_module_celltype
)

# Kept as an alias so earlier chunks referring to go_by_module_celltype still work
go_by_module_celltype <- go_bp_by_module_celltype

cat("\nTerms found, by database x module x cell type:\n")
print(enrichment_by_module_celltype %>%
        dplyr::count(database, module, celltype, name = "n_sig_terms") %>%
        tidyr::pivot_wider(names_from = celltype, values_from = n_sig_terms, values_fill = 0))
```

```{r}
library(tidyverse)
library(stringr)


output_dir <- "dotplots_GO_Reactome_by_module"
dir.create(output_dir, showWarnings = FALSE)


df_dotplot <- enrichment_by_module_celltype %>%
  dplyr::filter(database %in% c("GO_BP", "Reactome")) %>%
  dplyr::mutate(
    GeneRatio_num = sapply(GeneRatio, function(x) {
      nums <- as.numeric(unlist(strsplit(x, "/")))
      nums[1] / nums[2]
    })
  )


top_n_terms <- 5


modules_list <- unique(df_dotplot$module)

for (mod in modules_list) {
  df_mod <- df_dotplot %>%
    dplyr::filter(module == mod)
  
  if (nrow(df_mod) == 0) next
  
  
  df_top <- df_mod %>%
    dplyr::group_by(database, celltype) %>%
    dplyr::slice_min(order_by = p.adjust, n = top_n_terms, with_ties = FALSE) %>%
    dplyr::ungroup()
  
  
  df_top <- df_top %>%
    dplyr::mutate(
      Description_wrapped = stringr::str_wrap(Description, width = 45),
      Description_wrapped = factor(Description_wrapped, levels = unique(Description_wrapped[order(p.adjust, decreasing = TRUE)]))
    )
  
  
  n_unique_terms <- length(unique(df_top$Description_wrapped))
  n_celltypes    <- length(unique(df_top$celltype))
  
  pdf_height <- max(5, n_unique_terms * 0.35 + 2)
  pdf_width  <- max(7, n_celltypes * 1.2 + 3)
  
  
  p <- ggplot(df_top, aes(x = celltype, y = Description_wrapped, size = GeneRatio_num, color = p.adjust)) +
    geom_point(alpha = 0.9) +
    scale_color_gradient(low = "#d73027", high = "#4575b4", name = "p.adjust") +
    scale_size_continuous(name = "Gene Ratio", range = c(2, 7)) +
    facet_grid(database ~ ., scales = "free_y", space = "free_y") +
    theme_bw(base_size = 11) +
    labs(
      title = paste0("Module ", mod, " | Enrichment (GO_BP & Reactome)"),
      x = "Cell Type",
      y = NULL
    ) +
    theme(
      plot.title       = element_text(face = "bold", size = 12, hjust = 0.5),
      axis.text.x      = element_text(angle = 45, hjust = 1, vjust = 1, face = "bold", color = "black"),
      axis.text.y      = element_text(size = 9, color = "black"),
      strip.background = element_rect(fill = "grey90", color = "grey50"),
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )
  
  
  file_path <- file.path(output_dir, paste0("module_", mod, "_GO_Reactome_dotplot.pdf"))
  
  ggplot2::ggsave(
    filename = file_path,
    plot     = p,
    width    = pdf_width,
    height   = pdf_height,
    limitsize = FALSE
  )
  
  cat("Guardado:", file_path, "\n")
}
```

### Activity-score proxy for coloring (kME-weighted, per term x cell type)

`enrichGO`/`enrichKEGG`/`enrichPathway` are over-representation tests:
they return significance (p-value) and gene counts, but -- unlike a
differential-expression contrast -- no directional "up/down" statistic.
To still be able to color bubbles the way IPA-style figures do (a
Z-score-like diverging scale), we build a proxy: for every module x cell
type x term, take the **mean kME (module membership)** of the genes
annotated to that term in that cell type, then standardize (z-score)
within each module. Modules/terms with higher-than-average membership
strength score positive (red), lower than average score negative (blue)
-- this reflects how "core" vs "peripheral" the contributing genes are
to the module, not differential expression.

```{r activity-score-proxy}
# gene -> kME lookup, per module (mean kME across cell types the gene appears in)
gene_kme_lookup <- gene_celltype_module %>%
  filter(module != "grey") %>%
  group_by(module, gene) %>%
  summarise(kme = mean(kME, na.rm = TRUE), .groups = "drop")

add_activity_score <- function(df) {
  if (is.null(df) || nrow(df) == 0) return(df)

  kme_key <- paste(gene_kme_lookup$module, gene_kme_lookup$gene)
  kme_val <- gene_kme_lookup$kme

  df$mean_kme <- vapply(seq_len(nrow(df)), function(i) {
    genes <- strsplit(df$geneID[i], "/")[[1]]
    vals  <- kme_val[match(paste(df$module[i], genes), kme_key)]
    mean(vals, na.rm = TRUE)
  }, numeric(1))

  df <- df %>%
    group_by(module) %>%
    mutate(activity_score = as.numeric(scale(mean_kme))) %>%
    ungroup()

  df$activity_score[is.nan(df$activity_score) | is.na(df$activity_score)] <- 0
  df
}

enrichment_by_module_celltype <- add_activity_score(enrichment_by_module_celltype)
go_by_module_celltype         <- enrichment_by_module_celltype %>% filter(database == "GO_BP")
```

### Heatmaps

```{r plot-composition-heatmap-extension, fig.width=9, fig.height=6}
composition_pct <- gene_counts_by_module_celltype %>%
  group_by(module) %>%
  mutate(pct = n_genes / sum(n_genes) * 100) %>%
  ungroup()

p_composition <- ggplot(composition_pct, aes(x = module, y = celltype, fill = pct)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(name = "% genes") +
  labs(title = "Celular composition per module ",
       x = "Module", y = "Cell type") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_composition)
save_pub_pdf(p_composition, "heatmap_composicion_celular_por_modulo.pdf", width = 9, height = 6)
```

## Module trajectory shape (nested models)

Instead of a single Pearson correlation, each module's eigengene
trajectory across age is classified with **nested model comparison**:
null (`~1`) vs linear (`~age`) vs quadratic (`~age + age^2`), compared
pairwise with `anova()` (nested F-test). The quadratic term is kept only
if it explains significantly more variance than the linear model alone.
The winning model's fitted values are then passed to an edge-aware
classifier that looks at *where* along the age axis the min/max of the
curve falls, producing an honest label: Increasing, Decreasing,
U-shaped, or Inverted U-shaped (not just "significant/not significant").

Cached to `module_trajectory_shape.csv`: if that file exists it's read
instead of refit -- delete it (or set
`force_recompute_trajectory <- TRUE`) after any upstream change to
`datExpr`/module composition.

```{r module-trajectory-nested-models}
trajectory_cache <- "module_trajectory_shape.csv"
force_recompute_trajectory <- FALSE

df_traj <- data.frame(MEs, age_num = age_num, sample_id = rownames(MEs))
# module_names already defined in the module-eigengenes chunk above


classify_curve <- function(age_num, fit, edge_frac = 0.15) {
  ord   <- order(age_num)
  fit_s <- fit[ord]
  n     <- length(fit_s)

  idx_min <- which.min(fit_s); idx_max <- which.max(fit_s)
  pos_min <- (idx_min - 1) / (n - 1); pos_max <- (idx_max - 1) / (n - 1)

  net_change  <- fit_s[n] - fit_s[1]
  total_range <- max(fit_s) - min(fit_s)

  is_min_edge <- pos_min <= edge_frac || pos_min >= (1 - edge_frac)
  is_max_edge <- pos_max <= edge_frac || pos_max >= (1 - edge_frac)

  category <- if (is_min_edge && is_max_edge) {
    if (net_change >= 0) "Increasing" else "Decreasing"
  } else if (!is_min_edge && is_max_edge) {
    "U-shaped"
  } else if (!is_max_edge && is_min_edge) {
    "Inverted U-shaped"
  } else {
    if (net_change >= 0) "Increasing" else "Decreasing"
  }
  data.frame(net_change = net_change, total_range = total_range, category = category)
}

if (file.exists(trajectory_cache) && !force_recompute_trajectory) {

  module_trajectory <- read.csv(trajectory_cache, stringsAsFactors = FALSE)
  cat("Loaded cached trajectory results from", trajectory_cache, "(",
      nrow(module_trajectory), "modules). Set force_recompute_trajectory <- TRUE to redo.\n")

} else {

  trajectory_results <- list()

  for (mod in module_names) {
    f_null <- as.formula(paste0("`", mod, "` ~ 1"))
    f_lin  <- as.formula(paste0("`", mod, "` ~ age_num"))
    f_quad <- as.formula(paste0("`", mod, "` ~ age_num + I(age_num^2)"))

    m_null <- tryCatch(lm(f_null, data = df_traj), error = function(e) NULL)
    m_lin  <- tryCatch(lm(f_lin,  data = df_traj), error = function(e) NULL)
    m_quad <- tryCatch(lm(f_quad, data = df_traj), error = function(e) NULL)
    if (is.null(m_null) || is.null(m_lin) || is.null(m_quad)) next

    p_lin_vs_null <- anova(m_null, m_lin)$`Pr(>F)`[2]
    p_quad_vs_lin <- anova(m_lin, m_quad)$`Pr(>F)`[2]

    if (!is.na(p_quad_vs_lin) && p_quad_vs_lin < 0.05) {
      shape <- "quadratic"; best_model <- m_quad; best_p <- p_quad_vs_lin
    } else {
      shape <- "linear"; best_model <- m_lin; best_p <- p_lin_vs_null
    }

    cls <- classify_curve(df_traj$age_num, fitted(best_model))

    trajectory_results[[mod]] <- data.frame(
      module = sub("^ME", "", mod),
      shape = shape,
      p_value = best_p,
      r_squared = summary(best_model)$r.squared,
      category = cls$category,
      net_change = cls$net_change,
      total_range = cls$total_range,
      n_mice = nrow(df_traj)
    )
  }

  module_trajectory <- bind_rows(trajectory_results) %>%
    mutate(fdr = p.adjust(p_value, method = "BH")) %>%
    arrange(fdr)

  write.csv(module_trajectory, trajectory_cache, row.names = FALSE)
  cat("Computed fresh and saved to", trajectory_cache, "\n")
}

cat("Module trajectory shapes (n =", nrow(df_traj), "mice):\n")
print(module_trajectory)

significant_trajectory_modules <- module_trajectory %>% filter(p_value < 0.05) %>% pull(module)
cat("\nModules with a significant age trajectory (p_value<0.05):\n")
print(significant_trajectory_modules)
```

## Combined figure per module: heatmap + composition + trajectory + pathways, all split by cell type

One publication-style figure per module with a significant age
trajectory, echoing the layout of a canonical multi-panel WGCNA/aging
figure, with **every panel split by cell type**:

- **(A) Expression heatmap by cell type** -- z-scored expression of the
  module's top genes (mice ordered by age, left to right), with genes
  grouped into one facet block per cell type of origin -- the
  cell-type-resolved analogue of a standard "genes x samples" module
  heatmap.
- **(B) Composition** -- number of module genes contributed by each cell
  type.
- **(C) Trajectory by cell type** -- one age trajectory per cell type,
  computed from a **cell-type-restricted eigengene**: for each (module,
  cell type) pair, `moduleEigengenes()` is run again on only the subset
  of features belonging to that cell type, so this is not the single
  pooled module eigengene split visually -- it is a genuinely separate
  PC1 computed from that cell type's genes only.
- **(D) Pathway bubble matrix by cell type** -- one bubble grid per
  database (GO Biological Process, KEGG, Reactome), x = cell type, y =
  top pathways for that module, size = % of the module's genes (in that
  cell type) annotated to the term, color = kME-weighted activity-score
  proxy (see `activity-score-proxy` chunk above).
- **(E) Canonical-pathway lollipop** -- -log10(FDR) of each pathway's
  best result across cell types, one panel per database.

```{r module-trajectory-pathway-figure, fig.width=22, fig.height=14, fig.keep='all', results='asis'}

age_num_lookup <- setNames(age_num, rownames(MEs))

# Palette unificada de colores por tipo celular (reutilizada en TODOS los paneles)
celltype_colors <- setNames(
  viridis::viridis(length(target_celltypes), option = "turbo"),
  sort(target_celltypes)
)

# Bases de datos a plotear (excluyendo KEGG)
databases_to_plot <- c("GO_BP", "Reactome")

# --- Helper: safe, significance-ordered GO/pathway term factor -------------
order_go_terms <- function(df, desc_col = "Description", p_col = "p.adjust") {
  term_levels <- df %>%
    dplyr::group_by(.data[[desc_col]]) %>%
    dplyr::summarise(.best_p = min(.data[[p_col]], na.rm = TRUE), .groups = "drop") %>%
    dplyr::arrange(dplyr::desc(.best_p)) %>%
    dplyr::pull(.data[[desc_col]])
  
  df[[desc_col]] <- factor(df[[desc_col]], levels = term_levels)
  df
}

# --- Panel A: expression heatmap, split by cell type ------------------------
plot_module_heatmap_by_celltype <- function(mod, max_genes_per_ct = 15) {
  feats <- gene_celltype_module %>%
    filter(module == mod) %>%
    group_by(celltype) %>%
    arrange(desc(kME)) %>%
    slice_head(n = max_genes_per_ct) %>%
    ungroup()
  
  if (nrow(feats) == 0) return(NULL)
  
  mat <- pseudobulk_hvg[feats$gene_celltype, , drop = FALSE]
  z <- t(scale(t(mat)))
  
  sample_order <- names(sort(age_num_lookup[colnames(z)]))
  
  df <- as.data.frame(z) %>%
    tibble::rownames_to_column("gene_celltype") %>%
    tidyr::pivot_longer(-gene_celltype, names_to = "sample_id", values_to = "z") %>%
    left_join(feats %>% dplyr::select(gene_celltype, celltype, gene, kME), by = "gene_celltype") %>%
    mutate(sample_id = factor(sample_id, levels = sample_order),
           gene_celltype = factor(gene_celltype,
                                  levels = feats %>% arrange(celltype, desc(kME)) %>% pull(gene_celltype)))
  
  ggplot(df, aes(x = sample_id, y = gene_celltype, fill = z)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                         midpoint = 0, name = "z-score") +
    facet_grid(celltype ~ ., scales = "free_y", space = "free_y") +
    labs(x = "Mice (ordered by age, young -> old)", y = NULL,
         title = "Expression heatmap by cell type") +
    theme_pub(base_size = 10) +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      strip.text.y = element_text(angle = 0, size = 8, face = "bold"),
      panel.spacing = unit(0.15, "cm"),
      panel.grid = element_blank(),
      legend.position = "right",
      legend.key.height = unit(0.4, "cm"),
      plot.title = element_text(size = 11, face = "bold")
    )
}

# --- Panel B: gene composition per cell type (Corregido: colores consistentes y números) ---
plot_module_celltype_counts <- function(mod) {
  d <- gene_counts_by_module_celltype %>% filter(module == mod)
  if (nrow(d) == 0) return(NULL)
  
  ggplot(d, aes(x = reorder(celltype, n_genes), y = n_genes, fill = celltype)) +
    geom_col(width = 0.65, color = "grey20", linewidth = 0.3) +
    geom_text(aes(label = n_genes), hjust = -0.25, size = 3.2, fontface = "bold", color = "grey20") +
    scale_fill_manual(values = celltype_colors) +
    scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
    coord_flip() +
    labs(x = NULL, y = "Number of genes", title = "Gene composition") +
    theme_pub(base_size = 10) +
    theme(
      legend.position = "none",
      axis.text.y = element_text(size = 9, face = "bold"),
      panel.grid.major.x = element_line(color = "grey90", linewidth = 0.3),
      panel.grid.major.y = element_blank(),
      plot.title = element_text(size = 11, face = "bold")
    )
}

# --- Panel C: trajectory per cell type (Corregido: colores consistentes) -----
compute_celltype_module_eigengene <- function(mod, ct, min_features = 3) {
  feats <- gene_celltype_module %>%
    filter(module == mod, celltype == ct) %>%
    pull(gene_celltype)
  
  if (length(feats) < min_features) return(NULL)
  
  sub_mat <- pseudobulk_hvg[feats, , drop = FALSE]
  
  me <- tryCatch(
    moduleEigengenes(t(sub_mat), colors = rep("ct", length(feats)))$eigengenes,
    error = function(e) NULL
  )
  if (is.null(me)) return(NULL)
  
  mice <- colnames(sub_mat)
  data.frame(
    sample_id = mice,
    celltype  = ct,
    ME        = me[[1]],
    age       = age_num_lookup[mice],
    n_genes   = length(feats)
  )
}

plot_module_trend_by_celltype <- function(mod) {
  cts <- unique(gene_celltype_module$celltype[gene_celltype_module$module == mod])
  
  df_all <- bind_rows(lapply(cts, function(ct) compute_celltype_module_eigengene(mod, ct)))
  if (is.null(df_all) || nrow(df_all) == 0) return(NULL)
  
  row <- module_trajectory %>% filter(module == mod)
  subtitle_txt <- if (nrow(row) == 1) {
    paste0("Pooled module: ", tolower(row$shape), " / ", row$category,
           ", p-value = ", signif(row$p_value, 3))
  } else {
    NULL
  }
  
  ggplot(df_all, aes(x = age, y = ME, color = celltype, fill = celltype)) +
    geom_point(alpha = 0.6, size = 2) +
    geom_smooth(method = "loess", span = 1.3, se = TRUE, alpha = 0.12, linewidth = 0.9) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
    scale_color_manual(values = celltype_colors, name = "Cell type") +
    scale_fill_manual(values = celltype_colors, name = "Cell type") +
    labs(x = "Age", y = "Cell-type-restricted eigengene",
         title = paste0("Module ", mod, " -- trajectory by cell type"),
         subtitle = subtitle_txt) +
    theme_pub(base_size = 10) +
    theme(
      legend.position = "right",
      plot.title = element_text(size = 11, face = "bold"),
      plot.subtitle = element_text(size = 8.5, color = "grey40")
    )
}

# --- Helper: placeholder panel cuando un modulo/base de datos no tiene
# terminos de enriquecimiento -- asi GO_BP y Reactome siempre ocupan su
# columna en la figura combinada, en vez de desaparecer cuando no hay datos --
empty_pathway_panel <- function(title_txt) {
  ggplot() +
    annotate("text", x = 0, y = 0,
             label = "No enrichment terms\nfor this module/database",
             size = 3.5, color = "grey55", fontface = "italic") +
    xlim(-1, 1) + ylim(-1, 1) +
    labs(title = title_txt) +
    theme_void(base_size = 10) +
    theme(plot.title = element_text(size = 10.5, face = "bold", hjust = 0))
}

# --- Panel D: pathway bubble matrix split by cell type -----------------------
plot_module_pathway_bubble <- function(mod, database_label, n_top_per_ct = 5, n_terms_max = 12) {
  df_db <- enrichment_by_module_celltype %>%
    filter(module == mod, database == database_label)
  if (nrow(df_db) == 0) return(NULL)
  
  top_terms_per_ct <- df_db %>%
    group_by(celltype) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top_per_ct) %>%
    ungroup() %>%
    pull(Description) %>%
    unique()
  
  if (length(top_terms_per_ct) > n_terms_max) {
    top_terms_per_ct <- df_db %>%
      group_by(Description) %>%
      summarise(best_p = min(p.adjust), .groups = "drop") %>%
      arrange(best_p) %>%
      slice_head(n = n_terms_max) %>%
      pull(Description)
  }
  
  n_genes_mod <- gene_counts_by_module_celltype %>%
    filter(module == mod) %>%
    dplyr::select(celltype, n_genes)
  
  df_plot <- df_db %>%
    filter(Description %in% top_terms_per_ct) %>%
    left_join(n_genes_mod, by = "celltype") %>%
    mutate(pct_genes = 100 * Count / n_genes,
           neg_log_fdr = -log10(p.adjust)) %>%
    order_go_terms(desc_col = "Description", p_col = "p.adjust")
  
  ggplot(df_plot, aes(x = celltype, y = Description)) +
    geom_point(aes(size = pct_genes, fill = activity_score),
               pch = 21, color = "grey20", stroke = 0.3, alpha = 0.9) +
    scale_fill_gradient2(low = "#4575b4", mid = "white", high = "#d73027",
                         midpoint = 0, name = "Activity\nscore (Z)") +
    scale_size_continuous(name = "% genes", range = c(2, 7)) +
    scale_y_discrete(labels = scales::label_wrap(40)) +
    labs(x = "Cell type", y = NULL, title = paste0(database_label, " - Enriched Terms"),
         subtitle = "Size = % of module genes annotated to term (not filtered by significance)") +
    theme_pub(base_size = 9.5) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8.5),
          axis.text.y = element_text(size = 8, lineheight = 0.85),
          panel.grid.major = element_line(linewidth = 0.2, color = "grey92"),
          legend.position = "right",
          legend.box = "vertical",
          legend.key.size = unit(0.3, "cm"),
          plot.title = element_text(size = 10.5, face = "bold"),
          plot.subtitle = element_text(size = 7.5, color = "grey40"))
}

# --- Panel E: canonical-pathway lollipop (-log10 FDR) -----------------------
plot_module_pathway_lollipop <- function(mod, database_label, n_terms = 10) {
  df_db <- enrichment_by_module_celltype %>%
    filter(module == mod, database == database_label)
  if (nrow(df_db) == 0) return(NULL)
  
  df_top <- df_db %>%
    dplyr::group_by(Description) %>%
    dplyr::slice_min(p.adjust, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(p.adjust) %>%
    dplyr::slice_head(n = n_terms) %>%
    dplyr::mutate(neg_log_fdr = -log10(p.adjust),
                  Description = factor(Description, levels = rev(Description)))
  
  ggplot(df_top, aes(x = neg_log_fdr, y = Description)) +
    geom_segment(aes(x = 0, xend = neg_log_fdr, y = Description, yend = Description),
                 color = "grey60", linewidth = 0.4) +
    geom_point(aes(fill = celltype), pch = 21, color = "grey20", size = 3.5, stroke = 0.3) +
    scale_fill_manual(values = celltype_colors, name = "Best cell type", drop = FALSE) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "grey40", linewidth = 0.4) +
    scale_y_discrete(labels = scales::label_wrap(40)) +
    labs(x = "-log10(FDR)", y = NULL, title = paste0(database_label, " - Top Significance"),
         subtitle = "Color = cell type with lowest FDR") +
    theme_pub(base_size = 9.5) +
    theme(axis.text.y = element_text(size = 8, lineheight = 0.85),
          plot.title = element_text(size = 10.5, face = "bold"),
          plot.subtitle = element_text(size = 7.5, color = "grey40"),
          panel.grid.minor = element_blank(),
          legend.position = "right",
          legend.key.size = unit(0.3, "cm"))
}

# --- Standalone Heatmap por Base de Datos -----------------------------------
create_module_pathway_heatmap <- function(mod, database_label, n_top = 8) {
  df_db <- enrichment_by_module_celltype %>%
    filter(module == mod, database == database_label)
  if (nrow(df_db) == 0) return(NULL)
  
  top_terms_per_ct <- df_db %>%
    group_by(celltype) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_top) %>%
    ungroup() %>%
    pull(Description) %>%
    unique()
  
  if (length(top_terms_per_ct) > 12) {
    top_terms_per_ct <- df_db %>%
      group_by(Description) %>%
      summarise(best_p = min(p.adjust), .groups = "drop") %>%
      arrange(best_p) %>%
      slice_head(n = 12) %>%
      pull(Description)
  }
  
  df_plot <- df_db %>%
    filter(Description %in% top_terms_per_ct) %>%
    mutate(neg_log_fdr = -log10(p.adjust)) %>%
    order_go_terms(desc_col = "Description", p_col = "p.adjust")
  
  ggplot(df_plot, aes(x = celltype, y = Description)) +
    geom_tile(aes(fill = neg_log_fdr), color = "white", linewidth = 0.5) +
    geom_text(aes(label = sprintf("%.1f", neg_log_fdr)),
              size = 3.5, color = "white", fontface = "bold") +
    scale_fill_gradient(low = "#4575b4", high = "#d73027",
                        name = "-log10(FDR)",
                        limits = c(0, max(df_plot$neg_log_fdr, na.rm = TRUE))) +
    scale_y_discrete(labels = scales::label_wrap(48)) +
    labs(x = "Cell type",
         y = paste0(database_label, " term"),
         title = paste0("Module ", mod, " - ", database_label, " Terms by Cell Type (FDR < 0.05)")) +
    theme_pub(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, size = 11),
      axis.text.y = element_text(size = 10, lineheight = 0.85),
      plot.title = element_text(size = 13, face = "bold"),
      legend.position = "right",
      legend.key.height = unit(0.4, "cm"),
      panel.grid = element_blank()
    )
}

# --- Reestructuración completa del ensamblado PDF con Patchwork -------------
build_module_publication_figure <- function(mod) {
  p_heatmap <- plot_module_heatmap_by_celltype(mod)
  p_counts  <- plot_module_celltype_counts(mod)
  p_trend   <- plot_module_trend_by_celltype(mod)
  
  if (is.null(p_heatmap)) p_heatmap <- patchwork::plot_spacer()
  if (is.null(p_counts))  p_counts  <- patchwork::plot_spacer()
  if (is.null(p_trend))   p_trend   <- patchwork::plot_spacer()
  
  # Fila superior (Expresión, Composición y Trayectoria)
  top <- p_heatmap + p_counts + p_trend +
    patchwork::plot_layout(widths = c(1.2, 0.7, 1.2))
  
  # Fila inferior: Columnas estructuradas por Base de Datos (GO_BP vs Reactome).
  # Cada columna apila verticalmente Bubble (arriba) y Lollipop (abajo).
  # GO_BP y Reactome SIEMPRE ocupan su columna, incluso si el modulo no tiene
  # terminos para esa base de datos (panel vacio en ese caso, en vez de
  # desaparecer) -- no se filtra por significancia en ningun panel: eso ya
  # queda explicito en el panel lollipop (-log10 FDR) de cada columna.
  db_columns <- lapply(databases_to_plot, function(db) {
    p_bub <- plot_module_pathway_bubble(mod, db)
    p_lol <- plot_module_pathway_lollipop(mod, db)
    
    if (is.null(p_bub)) p_bub <- empty_pathway_panel(paste0(db, " - Enriched Terms"))
    if (is.null(p_lol)) p_lol <- empty_pathway_panel(paste0(db, " - Top Significance"))
    
    p_bub / p_lol + patchwork::plot_layout(heights = c(1, 1))
  })
  
  bottom <- patchwork::wrap_plots(db_columns, ncol = length(db_columns))
  
  # Integración final vertical
  top / bottom +
    patchwork::plot_layout(heights = c(1, 1.7)) +
    patchwork::plot_annotation(
      title = paste0("Module ", mod, " -- expression, composition, trajectory, and pathway enrichment, all by cell type"),
      theme = theme_pub(base_size = 12)
    )
}

# --- Main rendering and saving loop -----------------------------------------
# Figura combinada (heatmap + composicion + trayectoria + pathways) para
# TODOS los modulos, no solo los significativos. Los modulos con trayectoria
# de edad significativa (p_value < 0.05, no FDR) ademas se guardan por
# duplicado con el sufijo "_significant" en el nombre del PDF, para poder
# ubicarlos rapido sin perder el set completo.
dir.create("module_trajectory_pathway_by_celltype", showWarnings = FALSE)
dir.create("module_pathway_heatmaps", showWarnings = FALSE)

all_modules_for_figure <- module_trajectory %>%
  dplyr::arrange(p_value) %>%
  dplyr::pull(module)

for (mod in all_modules_for_figure) {
  
  cat("\n\n## Módulo:", mod, "\n\n")
  
  tryCatch({
    p_full <- build_module_publication_figure(mod)
    
    if (!is.null(p_full)) {
      print(p_full)
    }
    
    base_path <- file.path("module_trajectory_pathway_by_celltype",
                           paste0("module_", mod, "_trajectory_pathways.pdf"))
    save_pub_pdf(p_full, filename = base_path, width = 22, height = 14)
    
    row_traj <- module_trajectory %>% dplyr::filter(module == mod)
    is_significant <- nrow(row_traj) == 1 && !is.na(row_traj$p_value) && row_traj$p_value < 0.05
    
    if (is_significant) {
      sig_path <- file.path("module_trajectory_pathway_by_celltype",
                            paste0("module_", mod, "_trajectory_pathways_significant.pdf"))
      file.copy(base_path, sig_path, overwrite = TRUE)
      cat("  -> age-associated (p =", signif(row_traj$p_value, 3),
          "), copy saved as", basename(sig_path), "\n")
    }
    
    for (db in databases_to_plot) {
      p_heatmap_db <- create_module_pathway_heatmap(mod, db, n_top = 8)
      if (!is.null(p_heatmap_db)) {
        save_pub_pdf(p_heatmap_db,
                     filename = paste0("module_pathway_heatmaps/", db, "_heatmap_module_", mod, ".pdf"),
                     width = 10,
                     height = 8)
      }
    }
  }, error = function(e) {
    message(paste0("Error processing module ", mod, ": ", e$message))
  })
}

```

## Trajectory + GO terms per module x cell-type combination, stacked by cell type

The combined per-module figure above overlays all cell types in one
trajectory panel and pools GO terms across cell types. This section
instead builds **one row per (module, cell type) combination**: the
cell-type-restricted eigengene trajectory for that combination alone
(its own panel, not overlaid with other cell types), with the top
significant **GO Biological Process** terms *for that same cell type
only* right next to it. Rows are then stacked vertically into **one
figure per cell type** (4 figures total, one per target cell type),
ordered by how significant that module's overall age trajectory is.

Cell type color is kept consistent with the rest of the report (same
`viridis`/`"turbo"` palette used in the combined per-module figure), so
a given cell type is always the same color across every plot.

```{r module-celltype-combo-setup}
# `celltype_colors` and `order_go_terms()` were already defined earlier
# (module-trajectory-pathway-figure chunk) and are reused here so cell-type
# colors stay identical across every plot in the report.

# --- Trajectory panel: ONE cell type, its own cell-type-restricted eigengene
plot_celltype_module_trend <- function(mod, ct) {
  df <- compute_celltype_module_eigengene(mod, ct)
  if (is.null(df) || nrow(df) == 0) return(NULL)
  
  col_ct <- unname(celltype_colors[ct])
  
  ggplot(df, aes(x = age, y = ME)) +
    geom_point(color = col_ct, size = 2.2, alpha = 0.75) +
    geom_smooth(method = "loess", span = 1.3, se = TRUE,
                color = col_ct, fill = col_ct, alpha = 0.15, linewidth = 0.9) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
    labs(x = "Age", y = "Cell-type-restricted eigengene",
         title = paste0("Module ", mod, " -- ", ct),
         subtitle = paste0("n genes = ", unique(df$n_genes))) +
    theme_pub(base_size = 10) +
    theme(
      plot.title    = element_text(size = 10, face = "bold"),
      plot.subtitle = element_text(size = 8, color = "grey40")
    )
}

# --- GO panel: top significant GO_BP terms for THIS module x cell type only
plot_celltype_module_go_terms <- function(mod, ct, n_terms = 10) {
  df <- enrichment_by_module_celltype %>%
    filter(module == mod, celltype == ct, database == "GO_BP") %>%
    # guard against duplicate term rows (e.g. re-run enrichment merges) so
    # the same GO term is never plotted twice for this module x cell type
    dplyr::distinct(Description, .keep_all = TRUE) %>%
    arrange(p.adjust) %>%
    slice_head(n = n_terms)
  
  if (nrow(df) == 0) {
    return(
      ggplot() +
        annotate("text", x = 0, y = 0, label = paste0("No significant GO BP terms\n(", ct, ")"),
                 size = 3.2, color = "grey40") +
        theme_void()
    )
  }
  
  col_ct <- unname(celltype_colors[ct])
  
  df <- df %>%
    mutate(
      neg_log_fdr = -log10(p.adjust),
      Description = factor(Description, levels = rev(Description))
    )
  
  ggplot(df, aes(x = neg_log_fdr, y = Description)) +
    geom_segment(aes(x = 0, xend = neg_log_fdr, y = Description, yend = Description),
                 color = "grey70", linewidth = 0.5) +
    geom_point(aes(size = Count), color = col_ct, alpha = 0.9) +
    geom_vline(xintercept = -log10(0.05), linetype = "dashed", color = "grey40", linewidth = 0.4) +
    scale_size_continuous(name = "N genes", range = c(2, 7)) +
    scale_y_discrete(labels = scales::label_wrap(40)) +
    labs(x = "-log10(FDR)", y = NULL,
         title = paste0("Top GO Biological Process terms -- ", ct)) +
    theme_pub(base_size = 9) +
    theme(
      axis.text.y      = element_text(size = 7.5, lineheight = 0.85),
      plot.title       = element_text(size = 9, face = "bold", color = col_ct),
      panel.grid.minor = element_blank(),
      legend.position  = "right",
      legend.key.size  = unit(0.3, "cm")
    )
}

# --- One row = trajectory (left) + matching GO terms (right), for one combo
build_module_celltype_combo_row <- function(mod, ct) {
  p_trend <- plot_celltype_module_trend(mod, ct)
  if (is.null(p_trend)) return(NULL)
  p_go <- plot_celltype_module_go_terms(mod, ct)
  p_trend + p_go + patchwork::plot_layout(widths = c(1, 1.4))
}
```

```{r module-celltype-combo-stacked-by-celltype, fig.width=12, fig.height=14}
conflicts_prefer(base::unname)
dir.create("module_celltype_trajectory_GO_stacked", showWarnings = FALSE)

for (ct in sort(target_celltypes)) {
  
  # Módulos con trayectoria de edad significativa y genes presentes en este tipo celular,
  # ordenados por p-valor de la trayectoria (más significativo primero)
  mods_this_ct <- gene_celltype_module %>%
    dplyr::filter(celltype == ct, module %in% significant_trajectory_modules) %>%
    dplyr::distinct(module) %>%
    dplyr::pull(module)
  
  mods_this_ct <- module_trajectory %>%
    dplyr::filter(module %in% mods_this_ct) %>%
    dplyr::arrange(p_value) %>%
    dplyr::pull(module)
  
  if (length(mods_this_ct) == 0) {
    cat("[", ct, "] No hay módulos con trayectoria significativa y genes asignados -- omitiendo.\n", sep = "")
    next
  }
  
  # Generar lista de filas pareadas ignorando valores nulos
  row_list <- purrr::compact(
    lapply(mods_this_ct, function(mod) build_module_celltype_combo_row(mod, ct))
  )
  
  if (length(row_list) == 0) {
    cat("[", ct, "] Sin filas válidas para graficar -- omitiendo.\n", sep = "")
    next
  }
  
  col_ct <- if (exists("celltype_colors") && ct %in% names(celltype_colors)) {
    unname(celltype_colors[ct])
  } else {
    "grey20"
  }
  
  n_mods <- length(row_list)
  
  # Apilado vertical con Patchwork
  p_stack <- patchwork::wrap_plots(row_list, ncol = 1) +
    patchwork::plot_annotation(
      title = paste0(ct, " — Module Trajectories & Top GO Biological Process Terms"),
      subtitle = paste0(
        n_mods, " module", ifelse(n_mods > 1, "s", ""),
        " with a significant age trajectory (p_value < 0.05) and genes from ", ct, "\n",
        "(Trajectory restricted to this cell type; GO terms restricted to this cell type)"
      ),
      theme = theme_pub(base_size = 12) +
        theme(
          plot.title    = element_text(size = 14, face = "bold", color = col_ct),
          plot.subtitle = element_text(size = 9, color = "grey30", lineheight = 1.1)
        )
    )
  
  # Mostrar gráfica en la sesión interactiva / informe HTML
  print(p_stack)
  
  # Cálculo dinámico de altura para el PDF (base de título + 3.2 in por módulo)
  pdf_height <- max(4.5, 3.2 * n_mods + 1.2)
  file_out <- paste0("module_celltype_trajectory_GO_stacked/", ct, "_modules_trajectory_GO.pdf")
  
  save_pub_pdf(
    p_stack,
    filename = file_out,
    width = 12,
    height = pdf_height
  )
  
  cat("[", ct, "] PDF generado: ", file_out, " (", n_mods, " módulos, alto = ", round(pdf_height, 1), " in)\n", sep = "")
}
```

## Master heatmap: pathway enrichment across all modules and cell types (GO / KEGG / Reactome)

Large, publication-ready heatmap showing pathway enrichment across the
entire analysis, generated once per database via a shared function.

```{r master-pathway-heatmap, fig.width=16, fig.height=12}


build_master_pathway_heatmap <- function(database_label, n_per_module = 6, n_terms_max = 20) {
  df_all <- enrichment_by_module_celltype %>%
    filter(module %in% significant_trajectory_modules, database == database_label) %>%
    mutate(neg_log_fdr = -log10(p.adjust))

  if (nrow(df_all) == 0) return(NULL)

  top_terms_per_module <- df_all %>%
    group_by(module, Description) %>%
    summarise(best_fdr = min(p.adjust), n_ct = n_distinct(celltype), .groups = "drop") %>%
    arrange(best_fdr) %>%
    group_by(module) %>%
    slice_head(n = n_per_module) %>%
    ungroup() %>%
    pull(Description) %>%
    unique()

  if (length(top_terms_per_module) > n_terms_max) {
    top_terms_per_module <- df_all %>%
      group_by(Description) %>%
      summarise(best_p = min(p.adjust), .groups = "drop") %>%
      arrange(best_p) %>%
      slice_head(n = n_terms_max) %>%
      pull(Description)
  }

  df_master <- df_all %>%
    filter(Description %in% top_terms_per_module) %>%
    mutate(module_celltype = paste(module, celltype, sep = " / "),
           Description = factor(Description, levels = rev(sort(top_terms_per_module))))

  ggplot(df_master, aes(x = module_celltype, y = Description)) +
    geom_tile(aes(fill = neg_log_fdr), color = "white", linewidth = 0.6) +
    geom_text(data = df_master %>% filter(neg_log_fdr > 0),
              aes(label = sprintf("%.0f", neg_log_fdr)),
              size = 3, color = "white", fontface = "bold") +
    scale_fill_gradient(low = "#f7fbff", high = "#08519c",
                         name = "-log10(FDR)",
                         limits = c(0, max(df_master$neg_log_fdr, na.rm = TRUE)),
                         na.value = "#f7fbff") +
    labs(x = "Module / Cell Type",
         y = paste0(database_label, " term"),
         title = paste0(database_label, " Enrichment: All Significant Modules and Cell Types")) +
    theme_pub(base_size = 12) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 10),
      axis.text.y = element_text(size = 10),
      plot.title = element_text(size = 14, face = "bold", margin = margin(b = 15)),
      legend.position = "right",
      legend.key.height = unit(0.5, "cm"),
      panel.grid = element_blank(),
      plot.margin = margin(r = 20)
    )
}

for (db in c("GO_BP", "KEGG", "Reactome")) {
  p_master_db <- build_master_pathway_heatmap(db)
  if (!is.null(p_master_db)) {
    print(p_master_db)
    save_pub_pdf(p_master_db,
                 filename = paste0(db, "_master_heatmap_all_modules.pdf"),
                 width = 16,
                 height = 12)
    cat("Saved master", db, "heatmap to plots_pdf/", db, "_master_heatmap_all_modules.pdf\n")
  } else {
    cat("No significant", db, "results to build a master heatmap for.\n")
  }
}
```

## Additional exploratory visualizations

Six extra views of the same underlying tables (`gene_celltype_module`,
`MEs`, `module_trajectory`, `enrichment_by_module_celltype`) -- no new
data is computed upstream of this section, only new plots.

### 1. Module membership (kME) by cell type, hub genes highlighted

For every module, kME (module membership) of every gene is plotted
against the cell type it came from (jittered), with **hub genes** (the
same top-10% kME definition used in `hub_genes_table`) highlighted in
red. This is the plot-equivalent of `hub_genes_by_module.csv`: instead
of scrolling a table, you see at a glance whether hub genes concentrate
in one cell type or are spread across all of them, and how kME is
distributed overall.

```{r kme-jitter-setup}
gene_celltype_module <- gene_celltype_module %>%
  mutate(is_hub = gene_celltype %in% hub_genes_table$gene_celltype)

plot_kme_jitter_module <- function(mod) {
  df <- gene_celltype_module %>% filter(module == mod, !is.na(kME))
  if (nrow(df) == 0) return(NULL)

  ggplot(df, aes(x = celltype, y = kME)) +
    geom_jitter(aes(color = is_hub, size = is_hub, alpha = is_hub),
                width = 0.28, height = 0) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.3) +
    scale_color_manual(values = c("TRUE" = "#d73027", "FALSE" = "grey65"),
                        labels = c("TRUE" = "Hub gene (top 10% kME)", "FALSE" = "Other gene"),
                        name = NULL) +
    scale_size_manual(values = c("TRUE" = 2.3, "FALSE" = 1.1), guide = "none") +
    scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.35), guide = "none") +
    labs(x = "Cell type", y = "kME (module membership)",
         title = paste0("Module ", mod, " -- module membership by cell type")) +
    theme_pub(base_size = 11) +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 30, hjust = 1)
    )
}
```

```{r kme-jitter-overview, fig.width=13, fig.height=9}
# Quick multi-module overview, restricted to modules with a significant age
# trajectory (otherwise this gets crowded) -- one facet per module.
overview_modules <- significant_trajectory_modules
if (length(overview_modules) > 0) {
  df_overview <- gene_celltype_module %>%
    filter(module %in% overview_modules, !is.na(kME)) %>%
    mutate(module = factor(module, levels = overview_modules))

  p_kme_overview <- ggplot(df_overview, aes(x = celltype, y = kME)) +
    geom_jitter(aes(color = is_hub, size = is_hub, alpha = is_hub),
                width = 0.28, height = 0) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey60", linewidth = 0.3) +
    scale_color_manual(values = c("TRUE" = "#d73027", "FALSE" = "grey65"),
                        labels = c("TRUE" = "Hub gene (top 10% kME)", "FALSE" = "Other gene"),
                        name = NULL) +
    scale_size_manual(values = c("TRUE" = 2, "FALSE" = 0.9), guide = "none") +
    scale_alpha_manual(values = c("TRUE" = 0.9, "FALSE" = 0.3), guide = "none") +
    facet_wrap(~module, scales = "free_y") +
    labs(x = "Cell type", y = "kME (module membership)",
         title = "Module membership by cell type -- significant-trajectory modules") +
    theme_pub(base_size = 10) +
    theme(
      legend.position = "top",
      axis.text.x = element_text(angle = 40, hjust = 1, size = 8)
    )

  print(p_kme_overview)
  save_pub_pdf(p_kme_overview, "kme_jitter_overview_significant_modules.pdf", width = 13, height = 9)
} else {
  cat("No significant-trajectory modules -- skipping overview jitter plot.\n")
}
```

```{r kme-jitter-per-module, fig.width=6, fig.height=5}
dir.create("kme_jitter_by_module", showWarnings = FALSE)

for (mod in significant_trajectory_modules) {
  p_kme <- plot_kme_jitter_module(mod)
  if (is.null(p_kme)) next
  print(p_kme)
  save_pub_pdf(p_kme,
               filename = paste0("kme_jitter_by_module/module_", mod, "_kme_jitter.pdf"),
               width = 6, height = 5)
}
```

### 2. Alluvial / Sankey diagram: module -\> cell type composition

Flow diagram where the left axis is module, the right axis is cell type,
flow width is the number of genes -- the same information as the
composition heatmap (`heatmap_composicion_celular_por_modulo.pdf`), but
easier to scan when there are many modules: dominant cell types per
module jump out as thick flows. Flows and the module axis are colored by
the module's own WGCNA color.

```{r module-celltype-alluvial, fig.width=12, fig.height=max(6, 0.35 * length(unique(gene_counts_by_module_celltype$module)))}
alluvial_data <- gene_counts_by_module_celltype %>%
  filter(module != "grey") %>%
  mutate(module_color = tolower(module))

# Order modules by total size (top to bottom) for a cleaner read
module_order <- alluvial_data %>%
  group_by(module) %>%
  summarise(total = sum(n_genes), .groups = "drop") %>%
  arrange(desc(total)) %>%
  pull(module)

alluvial_data <- alluvial_data %>%
  mutate(module = factor(module, levels = module_order))

p_alluvial <- ggplot(alluvial_data,
                      aes(axis1 = module, axis2 = celltype, y = n_genes)) +
  geom_alluvium(aes(fill = module_color), width = 1/6, alpha = 0.85, color = "white", linewidth = 0.1) +
  geom_stratum(width = 1/6, fill = "grey92", color = "grey40", linewidth = 0.3) +
  geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 2.8) +
  scale_fill_identity() +
  scale_x_discrete(limits = c("Module", "Cell type"), expand = c(0.08, 0.08)) +
  labs(x = NULL, y = "Number of genes",
       title = "Module -> cell type composition (all genes)") +
  theme_pub(base_size = 11) +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    panel.grid = element_blank(),
    axis.text.x = element_text(face = "bold", size = 12)
  )

print(p_alluvial)
save_pub_pdf(p_alluvial, "alluvial_module_celltype_composition.pdf",
             width = 12, height = max(6, 0.35 * length(unique(alluvial_data$module))))

# Second version restricted to modules with a significant age trajectory,
# for a less crowded "what matters" view
if (length(significant_trajectory_modules) > 0) {
  alluvial_sig <- alluvial_data %>% filter(as.character(module) %in% significant_trajectory_modules)

  p_alluvial_sig <- ggplot(alluvial_sig,
                            aes(axis1 = module, axis2 = celltype, y = n_genes)) +
    geom_alluvium(aes(fill = module_color), width = 1/6, alpha = 0.85, color = "white", linewidth = 0.1) +
    geom_stratum(width = 1/6, fill = "grey92", color = "grey40", linewidth = 0.3) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.2) +
    scale_fill_identity() +
    scale_x_discrete(limits = c("Module", "Cell type"), expand = c(0.08, 0.08)) +
    labs(x = NULL, y = "Number of genes",
         title = "Module -> cell type composition (significant age-trajectory modules only)") +
    theme_pub(base_size = 12) +
    theme(
      axis.text.y = element_blank(),
      axis.ticks.y = element_blank(),
      panel.grid = element_blank(),
      axis.text.x = element_text(face = "bold", size = 13)
    )

  print(p_alluvial_sig)
  save_pub_pdf(p_alluvial_sig, "alluvial_module_celltype_composition_significant.pdf",
               width = 10, height = 7)
}
```

### 3. Jaccard overlap matrix between modules

Pairwise Jaccard index (shared genes / union of genes, gene identity
only -- cell-type origin is ignored so a gene shared across cell types
within a module counts once) between every pair of modules, as a
symmetric heatmap. High off-diagonal values flag potentially redundant
modules (near-duplicate gene sets) that might be candidates for merging
in a future run; low values everywhere confirm the modules are capturing
distinct biology.

```{r module-jaccard-overlap, fig.width=9, fig.height=8}
gene_sets_by_module <- gene_celltype_module %>%
  filter(module != "grey") %>%
  distinct(module, gene) %>%
  split(x = .$gene, f = .$module)

mods_jaccard <- names(gene_sets_by_module)

jaccard_mat <- matrix(NA_real_, nrow = length(mods_jaccard), ncol = length(mods_jaccard),
                       dimnames = list(mods_jaccard, mods_jaccard))

for (i in seq_along(mods_jaccard)) {
  for (j in i:length(mods_jaccard)) {
    a <- gene_sets_by_module[[mods_jaccard[i]]]
    b <- gene_sets_by_module[[mods_jaccard[j]]]
    j_val <- length(intersect(a, b)) / length(union(a, b))
    jaccard_mat[i, j] <- j_val
    jaccard_mat[j, i] <- j_val
  }
}

# Order modules by hierarchical clustering on (1 - Jaccard) distance, so
# similar modules sit next to each other in the heatmap
jaccard_dist  <- as.dist(1 - jaccard_mat)
jaccard_hclust <- hclust(jaccard_dist, method = "average")
mod_order_jaccard <- mods_jaccard[jaccard_hclust$order]

jaccard_df <- as.data.frame(as.table(jaccard_mat)) %>%
  setNames(c("module_1", "module_2", "jaccard")) %>%
  mutate(
    module_1 = factor(module_1, levels = mod_order_jaccard),
    module_2 = factor(module_2, levels = mod_order_jaccard)
  )

p_jaccard <- ggplot(jaccard_df, aes(x = module_1, y = module_2, fill = jaccard)) +
  geom_tile(color = "white", linewidth = 0.2) +
  scale_fill_gradient(low = "white", high = "#08519c", limits = c(0, 1),
                       name = "Jaccard\nindex") +
  labs(x = NULL, y = NULL,
       title = "Pairwise gene-set overlap between modules (Jaccard index)",
       subtitle = "Off-diagonal values close to 1 flag potentially redundant modules") +
  theme_pub(base_size = 10) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 8),
    axis.text.y = element_text(size = 8),
    panel.grid = element_blank(),
    plot.title = element_text(size = 13, face = "bold"),
    plot.subtitle = element_text(size = 9, color = "grey35")
  ) +
  coord_fixed()

print(p_jaccard)
save_pub_pdf(p_jaccard, "module_jaccard_overlap_heatmap.pdf", width = 9, height = 8)

# Flag any pair (excluding self) above a redundancy threshold, for a quick text summary
redundancy_threshold <- 0.3
redundant_pairs <- jaccard_df %>%
  filter(as.character(module_1) < as.character(module_2), jaccard >= redundancy_threshold) %>%
  arrange(desc(jaccard))

cat("Module pairs with Jaccard >=", redundancy_threshold, "(possible redundancy):\n")
if (nrow(redundant_pairs) > 0) print(redundant_pairs) else cat("None.\n")
```

### 4. PCA of module eigengenes (MEs) by mouse, colored by age

Instead of looking at trajectories one module at a time, this projects
**all mice into the joint module-eigengene space** (PCA on `MEs`, grey
module excluded) and colors each mouse by its real age. If aging has a
coherent, dataset-wide signature across modules (not just isolated
single-module trends), mice should separate along a smooth age gradient
in this space.

```{r me-pca-by-age, fig.width=9, fig.height=7}
MEs_no_grey <- MEs[, !grepl("^MEgrey$", colnames(MEs)), drop = FALSE]

pca_me <- prcomp(MEs_no_grey, center = TRUE, scale. = TRUE)
pca_var <- round(100 * (pca_me$sdev^2) / sum(pca_me$sdev^2), 1)

pca_df <- data.frame(
  sample_id = rownames(MEs_no_grey),
  PC1 = pca_me$x[, 1],
  PC2 = pca_me$x[, 2],
  age = age_num_lookup[rownames(MEs_no_grey)]
)

p_pca_me <- ggplot(pca_df, aes(x = PC1, y = PC2, color = age)) +
  geom_point(size = 3.2, alpha = 0.9) +
  ggrepel::geom_text_repel(aes(label = sample_id), size = 2.5, color = "grey30",
                            max.overlaps = 15, show.legend = FALSE) +
  scale_color_viridis_c(name = "Age", option = "plasma") +
  labs(x = paste0("PC1 (", pca_var[1], "% variance)"),
       y = paste0("PC2 (", pca_var[2], "% variance)"),
       title = "PCA of module eigengenes (MEs) across mice",
       subtitle = "Grey (unassigned) module excluded; color = real age") +
  theme_pub(base_size = 12) +
  theme(plot.title = element_text(size = 13, face = "bold"),
        plot.subtitle = element_text(size = 9, color = "grey35"))

print(p_pca_me)
save_pub_pdf(p_pca_me, "PCA_module_eigengenes_by_age.pdf", width = 9, height = 7)

# UMAP version, only if the 'uwot' package is available (optional dependency
# -- PCA above already gives the core result, UMAP is a bonus nonlinear view)
if (requireNamespace("uwot", quietly = TRUE) && nrow(MEs_no_grey) > 15) {
  set.seed(1)
  umap_me <- uwot::umap(scale(MEs_no_grey), n_neighbors = min(10, nrow(MEs_no_grey) - 1))
  umap_df <- data.frame(
    sample_id = rownames(MEs_no_grey),
    UMAP1 = umap_me[, 1],
    UMAP2 = umap_me[, 2],
    age = age_num_lookup[rownames(MEs_no_grey)]
  )

  p_umap_me <- ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = age)) +
    geom_point(size = 3.2, alpha = 0.9) +
    ggrepel::geom_text_repel(aes(label = sample_id), size = 2.5, color = "grey30",
                              max.overlaps = 15, show.legend = FALSE) +
    scale_color_viridis_c(name = "Age", option = "plasma") +
    labs(title = "UMAP of module eigengenes (MEs) across mice",
         subtitle = "Grey (unassigned) module excluded; color = real age") +
    theme_pub(base_size = 12) +
    theme(plot.title = element_text(size = 13, face = "bold"),
          plot.subtitle = element_text(size = 9, color = "grey35"))

  print(p_umap_me)
  save_pub_pdf(p_umap_me, "UMAP_module_eigengenes_by_age.pdf", width = 9, height = 7)
} else {
  cat("'uwot' not installed (or too few mice) -- skipping the UMAP version; PCA above already computed.\n")
}
```

### 5. All module trajectories, combined in a single figure (small multiples)

One combined figure with **every module** (not only the significant
ones), one facet each, ordered by trajectory p-value -- a fast visual
scan of shapes (Increasing / Decreasing / U-shaped / Inverted U-shaped)
without opening every individual PDF. Point/line color = the module's
own WGCNA color, for consistency with the other figures in this report.
Modules associated with age are defined here as **p-value \< 0.05**
(raw, not FDR-adjusted) and are flagged with a "\*" in the facet label,
so the combined figure and the age-association call use the same, single
criterion.

```{r all-trajectories-facet, fig.width=14, fig.height=10}
all_trajectory_modules <- module_trajectory %>% dplyr::arrange(p_value) %>% dplyr::pull(module)

if (length(all_trajectory_modules) > 0) {
  traj_long <- df_traj %>%
    dplyr::select(sample_id, age_num,
                  all_of(paste0("ME", all_trajectory_modules))) %>%
    tidyr::pivot_longer(-c(sample_id, age_num), names_to = "module", values_to = "ME") %>%
    mutate(module = sub("^ME", "", module)) %>%
    left_join(module_trajectory %>% dplyr::select(module, category, shape, p_value),
              by = "module") %>%
    mutate(
      module = factor(module, levels = all_trajectory_modules),
      module_color = tolower(as.character(module)),
      is_age_associated = p_value < 0.05,
      module_label = paste0(as.character(module), ifelse(is_age_associated, " *", ""),
                             " -- ", category,
                             " (p = ", signif(p_value, 2), ")")
    )

  # keep facet order = p-value order (most significant first)
  label_order <- traj_long %>% distinct(module, module_label) %>% arrange(module) %>% pull(module_label)
  traj_long <- traj_long %>% mutate(module_label = factor(module_label, levels = label_order))

  p_traj_facet <- ggplot(traj_long, aes(x = age_num, y = ME)) +
    geom_point(aes(color = module_color), size = 1.8, alpha = 0.65) +
    geom_smooth(aes(color = module_color, fill = module_color),
                method = "loess", span = 1.3, se = TRUE, alpha = 0.15, linewidth = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "grey50", linewidth = 0.3) +
    scale_color_identity() +
    scale_fill_identity() +
    facet_wrap(~module_label, scales = "free_y", ncol = 4) +
    labs(x = "Age", y = "Module eigengene",
         title = "All module trajectories, combined",
         subtitle = "\"*\" in facet title = module associated with age (p-value < 0.05, not FDR)") +
    theme_pub(base_size = 10) +
    theme(
      strip.text = element_text(size = 8, face = "bold"),
      plot.title = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 9, color = "grey30")
    )

  print(p_traj_facet)
  n_facets <- length(unique(traj_long$module))
  save_pub_pdf(p_traj_facet, "all_module_trajectories_combined.pdf",
               width = 14, height = max(6, 3 * ceiling(n_facets / 4)))
} else {
  cat("No modules with a computed trajectory -- skipping the facet comparison.\n")
}
```

Modules associated with age (p-value \< 0.05, not FDR):

```{r age-associated-modules-summary}
age_associated_modules_table <- module_trajectory %>%
  dplyr::filter(p_value < 0.05) %>%
  dplyr::arrange(p_value) %>%
  dplyr::select(module, shape, category, p_value, fdr, r_squared)

cat("Modules associated with age (p_value < 0.05):\n")
print(age_associated_modules_table)
```

### 6. Gene-concept network (cnetplot) for the top modules

For the modules with the strongest age trajectories, a gene-concept
network connects each module's top enriched GO Biological Process terms
to the genes that support them (pooled across cell types, unlike the
by-cell-type enrichment used elsewhere). Node color = kME (module
membership) of that gene, so hub genes stand out visually inside the
network. Recomputed fresh here (readable `enrichResult` objects aren't
kept from the by-cell-type enrichment above, since `cnetplot()` needs
the actual `enrichResult`, not the flattened data frame) -- limited to a
handful of top modules to keep this fast.

```{r}

library(tidyverse)
library(clusterProfiler)
library(enrichplot)

if (requireNamespace("conflicted", quietly = TRUE)) {
  library(conflicted)
  conflicts_prefer(dplyr::select,   .quiet = TRUE)
  conflicts_prefer(dplyr::filter,   .quiet = TRUE)
  conflicts_prefer(dplyr::mutate,   .quiet = TRUE)
  conflicts_prefer(dplyr::arrange,  .quiet = TRUE)
  conflicts_prefer(dplyr::count,    .quiet = TRUE)
  conflicts_prefer(dplyr::rename,   .quiet = TRUE)
  conflicts_prefer(dplyr::slice,    .quiet = TRUE)
  conflicts_prefer(dplyr::first,    .quiet = TRUE)
  conflicts_prefer(base::intersect, .quiet = TRUE)
  conflicts_prefer(base::union,     .quiet = TRUE)
  conflicts_prefer(base::setdiff,   .quiet = TRUE)
  conflicts_prefer(stats::cor,      .quiet = TRUE)
}

n_top_modules_cnet <- min(6, length(significant_trajectory_modules))
top_modules_cnet <- head(significant_trajectory_modules, n_top_modules_cnet)

dir.create("cnetplot_top_modules_by_celltype", showWarnings = FALSE)

module_celltype_combos <- gene_celltype_module %>%
  dplyr::filter(module %in% top_modules_cnet) %>%
  dplyr::distinct(module, celltype)

cnet_results_summary <- list()

for (i in seq_len(nrow(module_celltype_combos))) {
  mod <- module_celltype_combos$module[i]
  ct  <- module_celltype_combos$celltype[i]
  combo_label <- paste(mod, ct, sep = "_")
  
  tryCatch({
    genes_mod_ct <- gene_celltype_module %>%
      dplyr::filter(module == mod, celltype == ct) %>%
      dplyr::distinct(gene) %>%
      dplyr::pull(gene)
    
    if (length(genes_mod_ct) < 5) {
      message("Module ", mod, " / ", ct, ": <5 genes -- skipping.")
      next
    }
    
    entrez_mod_ct <- convert_to_entrez_flexible(
      setNames(list(genes_mod_ct), combo_label), min_genes = 5, verbose = FALSE
    )
    if (length(entrez_mod_ct) == 0 || is.null(entrez_mod_ct[[combo_label]])) {
      message("Module ", mod, " / ", ct, ": too few genes mapped to Entrez -- skipping cnetplot.")
      next
    }
    
    ego_mod_ct <- clusterProfiler::enrichGO(
      gene          = entrez_mod_ct[[combo_label]],
      universe      = universe_entrez,
      OrgDb         = org.Mm.eg.db,
      ont           = "BP",
      pAdjustMethod = "BH",
      pvalueCutoff  = 0.05,
      qvalueCutoff  = 0.2,
      readable      = TRUE
    )
    
    if (is.null(ego_mod_ct) || nrow(as.data.frame(ego_mod_ct)) == 0) {
      message("Module ", mod, " / ", ct, ": no significant GO_BP terms -- skipping cnetplot.")
      next
    }
    
    # 1. Asegurar vector fc_vec sin duplicados ni NAs
    fc_df <- gene_kme_lookup %>%
      dplyr::filter(module == mod) %>%
      dplyr::distinct(gene, .keep_all = TRUE)
    
    fc_vec <- fc_df$kme
    names(fc_vec) <- fc_df$gene
    fc_vec <- na.omit(fc_vec)
    
    # 2. Generar cnetplot (se retira scale_color_gradient2 directo para evitar conflictos de ggraph)
    p_cnet <- enrichplot::cnetplot(
      ego_mod_ct, 
      foldChange = fc_vec, 
      showCategory = min(6, nrow(as.data.frame(ego_mod_ct))),
      node_label = "all", 
      cex_label_gene = 0.6, 
      cex_label_category = 0.8
    ) +
      labs(title = paste0("Module ", mod, " | ", ct, " -- gene-concept network (GO_BP)")) +
      theme(plot.title = element_text(size = 12, face = "bold"))
    
    # Renderizar en consola/pantalla
    print(p_cnet)
    
    # 3. Guardar con ggsave (o sustituir por save_pub_pdf si está definida en tu entorno)
    ggplot2::ggsave(
      filename = paste0("cnetplot_top_modules_by_celltype/module_", combo_label, "_cnetplot.pdf"),
      plot = p_cnet,
      width = 11, 
      height = 9
    )
    
    # 4. Registrar solo si la gráfica y el guardado se completaron sin errores
    cnet_results_summary[[combo_label]] <- nrow(as.data.frame(ego_mod_ct))
    
  }, error = function(e) {
    message("Module ", mod, " / ", ct, ": cnetplot failed -> ", conditionMessage(e))
  })
}

cat("\nCombinaciones módulo x celltype procesadas con éxito:\n")
if (length(cnet_results_summary) > 0) {
  print(cnet_results_summary)
} else {
  cat("Ninguna combinación generó un gráfico con éxito.\n")
}
```
